---
title: "hypothesis testing Raza fall 2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### This dataset contains simulated data. It is based on the following covid dataset: https://www.kaggle.com/sudalairajkumar/novel-corona-virus-2019-dataset

Import the data
```{r}
covid=read.csv("covid_simulated.csv")

```


Is there a difference between the mean age of people who survive and those who dont?
```{r}

alive=covid[covid$Survival=="Yes","Age"]
dead=covid[covid$Survival=="No","Age"]

#check if the two groups are normal with histogram, qqnorm and shapiro.tets
hist(alive,breaks = 10)
hist(dead,breaks = 10)

qqnorm(alive)
qqnorm(dead)

shapiro.test(alive)  
shapiro.test(dead)


#Age of dead patients is not normal. Use wilcox.test
wilcox.test(alive,dead,paired = F)


```


Is the mean age of all covid patients greater than 60?
```{r}
hist(covid$Age,breaks = 10)
qqnorm(covid$Age)
shapiro.test(log(covid$Age))

#data is not normal, use non-paramtric test
#but since sample size is large t.test might also work
wilcox.test(covid$Age,mu = 60,alternative = "g")


```


Is their a difference between the mean body temperature of all patients on day1 and day7 after infection
```{r}
hist(covid$Temp_day7-covid$Temp_day1)
qqnorm(covid$Temp_day7-covid$Temp_day1)
shapiro.test(covid$Temp_day7-covid$Temp_day1)

#data is not normal. Use non-parametric test with paired=T
#but since sample size is large paired t.test might also work

wilcox.test(covid$Temp_day1,covid$Temp_day7,paired = T)
wilcox.test((covid$Temp_day1-covid$Temp_day7),mu=0)

```


Is their a difference between the mean body temperature of male patients on day1 and day7 after infection
```{r}
males_day1=covid[covid$Gender=="Male","Temp_day1"]
males_day7=covid[covid$Gender=="Male","Temp_day7"]
diff_males=males_day1 - males_day7
hist(diff_males)
shapiro.test(diff_males)  
qqnorm(diff_males)


#data is normal, use paired t.test
t.test(diff_males,mu=0)
#or
t.test(males_day1,males_day7,paired = T)

#manual
t=mean(diff_males)/(sd(diff_males)/sqrt(length(diff_males)))
2*pt(t,df=19,lower.tail = T)

```


Use a permutation test to answer the previous question and calculate the p-value
```{r}
diff=mean(males_day1)-mean(males_day7)

p=c(males_day1,males_day7)
fac=c(rep("one",10),rep("seven",10))

perm=as.numeric()

for (i in 1:10000){
perm[i]=diff(tapply(sample(p,20,replace = F), fac, mean))
  }

hist(perm)


sum(abs(perm)>=abs(diff))/10000




```


Fever is defined as body temp of >100.4 F. Is there any relationship between the patients gender and having fever on day 7 after infection?
```{r}
covid$Fever=covid$Temp_day7>100.4
tab=table(covid$Gender,covid$Fever)

#Use fisher exact test as assumptions of chisq are not met
chisq.test(tab)
fisher.test(tab)
```
