---
title: "XDAS ANOVA Exercise"
author: "Manpreet S. Katari"
date: "11/9/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## NPK

For this exercise, we will be using the `npk` dataset that is provided in R. The experiment looked at the growth of peas and how nitrogen(N), phosphate(P), and potassium(K) had an effect.

The experiment was performed on 6 blocks, however only three blocks contains the full 3-way interaction combinations.

### 1) Look at the distribution of values from the different plots

Use a graph to visually test wheter the distributions of `yield` is the same across all 6 blocks.

```{r}
#
```

### 2) Use a model to determine which nutrient, on its own, gave the best yield.

```{r}
#
```

# Interactions

For the interaction example, we will look at a dataset called `ToothGrowth`. The dataset measures tooth group in guinea pics. Each has recieved Vitaming C either from Orange juice, `OJ`, or ascorbic acid `VC`. The two treatments were also provided in different concentrations. We will look to see if there is a something going on.

## 3) Create a chart that that shows the result based on the type and amount of treatment. Do you expect an interaction between Treatment and Dose? 

```{r}
#
```
## 4) Test to see if there is a significant interaction between the dosage of the treatment and the treatment.

```{r}

#

```

## 5) Compare the interaction model with a model without the interaction. Is there a significant difference between the two? Which model should you keep?

```{r}
#
```

```{r}

#

```

```{r}
#

```