---
title: "Two group T-test"
author: "Manpreet S. Katari"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Let's create two random values from populations that we know have different means

```{r}
sample_size=10
sampleNormValues1 = rnorm(sample_size, mean=10, sd=1)
sampleNormValues2 = rnorm(sample_size, mean=11, sd=1)

hist1=hist(sampleNormValues1, plot = F)
hist2=hist(sampleNormValues2, plot = F)

# the fourth value of the color parameter is the opacity (here 0.25)
plot(hist1, xlim=c(5,20), col=rgb(0,0,1,.25))
plot(hist2, xlim=c(5,20), col=rgb(1,0,0,.25), add=T)

```

What's the difference in sample means for these two samples?
```{r}
sample_diff=mean(sampleNormValues1) - mean(sampleNormValues2)
sample_diff
```


#### We can use the mean and sd of our random values to generate more random values to determine the distribution of difference of the means we can obtain.

```{r}
sampleMeansDiff = numeric()  # initialize an empty loop
for ( i in 1:10000) {
  # take a new sample from each of the two normal distributions above
  sampleNormValues1loop = rnorm(10, 
                                mean=mean(sampleNormValues1),
                                sd=sd(sampleNormValues1))  
  sampleNormValues2loop = rnorm(10, 
                                mean=mean(sampleNormValues2),
                                sd=sd(sampleNormValues2))
  
  sampleMeansDiff[i]=mean(sampleNormValues1loop)-mean(sampleNormValues2loop)
}

# What's the mean of the samples? 
mean(sampleMeansDiff)

# let's look at the data.
hist(sampleMeansDiff, breaks=50)
abline(v=mean(sampleMeansDiff), col="red", lwd=2)
```

```{r}
# 95% of the time the difference in the means will be between

sort(sampleMeansDiff)[.025*100000]
sort(sampleMeansDiff)[.975*100000]

```

#### The standard deviation gives us the standard error

```{r}

mean(sampleMeansDiff)
sd(sampleMeansDiff)

```

#### Instead of calculating it empirically, we can use one of the formulas.
```{r}

# sd of the difference between the means

var_normval1 = var(sampleNormValues1)
var_normval2 = var(sampleNormValues2)
normval1size = length(sampleNormValues1)
normval2size = length(sampleNormValues2)


se_diff = sqrt((var_normval1/normval1size) + (var_normval2/normval2size))

print(se_diff)
```

#### Theoretical limits

```{r}
lower_limit = sample_diff - 1.96*se_diff
higher_limit = sample_diff + 1.96*se_diff

print(lower_limit)
print(higher_limit)

```

#### T-test is the difference of the means over the standard error
```{r}
(sample_diff) / ( se_diff )

```

```{r}
t.test(sampleNormValues1, sampleNormValues2, var.equal = F)
```

#### shuffle test

```{r}
data = c(sampleNormValues1, sampleNormValues2)
expgroup = factor(rep(c("A","B"), each=sample_size))

diff(tapply(data, expgroup, mean))

shufflediff = numeric()
for ( i in 1:10000) {
  shufflediff[i]=diff(tapply(sample(data, length(data), replace = F),
                             expgroup, mean))
}

hist(shufflediff, breaks = 50)

shufflepvalue = sum(shufflediff >= abs(sample_diff))/10000
print(shufflepvalue)
```
