
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# These practice questions are based on HW5-Q1. 

1.Create a vector representing a population of 100 Y and P cells. Use 1 and 0 to represent Y and P cells in the ratio of 0.65 and 0.35 respectively
```{r}


```

2.In HW5 we sampled 6 cells from this population. Write a loop to create a sampling distribution of means based on sampling 6 cells 1000 times. Store the results as 'nulldist' and plot its histogram
```{r}


```

3.Use the sampling distribution of means to find the probability that exactly 4 out of 6 cells will be yellow. Confirm your answer with dbinom() 
```{r}


```

4.Use the sampling distribution of means to find the probability of finding 4 or more Y cells out of 6. Confirm your answer with pbinom()
```{r}


```

5.Instead of using a loop, we could have simulated a binomial distribution with the rbinom function.Use rbinom to recreate the null distribution vector containing 1000 values and use it to answer the previous question 
```{r}


```

6.We suspect that the ratio of Y and P cells is regulated by Gene X and knocking out this gene may change this ratio. To test this hypothesis, we generate Gene X knockout flies and sample 6 cells from their eyes. Our sample contains 1 Y and 5 P cells.Write the null and alternative hypothesis for this experiment and calculate the p-value of the sampling result.  

```{r}


```

7.What would be the p-value if our hypothesis was that knocking down Gene X increases the number of Y cells compared to P cells
```{r}



```

8.Assume we took 100 samples of 6 cells from two different fly populations and counted the number of Y cells in each sample. The results for this sampling are stored in the dataframe 'results'. Use the chisquared test to determine if any of these populations fit a binomial distribution with a Y:P ratio of 0.65:0.35
```{r}
results=data.frame(No_of_Ycells=seq(0,6,1),pop1_freq=c(0,2,10,24,36,19,9),pop2_freq=c(15,15,14,14,14,14,14))

#add a column called 'expected' to the dataframe containing expected frequencies


#calculate the chi-squared statistic for each population using the observed and expected populations


#check your results using the chis.test() function and look at the p-values




```


