---
title: "XDASI-2021 HW1: R Basics"
author: "YOUR NAME HERE"
date: "9/7/2021"
output: 
  html_document:
    toc: TRUE
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

As you go through each section below, you may wonder about some of the markdown syntax you see (e.g. "##", "+", etc.) in the text. If so, take a look at an R Markdown cheatsheet or the R Markdown documentation. Links to these resources are provided on the course website in the "R Resources => R/RStudio Basics" section.

## Part I: R and RStudio basics

I.1) What is an R workspace? How do you save and load workspaces manually?

```{r eval=FALSE}

```

I.2) What is a working directory? How do you set it and how do you determine which directory is the working directory?

```{r eval=FALSE}

```

I.3) Where does R save the history of your commands?

```{r eval=FALSE}

```

I.4) How is an RStudio Project related to the R workspace and working directory? What is the big advantage of using R Projects?

```{r eval=FALSE}

```

I.5) How do you start a new R script in RStudio?

```{r eval=FALSE}

```

I.6) What does it mean then R prompts you with a  "+" in the beginning of the line?

```{r eval=FALSE}

```

I.7) What are three ways you can retrieve documentation related to a function and its usage in RStudio?

```{r eval=FALSE}
# 1)
# 2) 
# 3)
```

I.8) What are the three basic components of R Markdown files?

```{r eval=FALSE}

```

I.9) What happens when you knit an R Markdown file?

```{r eval=FALSE}

```

I.10) R Markdown code blocks

```{r eval=FALSE}
# what is the keyboard shortcut to create a new code block?

# what are two ways you can run a single line in a code block?

# what are two ways you can run an entire code block at once?

# how do you prevent a code block from being executed when you knit the document?

# how do you prevent a code block from being displayed in a rendered document?

```

---

## Part II: Working with data

### II.1) Matrices

a. Create three vectors and display their contents.
  + "GeneA", with the values 3,4,5,7
  + "GeneB", with the values 4,6,8,5
  + "GeneC", with the values 5,3,9,8

```{r collapse=TRUE}
GeneA = c(3:5,7)
GeneB = c(4,6,8,5)
GeneC = c(5,3,9,8)
GeneA
GeneB
GeneC
```

b. Create a character vector with the words, "control", "nitrogen","potassium", and "phosphate". Call it "Experiments" and display the contents.
```{r}
Experiments = c("control", "nitrogen","potassium", "phosphate")
```

c. Use two different methods to create a **3x4** matrix where each row contains data for one gene (don't assign them to a variable, just make them):

  - `matrix()` - look up the documentation, and make sure you fill the matrix by *rows*, not by *columns*
  - `rbind()` - combine vectors by rows
```{r}
# method 1: use matrix()
matrix(c(GeneA,GeneB,GeneC), nrow = 3, ncol = 4, byrow = TRUE)

# method 2: use rbind()
rbind(GeneA,GeneB,GeneC)
```

d. What's the difference between the output when you use these two different methods to create the matrix?
```{r eval=FALSE}
# matrix() did not propagate the vector names because they were combined as an unnamed vector, whereas 
# rbind() used the named vectors directly and those were retained in the new data structure.
```

e. Use the `rbind()` method to make a matrix called "expvalues", and add column names to it using the "Experiments" vector. Display the resulting matrix.
```{r}
expvalues = rbind(GeneA,GeneB,GeneC)
colnames(expvalues) = Experiments
colnames(expvalues)
expvalues
```


### II.2) Dataframes

a. What's the key difference between a matrix and a data frame? In what situation would you prefer a dataframe over a matrix?
```{r eval=FALSE}
# A matrix can only contain one data type, whereas a data frame can contain column vectors of different types
# so you can mix numerical columns with character columns

# In addition, character columns can be also be designated as factors to designate rows as belonging to different experimental groups
```

b. Convert the "expvalues" matrix you just made into a data frame. Call it "expvalues.df" and display the resulting object. Also, take a look at the `class()` and `structure()` of the object.
```{r collapse=TRUE}
# convert to df
expvalues.df = as.data.frame(expvalues) # explicit conversion
expvalues.df = data.frame(expvalues)    # also works
expvalues.df

# check the class and structure
class(expvalues.df)
str(expvalues.df)
```

c. Use two different methods to view just the data from the "potassium" column.
```{r collapse=TRUE}
# a) square brackets notation
expvalues.df[,3]

# b) dollar sign notation
expvalues.df$potassium
```

d. Get just the control and nitrogen values for GeneA from "expvalues.df".
```{r collapse=TRUE}
# a) select desired columns using numerical indices
expvalues.df[1,1:2]
expvalues.df[1,c(1,2)]

# b) alternatively, you can exclude unwanted columns using -c()
expvalues.df[1,-c(3:4)]
```

e. You can also select rows from data frames based on their values using the `which()` function with logical expressions!
```{r collapse=TRUE}
# select rows whose values satisfy specific criteria
# here we are asking for genes whose expression in the potassium condition is above 7
interesting_genes = which(expvalues.df$potassium > 7)
interesting_genes
expvalues.df[interesting_genes,]

# do it all at once
expvalues.df[which(expvalues.df$potassium > 7),]
```

---

## Part III: Data Import / Export

a. Inside RStudio (not in your file system):
  + get your current working directory
  + create a new directory inside of it called "data" using `dir.create()`
    + Note: you will get a warning if you run this more than once, but it's ok
  + list the contents of the directory using either `dir()` or `list.files()`
```{r collapse=TRUE}
getwd()
dir.create("data")
dir()
```

b. Download the data file called "hw1_fly_phenotypes.tsv" and put it into the data directory you just created. (Don't try to use RStudio for this, just use your filesytem.) 

  - First, check that the file is where you think it is from within RStudio using the appropriate command to list all the files in the data directory.
```{r}
list.files("data")
```

  - Next, look at the documentation for `read.table()` and compare the default settings for the different variations of this function.
  - Decide which function you will use, paying attention to the default parameter settings for `header`,  `sep`, and `stringsAsFactors`. Note that any of the arguments can be overridden by explicitly specifying them, but it's easier to just pick the function with the right defaults. Some things to think about: 
    - Does the file contain column headers? 
    - What are the column delimiters? 
    - What kind of data does the file contain? 
    - Do you want to make factors right away?

  - Once you've decided on how you will import the data, read in the data file using one of these functions and store it in a data frame called "fly_pheno".
   - You should do this *without changing your current working directory* from within RStudio. (Hint: You can use a relative path name to do this.)
```{r}
#  make factors during import
fly_pheno = read.delim("data/hw1_fly_phenotypes.tsv", stringsAsFactors = TRUE)

# make factors later
#fly_pheno = read.delim("data/hw1_fly_phenotypes.tsv", stringsAsFactors = FALSE)
```

  - Finally, take a look at the structure of the object you have created using `str()`. 
  - Also, use the `head()` command to just take a peek at the first few lines of the file (it's pretty long). Does everything look ok? 
```{r}
str(fly_pheno)   # structure
head(fly_pheno)  # contents
```

c. Rename the column labels by making them into lower-case font. You can do this using the `tolower()` function. Check to see if this worked.
```{r}
colnames(fly_pheno) = tolower(colnames(fly_pheno))
head(fly_pheno)
```

d. The eye colors have been encoded as three-letter strings. This seems unnecessarily cryptic, so change the strings for white eyes back to "WHITE". 

Depending on whether you set `stringsToFactors` as `TRUE` or `FALSE` when you imported the data, you will need to choose a different method to do this:

- `TRUE`:  use the `levels()` command
  - this will automatically change both the factors and the corresponding values in the column vector
  - WARNING: if you use the `levels()` command on a column that is not already a factor, you will be setting levels, but you won't actually change the values in the column.
  
- `FALSE`: use the `which()` command
  - find elements in the "eye_color" vector whose value is "WHI" (see Part II.d), and then reassign them as "WHITE"
  - WARNING: if you try to use `which()` and the column is already a factor, then if "WHITE" is not already one of the factor levels, all of the elements you tried to set to "WHITE" will now contain NA's! (R will show you a warning if this happens)

Check the structure of the data frame again and take a look at the first few lines of the file to make sure everything looks ok.
```{r}
# If you used stringsToFactors = TRUE
levels(fly_pheno$eye_color) = c("RED","WHITE")

# If you used stringsToFactors = FALSE
  # if "eye_color" is a factor, this will produce an error unless you've already changed "WHI" to "WHITE"
  # otherwise it won't do anything
fly_pheno[which(fly_pheno$eye_color == "WHI"),2] = "WHITE"

str(fly_pheno)
head(fly_pheno)
```

e. If it isn't already, convert the "sex" column into a factor. Then, make a table that summarizes the number of flies of each sex that have red or white eyes.
```{r}
# make "sex" a factor (this command won't change anything if it's already a factor)
fly_pheno$sex = factor(fly_pheno$sex)
levels(fly_pheno$sex)
str(fly_pheno)

# make a table
table(fly_pheno)
```

f. Great! You now have a data frame that you like! Let's save the massaged data in the "data" directory. We will work with this dataset again in a future homework.

  + First write it to a **comma-delimited** text file with the file extension ".csv". Set `row.names = FALSE`.
  + Also save it as an **RDS** object in a file with the "rds" extension.
  + Just be sure, check the contents of the "data" directory.
```{r}
write.csv(fly_pheno, file="data/fly_pheno.csv", row.names = FALSE)
saveRDS(fly_pheno, file="data/fly_pheno.rds")
dir("data")
```

g. Now you are ready to knit your document to hand it in. 

+ Sometimes, if you've been working on something for a while, you might be executing some commands on variables that do not contain what you think they do (e.g. because you made mistakes in naming something, and/or assigned the wrong data to a particular variable and failed to correct it).
+ First, check the variables that you have in your current environment using `ls()`.
+ Next, to make sure everything is running the way you expect it to, remove all of the variables in your environment by running the command `rm(list = ls())` and check the result using `ls()`.

+ Finally, knit the file using the "Knit" button at the top of the editor window and review the html output that is created. 
  + Are you getting any errors? If so, you will see a message indicating where things went wrong in the "R Markdown" pane (another tab in the window with the Console that opens when you try to knit). 
  + If your file won't knit, go back and try to correct the mistakes you made above, then remove all your environment variables and try again.
  + If your document still won't knit, feel free to reach out to us for help, or ask one of your classmates if they can detect what's wrong with your code.

```{r}
ls()
rm(list = ls())
ls()
```

**Voilà! First homework exercise completed! Yay!!! You may now submit your final .Rmd file through Brightspace.**
