---
title: "Paired Test example"
author: "Manpreet S. Katari"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Sleep dataset.

Simply type the name of the dataset, **sleep**, and you will be able to access it in our R workspace.

```{r}
data = sleep
```

Two drugs were studied to see their effects on sleep. The column *extra* contains the number of hours of sleep that were different from when the patient was not given a drug. The column *group* defines the different medications. The column *ID* defines the individuals. The two different drugs were tested on same patients at different times.

### 1)Determine whether the *extra* hours of sleep from group1 is normally distributed. Do the same for the *extra* sleep for group2.

```{r}
group1 = subset(data, group==1)
group2 = subset(data, group==2)
shapiro.test(group1$extra)
shapiro.test(group2$extra)


```


### 2)Determine whether the variance of the *extra* hours of sleep from the two different groups is equal. Check out documentation for **var.test()**. Explain what the results of the command means.

```{r}
var.test(group1$extra, group2$extra)

```

### 3)Is the dataset paired ? Explain.

```{r}
# yes
```

### 4)Based on your answers for 1-3, use the appropriate statistical test to determine if the two drugs have the same effect on amount of sleep the patients get. What is your conclusion ? Explain your answer.

```{r}

t.test(group1$extra, group2$extra, paired=T, var.equal = T)

t.test(data$extra ~ data$group, paired=T, var.equal=T)

```


### 5)Draw a plot that demonstrates your conclusions are correct. *EXPLAIN* your results.

```{r}
# if the command has a data option, it usually means you can refer to the
# column by its name
boxplot(extra ~ group, data=sleep)

histgroup1 = hist(group1$extra, col="lightgreen", xlim=c(-7,7))
histgroup2 = hist(group2$extra, col="lightblue", add=T, alpha=.5)

```
here
### 6)Since the extra defines the change in sleep, we can assume that normally (without any medication) the patients get an average of 0 extra hours of sleep. So if we assume that **population** average is 0, use a statistical test to show which of the two drugs is effective in changing the number of hours of sleep.

```{r}
t.test(group1$extra, mu=0)
t.test(group2$extra, mu=0)
```

### 7)For the group that is statistically different from the population mean, calculate the mean and stand error.

```{r}
sd(group2$extra)/sqrt(length(group2$extra))
mean(group2$extra)
  
```

### 8)If you divide the mean by the standard error, what statistic have you calculated? **Hint, check the results from q 6.**

```{r}
mean(group2$extra) / (sd(group2$extra)/sqrt(length(group2$extra)))

```

