---
title: "Exercise: Chi-Square Test"
subtitle: "XDASI Fall 2021"
date: "10/4/2021"
output:
  html_document:
    toc: no
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Effect of aspirin on heart attack risk

22,071 physicians served as subjects to study the effect of aspirin on the incidence of heart attacks. There were two groups: 

+ 11,028 subjects took aspirin regularly for 5 years
+ 11,043 received a placebo instead of aspirin 

At the end of the study, 189 in the placebo group and 104 of those who took aspirin had suffered a heart attack. 

***Does aspirin have a significant association with physicians suffering from a heart attack?***

### 1. Create a contingency table for these data.

```{r}
# data (quadrants a-d)
placebo_heart   =     # a: took placebo and heart attack
aspirin_heart   =     # b: took aspirin and heart attack
placebo_healthy =     # c: took placebo and healthy
aspirin_healthy =     # d: took aspirin and healthy

placebo_total =       # column 1 margin
aspirin_total =       # column 2 margin

heart_total   =       # row 1 margin
healthy_total =       # row 2 margin

Observed = 

Observed
```

### 2. Calculate the expected values for each group.

```{r}
N =   # total

placebo_heart_expected   = 
aspirin_heart_expected   = 
placebo_healthy_expected = 
aspirin_healthy_expected = 

Expected = 
  
Expected
```

### 3. Calculate the $X^2$ test statistic using observed and expected counts.

```{r}
chisq = 
chisq
```


### 4. Compute the $p$-value of the test statistic using the $\chi^2$ CDF.

```{r}

```

### 5. Calculate the $p$-value using the `chisq.test()` function.

```{r}

```
### 6. Apply Yates' continuity correction to your manual calculation of the $X^2$ test statistic.

```{r}
new.chisq = 
new.chisq
```

### 7. Recalculate the $p$-value using `pchisq()`.

```{r}

```

### 8. Why did we need to apply Yates' correction?

```{r}
# your answer here
```

