---
title: 'Homework 4: Probability'
subtitle: "XDASI Fall 2021"
author: "Sir Ronald Fisher [YOUR NAME HERE]"
date: "Due: 9/30/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
library(ggplot2)
library(gmodels)
# remove "error = TRUE" to make knitr halt on errors
# (do this when you are ready to check if your code will knit without errors)
```

## 1. Probability density and cumulative probability

You are screening for genetic interactions between genes in yeast that cause deleterious fitness effects. To establish a baseline for your experiments, you decide to first estimate the probability density of colony diameters (in mm) for a population of WT yeast grown in controlled conditions (48 hours at 37°C). Fortunately, you have an automated image recognition system that allows you to do this for a large number of individual colonies!

a. Simulate a sample of 1000 measured colonies with an average diameter of 1mm and a standard deviation of 0.1mm and store it in a variable. Set a seed to make your work reproducible. 

What are the actual mean and SD? What about the IQR? Draw a boxplot to get a feel for the sample distribution.

```{r}
# your answer here
```

b. What are the SEM and 95%CI for this sample? How are these summary statistics computed, and what do they mean?

```{r}
# Option 1: compute by hand


# Option 2: using gmodels package


# Explanation

```


c. Draw a histogram of the sampled data, and make a qqplot to confirm that the sample really comes from a normal distribution.

*Extra credit: See if you can overlay some additional information on the histogram, such as vertical lines showing some statistics, a smoothened density line for your sample distribution, and/or an "ideal" normal curve with the same parameters.*

```{r}
# Histograms

# Option 1: base R

# Option 2: ggplot


# qqplot

```


Use the R functions for the CDF (`pnorm()`) and inverse CDF (`qnorm()`) to answer the questions below.

d. What is the total probability that a randomly sampled colony from the parent population measures between 0.93 and 1.07 mm?

```{r}
# your answer here
```


e. What is the proportion of data that are <1 or >1 standard deviations away from the mean?

```{r}
# your answer here
```


f. What diameters correspond to the 10% and 90% quantiles?

```{r}
# your answer here
```



## 2. Co-occurrence of infection

Among women voluntarily tested for sexually transmitted diseases in one university (modified from Tábora et al. 2005):

+ 18% tested positive for human papilloma virus (HPV) only,
+ 2% tested positive for Chlamydia only, and
+ 3% tested positive for both HPV and Chlamydia. 

a. What is the total frequency of each STD in the population?

```{r}
# given probabilities
p.hpv = 0.18
p.chl = 0.02
p.hpv.chl = 0.03

# your answer here
```


For the next 3 questions, use set notation to write out the question in terms of set operations (as a comment) and then compute the answer. You may write probabilities using the following syntax, e.g.:

+ P(A) * P(B)
+ P(A) + P(B)
+ P(A AND B)
+ P(A OR B)
+ P(A | B)
+ etc.

b. What is the probability that a woman will test positive for Chlamydia, given that she is positive for HPV?

```{r}
# notation

# computation

```


b. What is the probability that a woman who tests positive for Chlamydia will also test positive HPV?

```{r}
# notation

# computation

```


c. What is the total probability that a woman from this population who gets tested would test positive for either HPV and/or Chlamydia?

```{r}
# notation

# computation

```


d. Is the occurrence of these infections independent? Explain.

```{r}
# your answer here
```


## 3. Allele frequencies

Many gene loci have a major allele and a number of minor alleles in the population. Let's say there are 5 alleles for a particular locus, that individuals mate randomly with respect to the genotype at this locus, and that allele frequencies for A1-A5 are 0.82, 0.06, 0.05, 0.04, and 0.03 respectively.

For each of the questions below, write out the probability equations (e.g. P(Ai) for each term) and then compute the answers numerically.

a. What is the probability that any single allele chosen at random from the population is either A2 or A5? Write out the probability statement as a comment below and then computer the answer.

```{r}
# Allele frequencies
A1 = 0.82
A2 = 0.06
A3 = 0.05
A4 = 0.04
A5 = 0.03

# notation

# computation

```

b. What is the probability that an individual carries two A3 alleles?

```{r}
# notation

# computation

```

c. What is the probability that someone does NOT carry two A3 alleles?

```{r}
# notation

# computation

```

d. What is the probability that someone is heterozygous for A1 and A4?

```{r}
# notation

# computation

```

e. What is the probability that *neither* of two random individuals in the population would carry two A1 alleles?

```{r}
# notation

# computation

```

f. ***Extra credit:*** What is the probability that 3 random individuals carry no A3 or A4 alleles at all? (Remember that each individual carries two alleles.) 

```{r}
# notation

# computation

```
