---
title: 'XDASI-2021 HW1: R Basics'
author: "MARI SHIRATORI"
date: "9/7/2021"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    toc: yes
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

As you go through each section below, you may wonder about some of the markdown syntax you see (e.g. "##", "+", etc.) in the text. If so, take a look at an R Markdown cheatsheet or the R Markdown documentation. Links to these resources are provided on the course website in the "R Resources => R/RStudio Basics" section.

## Part I: R and RStudio basics

I.1) What is an R workspace? How do you save and load workspaces manually?

```{r eval=FALSE}
An R workspace is the collection of all of the objects you've assigned during an R session, such as vectors, dataframes, matrices, etc, that can be saved and reloaded.

In order to save a workplace, you can go to the Session -> Save workplace as... in the upper toolbar. Alternatively, you can also type in your R console: save(file = "d:/specificFileName.RData")
save(file = "d:/Week1_Homework.RData")

In order to load workspaces, you can go to Session -> Load Workspace... in the upper toolbar. Alternatively, you can type into the R console: load(file = "d:/specificFileName.RData")

```

I.2) What is a working directory? How do you set it and how do you determine which directory is the working directory?

```{r eval=FALSE}
# The working directory is the directory or file path that your R session is drawing from and saving from during a particular command.

# In order to find what your working directory is, you can use
getwd()

# In order to set your working directory, you can use
setwd()

Within the parentheses above, you would want to add the specific file path you want to use. For instance, if I wanted to set my wd to my desktop, I could enter the following into the console:
  
setwd("/Users/maricobb/Desktop")

```

I.3) Where does R save the history of your commands?

```{r eval=FALSE}
R saves your history of commands in your environment in the history tab. By default, this is the upper right window within the RStudio main window.
```

I.4) How is an RStudio Project related to the R workspace and working directory? What is the big advantage of using R Projects?

```{r eval=FALSE}
An RStudio project organizes your R workspaces and associated outputs nd files within a particular working directory. The main advantage is that all of your pertinent files will be in one space.
```

I.5) How do you start a new R script in RStudio?

```{r eval=FALSE}
You can either type shift+command+N, or go to File -> New File -> R Script in the upper toolbar.
```

I.6) What does it mean then R prompts you with a  "+" in the beginning of the line?

```{r eval=FALSE}
The "+" at the beginning of the line means that your command was not complete and thus not executed. This could be due to a missing necessary argument or a missing parentheses.
```

I.7) What are three ways you can retrieve documentation related to a function and its usage in RStudio?

```{r eval=FALSE}
1) You can type ?function() into the console
2) You can search the function name into the searchbar in the Help section of the Files/Plots/Packages/Help window
3) When you type in your function in the console, R will offer autofill suggestions on what it thinks you're trying to type in and if you mouse over these options, it will provide basic information about that function.
```

I.8) What are the three basic components of R Markdown files?

```{r eval=FALSE}
There are 1) code blocks, which are the the code parts written within three backticks (```) at the beginning and end of the section. After the first set of backticks is also a part that includes {r } where the /'r' is describing the language and other specifications like eval = FALSE which basically tells knitr to not run the code in the code block.

Then there is 2) metadata, which is the header at the beginning of the .RMD file that contains the title, author, output file specifications. The metadata is sandwiched between a pair of 3 dashes (---).

Finally there is the 3) text, which is the text outside of the code blocks. 
```

I.9) What happens when you knit an R Markdown file?

```{r eval=FALSE}
When you knit, you compile the .RMD file into a .pdf, html, or word doc so it can be easily read.
```

I.10) R Markdown code blocks

```{r eval=FALSE}
# what is the keyboard shortcut to create a new code block?
Command + option + I on macs

# what are two ways you can run a single line in a code block?
You can either use Command + return or highlight a line of code, and click "Run Selected line(s)" in the Source window tool bar in R Studio

# what are two ways you can run an entire code block at once?
You can either use Command + option + R, or click "Run All" in the Source window tool bar in R Studio

# how do you prevent a code block from being executed when you knit the document?
At the top of the code block, you can type r eval = FALSE within the curly braces after the first set of backticks at the start of the code block.

# how do you prevent a code block from being displayed in a rendered document?
You can type echo = FALSE within the curly braces after the first set of backticks at the start of the code block.

```

---

## Part II: Working with data

### II.1) Matrices

a. Create three vectors and display their contents.
  + "GeneA", with the values 3,4,5,7
  + "GeneB", with the values 4,6,8,5
  + "GeneC", with the values 5,3,9,8

```{r collapse=TRUE}
# create the vectors
GeneA <- c(3, 4, 5, 7)
GeneB <- c(4, 6, 8, 5)
GeneC <- c(5, 3, 9, 8)

# take a look at them
GeneA
GeneB
GeneC

```

b. Create a character vector with the words, "control", "nitrogen","potassium", and "phosphate". Call it "Experiments" and display the contents.
```{r}
Experiments <- c("control", "nitrogen", "potassium", "phosphate")
Experiments

```

c. Use two different methods to create a **3x4** matrix where each row contains data for one gene (don't assign them to a variable, just make them):

  - `matrix()` - look up the documentation, and make sure you fill the matrix by *rows*, not by *columns*
  - `rbind()` - combine vectors by rows
```{r}
# method 1: use matrix()
matrix(c(GeneA, GeneB, GeneC), nrow = 3, ncol = 4, byrow = TRUE)

# method 2: use rbind()
rbind(GeneA, GeneB, GeneC)

```

d. What's the difference between the output when you use these two different methods to create the matrix?
```{r eval=FALSE}
# rbind() outputs row names along side of the matrix (i.e. GeneA, GeneB, and GeneC), while the output from matrix() just labels the row names as its index position (i.e. [1,], [2,], [3,])

```

e. Use the `rbind()` method to make a matrix called "expvalues", and add column names to it using the "Experiments" vector. Display the resulting matrix.
```{r}
# make the matrix "expvalues"
expvalues <- rbind(GeneA, GeneB, GeneC)
# assign column names
colnames(expvalues) <- Experiments
# check your work
expvalues
```


### II.2) Dataframes

a. What's the key difference between a matrix and a data frame? In what situation would you prefer a dataframe over a matrix?
```{r eval=FALSE}
# A matrix can only hold 1 type of data while a dataframe can contain multiple, thus in situations where you have multiple types of data (numbers, strings, logicals, etc) that you want to put together in the same matrix, you would use dataframes instead of matrices.
```

b. Convert the "expvalues" matrix you just made into a data frame. Call it "expvalues.df" and display the resulting object. Also, take a look at the `class()` and `structure()` of the object.
```{r collapse=TRUE}
# convert to df
expvalues.df <- as.data.frame(expvalues)
# check the class and structure
class(expvalues.df)
str(expvalues.df)

```

c. Use two different methods to view just the data from the "potassium" column.
```{r collapse=TRUE}
# a) square brackets notation
expvalues.df[,3]

# b) dollar sign notation
expvalues.df$potassium

```

d. Get just the control and nitrogen values for GeneA from "expvalues.df".
```{r collapse=TRUE}
# a) select desired columns using numerical indices
expvalues.df[1, 1:2]

# b) alternatively, you can exclude unwanted columns using -c()
expvalues.df[1, -c(3, 4)]

```

e. You can also select rows from data frames based on their values using the `which()` function with logical expressions!
```{r collapse=TRUE}
# select rows whose values satisfy specific criteria
# here we are asking for genes whose expression in the potassium condition is above 7
interesting_genes = which(expvalues.df$potassium > 7)
interesting_genes
expvalues.df[interesting_genes,]

# do it all at once
expvalues.df[which(expvalues.df$potassium > 7),]
```

---

## Part III: Data Import / Export

a. Inside RStudio (not in your file system):
  + get your current working directory
  + create a new directory inside of it called "data" using `dir.create()`
    + Note: you will get a warning if you run this more than once, but it's ok
  + list the contents of the directory using either `dir()` or `list.files()`
```{r collapse=TRUE}
getwd()
setwd("/Users/maricobb/Desktop/Mari/NewYork/BioStats/BioStats2021/Week1")
dir.create("/Users/maricobb/Desktop/Mari/NewYork/BioStats/BioStats2021/Week1/data")
```

b. Download the data file called "hw1_fly_phenotypes.tsv" and put it into the data directory you just created. (Don't try to use RStudio for this, just use your filesytem.) 

  - First, check that the file is where you think it is from within RStudio using the appropriate command to list all the files in the data directory.
```{r}
list.files(path = "/Users/maricobb/Desktop/Mari/NewYork/BioStats/BioStats2021/Week1/data")
```

  - Next, look at the documentation for `read.table()` and compare the default settings for the different variations of this function.
  - Decide which function you will use, paying attention to the default parameter settings for `header`,  `sep`, and `stringsAsFactors`. Note that any of the arguments can be overridden by explicitly specifying them, but it's easier to just pick the function with the right defaults. Some things to think about: 
    - Does the file contain column headers? 
    - What are the column delimiters? 
    - What kind of data does the file contain? 
    - Do you want to make factors right away?

  - Once you've decided on how you will import the data, read in the data file using one of these functions and store it in a data frame called "fly_pheno".
   - You should do this *without changing your current working directory* from within RStudio. (Hint: You can use a relative path name to do this.)
```{r}
fly_pheno <- read.table("./data/hw1_fly_phenotypes.tsv", header = TRUE, stringsAsFactors = TRUE)

head(fly_pheno)
```

  - Finally, take a look at the structure of the object you have created using `str()`. 
  - Also, use the `head()` command to just take a peek at the first few lines of the file (it's pretty long). Does everything look ok? 
```{r}
str(fly_pheno)
head(fly_pheno)
```

c. Rename the column labels by making them into lower-case font. You can do this using the `tolower()` function. Check to see if this worked.
```{r}
colnames(fly_pheno) <- tolower(colnames(fly_pheno))
head(fly_pheno)
```

d. The eye colors have been encoded as three-letter strings. This seems unnecessarily cryptic, so change the strings for white eyes back to "WHITE". 


Depending on whether you set `stringsToFactors` as `TRUE` or `FALSE` when you imported the data, you will need to choose a different method to do this:

- `TRUE`:  use the `levels()` command
  - this will automatically change both the factors and the corresponding values in the column vector
  - WARNING: if you use the `levels()` command on a column that is not already a factor, you will be setting levels, but you won't actually change the values in the column.
  
- `FALSE`: use the `which()` command
  - find elements in the "eye_color" vector whose value is "WHI" (see Part II.d), and then reassign them as "WHITE"
  - WARNING: if you try to use `which()` and the column is already a factor, then if "WHITE" is not already one of the factor levels, all of the elements you tried to set to "WHITE" will now contain NA's! (R will show you a warning if this happens)

Check the structure of the data frame again and take a look at the first few lines of the file to make sure everything looks ok.
```{r}
# If you used stringsToFactors = TRUE
levels(fly_pheno$eye_color) <- c("RED", "WHITE")

# If you used stringsToFactors = FALSE

# check what happened
head(fly_pheno)

```

e. If it isn't already, convert the "sex" column into a factor. Then, make a table that summarizes the number of flies of each sex that have red or white eyes.
```{r}
# make "sex" a factor if not already
# Already a factor

# check your work
head(fly_pheno)

# make a table
table(fly_pheno)
```

f. Great! You now have a data frame that you like! Let's save the massaged data in the "data" directory. We will work with this dataset again in a future homework.

  + First write it to a **comma-delimited** text file with the file extension ".csv". Set `row.names = FALSE`.
  + Also save it as an **RDS** object in a file with the "rds" extension.
  + Just be sure, check the contents of the "data" directory.
```{r}
# CSV file
write.csv(fly_pheno,"./data/Week1_Homework_fly.csv")
# RDS file
saveRDS(fly_pheno,"./data/Week1_Homework_fly.RDS")
# Check what's in the "data" directory
list.files("./data")

```

g. Now you are ready to knit your document to hand it in. 

+ Sometimes, if you've been working on something for a while, you might be executing some commands on variables that do not contain what you think they do (e.g. because you made mistakes in naming something, and/or assigned the wrong data to a particular variable and failed to correct it).
+ First, check the variables that you have in your current environment using `ls()`.
+ Next, to make sure everything is running the way you expect it to, remove all of the variables in your environment by running the command `rm(list = ls())` and check the result using `ls()`.

+ Finally, knit the file using the "Knit" button at the top of the editor window and review the html output that is created. 
  + Are you getting any errors? If so, you will see a message indicating where things went wrong in the "R Markdown" pane (another tab in the window with the Console that opens when you try to knit). 
  + If your file won't knit, go back and try to correct the mistakes you made above, then remove all your environment variables and try again.
  + If your document still won't knit, feel free to reach out to us for help, or ask one of your classmates if they can detect what's wrong with your code.

```{r}
# check the objects in the current environment
ls()
# remove all of them
rm(list = ls())
# check to make sure they are all gone
ls()
```

**Voilà! First homework exercise completed! Yay!!! You may now submit your final .Rmd file through Brightspace.**
