---
title: "XDASI-2021 HW1: R Basics"
author: "Graeme Vissers"
date: "9/7/2021"
output: 
  html_document:
    toc: TRUE
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

As you go through each section below, you may wonder about some of the markdown syntax you see (e.g. "##", "+", etc.) in the text. If so, take a look at an R Markdown cheatsheet or the R Markdown documentation. Links to these resources are provided on the course website in the "R Resources => R/RStudio Basics" section.

## Part I: R and RStudio basics

I.1) What is an R workspace? How do you save and load workspaces manually?

```{r eval=FALSE}
# A workspace contains all of the R objects that you create in a session and is contained in a file with a .Rdata extension. You can save or load workspaces with the following commands:
save.image()
load(".Rdata")
```

I.2) What is a working directory? How do you set it and how do you determine which directory is the working directory?

```{r eval=FALSE}
# A working directory is the location within your computer's file system that your project's features are saved. You can set the current working directory with:
setwd()

# You can determine your current working directory with the following command:
getwd()
```

I.3) Where does R save the history of your commands?

```{r eval=FALSE}
# a .Rhistory file in your working directory.
```

I.4) How is an RStudio Project related to the R workspace and working directory? What is the big advantage of using R Projects?

```{r eval=FALSE}
# Within a workspace, there are many files that are associated with a project. An R Project will organize these files all in one place and set the working directory to the folder of the project.
```

I.5) How do you start a new R script in RStudio?

```{r eval=FALSE}
# Shift + command + N
```

I.6) What does it mean then R prompts you with a  "+" in the beginning of the line?

```{r eval=FALSE}
# R assumes you are still typing out your command and hasn't attempted the command yet. It's still "listening".
```

I.7) What are three ways you can retrieve documentation related to a function and its usage in RStudio?

```{r eval=FALSE}
# 1) Type help("Your function here")
# 2) Search for the function of interest in the help search bar
# 3) Type ?Your function here
```

I.8) What are the three basic components of R Markdown files?

```{r eval=FALSE}
# 1) Readable text that demonstrates your work and thoughts
# 2) Embedded code
# 3) Metadata
```

I.9) What happens when you knit an R Markdown file?

```{r eval=FALSE}
# An output document will be created that will show a preview of what the .Rmd file looks like. The output format is specified in the YAML at the top of the .Rmd document.
```

I.10) R Markdown code blocks

```{r eval=FALSE}
# what is the keyboard shortcut to create a new code block?
# Command + option + I

# what are two ways you can run a single line in a code block?
# 1) Command + enter on the line that your cursor is on
# 2) Click run selected line(s) in the toolbar at the top

# what are two ways you can run an entire code block at once?
# 1) Shift + command + enter
# 2) Click "Run Current Chunk" in the toolbar

# how do you prevent a code block from being executed when you knit the document?
# Write 'eval=FALSE' in the header of the code chunk

# how do you prevent a code block from being displayed in a rendered document?
# Write 'echo=FALSE' in the header of the code chunk

```

---

## Part II: Working with data

### II.1) Matrices

a. Create three vectors and display their contents.
  + "GeneA", with the values 3,4,5,7
  + "GeneB", with the values 4,6,8,5
  + "GeneC", with the values 5,3,9,8

```{r collapse=TRUE}
# create the vectors
GeneA = c(3, 4, 5, 7)
GeneB = c(4, 6, 8, 5)
GeneC = c(5, 3, 9, 8)

# take a look at them
GeneA
GeneB
GeneC

```

b. Create a character vector with the words, "control", "nitrogen","potassium", and "phosphate". Call it "Experiments" and display the contents.
```{r}
Experiments = c("control", "nitrogen", "potassium", "phosphate")
Experiments
```

c. Use two different methods to create a **3x4** matrix where each row contains data for one gene (don't assign them to a variable, just make them):

  - `matrix()` - look up the documentation, and make sure you fill the matrix by *rows*, not by *columns*
  - `rbind()` - combine vectors by rows
```{r}
# method 1: use matrix()
mat1 = matrix(data=c(GeneA, GeneB, GeneC), nrow=3, ncol=4, byrow=TRUE)
mat1

# method 2: use rbind()
mat2 = rbind(GeneA, GeneB, GeneC)
mat2

```

d. What's the difference between the output when you use these two different methods to create the matrix?
```{r eval=FALSE}
# Using rbind() will automatically lable the rows based on the parameters that you used. Each row is labeled GeneA, GeneB, or GeneC
```

e. Use the `rbind()` method to make a matrix called "expvalues", and add column names to it using the "Experiments" vector. Display the resulting matrix.
```{r}
# make the matrix "expvalues"
expvalues = rbind(GeneA, GeneB, GeneC)

# assign column names
colnames(expvalues) = Experiments

# check your work
expvalues

```


### II.2) Dataframes

a. What's the key difference between a matrix and a data frame? In what situation would you prefer a dataframe over a matrix?
```{r eval=FALSE}
# Matrices can only store one data type while dataframes can store multiple. Dataframes may be useful when plotting real world data that contains multiple data types.
```

b. Convert the "expvalues" matrix you just made into a data frame. Call it "expvalues.df" and display the resulting object. Also, take a look at the `class()` and `structure()` of the object.
```{r collapse=TRUE}
# convert to df
expvalues.df = data.frame(expvalues)

# check the class and structure
structure(expvalues.df)
class(expvalues.df)

```

c. Use two different methods to view just the data from the "potassium" column.
```{r collapse=TRUE}
# a) square brackets notation
expvalues.df["potassium"]

# b) dollar sign notation
expvalues.df$potassium

```

d. Get just the control and nitrogen values for GeneA from "expvalues.df".
```{r collapse=TRUE}
# a) select desired columns using numerical indices
expvalues.df[1, 1:2]

# b) alternatively, you can exclude unwanted columns using -c()
expvalues.df[1, -c(3, 4)]

```

e. You can also select rows from data frames based on their values using the `which()` function with logical expressions!
```{r collapse=TRUE}
# select rows whose values satisfy specific criteria
# here we are asking for genes whose expression in the potassium condition is above 7
interesting_genes = which(expvalues.df$potassium > 7)
interesting_genes
expvalues.df[interesting_genes,]

# do it all at once
expvalues.df[which(expvalues.df$potassium > 7),]
```

---

## Part III: Data Import / Export

a. Inside RStudio (not in your file system):
  + get your current working directory
  + create a new directory inside of it called "data" using `dir.create()`
    + Note: you will get a warning if you run this more than once, but it's ok
  + list the contents of the directory using either `dir()` or `list.files()`
```{r collapse=TRUE}
getwd()
dir.create("data")
dir()
```

b. Download the data file called "hw1_fly_phenotypes.tsv" and put it into the data directory you just created. (Don't try to use RStudio for this, just use your filesytem.) 

  - First, check that the file is where you think it is from within RStudio using the appropriate command to list all the files in the data directory.
```{r}
list.files(path="./data")
```

  - Next, look at the documentation for `read.table()` and compare the default settings for the different variations of this function.
  - Decide which function you will use, paying attention to the default parameter settings for `header`,  `sep`, and `stringsAsFactors`. Note that any of the arguments can be overridden by explicitly specifying them, but it's easier to just pick the function with the right defaults. Some things to think about: 
    - Does the file contain column headers? 
    - What are the column delimiters? 
    - What kind of data does the file contain? 
    - Do you want to make factors right away?

  - Once you've decided on how you will import the data, read in the data file using one of these functions and store it in a data frame called "fly_pheno".
   - You should do this *without changing your current working directory* from within RStudio. (Hint: You can use a relative path name to do this.)
```{r}
help(read.table)
fly_pheno = read.table(file="./data/hw1_fly_phenotypes.tsv", header = TRUE)
```

  - Finally, take a look at the structure of the object you have created using `str()`. 
  - Also, use the `head()` command to just take a peek at the first few lines of the file (it's pretty long). Does everything look ok? 
```{r}
str(fly_pheno)
head(fly_pheno)
```

c. Rename the column labels by making them into lower-case font. You can do this using the `tolower()` function. Check to see if this worked.
```{r}
colnames(fly_pheno) = tolower(colnames(fly_pheno))
head(fly_pheno)
```

d. The eye colors have been encoded as three-letter strings. This seems unnecessarily cryptic, so change the strings for white eyes back to "WHITE". 

Depending on whether you set `stringsToFactors` as `TRUE` or `FALSE` when you imported the data, you will need to choose a different method to do this:

- `TRUE`:  use the `levels()` command
  - this will automatically change both the factors and the corresponding values in the column vector
  - WARNING: if you use the `levels()` command on a column that is not already a factor, you will be setting levels, but you won't actually change the values in the column.
  
- `FALSE`: use the `which()` command
  - find elements in the "eye_color" vector whose value is "WHI" (see Part II.d), and then reassign them as "WHITE"
  - WARNING: if you try to use `which()` and the column is already a factor, then if "WHITE" is not already one of the factor levels, all of the elements you tried to set to "WHITE" will now contain NA's! (R will show you a warning if this happens)

Check the structure of the data frame again and take a look at the first few lines of the file to make sure everything looks ok.
```{r}
# If you used stringsToFactors = TRUE

# If you used stringsToFactors = FALSE
fly_pheno[which(fly_pheno$eye_color == "WHI"), "eye_color"] = "WHITE"

# check what happened
head(fly_pheno)
```

e. If it isn't already, convert the "sex" column into a factor. Then, make a table that summarizes the number of flies of each sex that have red or white eyes.
```{r}
# make "sex" a factor if not already
fly_pheno$sex = factor(fly_pheno$sex)

# check your work
head(fly_pheno)

# make a table
table(fly_pheno)

```

f. Great! You now have a data frame that you like! Let's save the massaged data in the "data" directory. We will work with this dataset again in a future homework.

  + First write it to a **comma-delimited** text file with the file extension ".csv". Set `row.names = FALSE`.
  + Also save it as an **RDS** object in a file with the "rds" extension.
  + Just be sure, check the contents of the "data" directory.
```{r}
# CSV file
save(fly_pheno, file="./data/fly_pheno.csv")

# RDS file
save(fly_pheno, file = "./data/fly_pheno.rds")

# Check what's in the "data" directory
list.files("./data")

```

g. Now you are ready to knit your document to hand it in. 

+ Sometimes, if you've been working on something for a while, you might be executing some commands on variables that do not contain what you think they do (e.g. because you made mistakes in naming something, and/or assigned the wrong data to a particular variable and failed to correct it).
+ First, check the variables that you have in your current environment using `ls()`.
+ Next, to make sure everything is running the way you expect it to, remove all of the variables in your environment by running the command `rm(list = ls())` and check the result using `ls()`.

+ Finally, knit the file using the "Knit" button at the top of the editor window and review the html output that is created. 
  + Are you getting any errors? If so, you will see a message indicating where things went wrong in the "R Markdown" pane (another tab in the window with the Console that opens when you try to knit). 
  + If your file won't knit, go back and try to correct the mistakes you made above, then remove all your environment variables and try again.
  + If your document still won't knit, feel free to reach out to us for help, or ask one of your classmates if they can detect what's wrong with your code.

```{r}
# check the objects in the current environment
ls()

# remove all of them
rm(list=ls())

# check to make sure they are all gone
ls()

```

**Voilà! First homework exercise completed! Yay!!! You may now submit your final .Rmd file through Brightspace.**
