
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loops and functions

Load libraries
```{r}
library(tidyverse)
```

create a vector called 'chromosomes' that contains numbers from 1:20 
```{r}


```

Lets assume we want to add the prefix prefix "chr" before each number to signify that it is a chromosome. 
We will write a simple loop to do this (this can be easily done without a loop but we are using this example for practice).

```{r}


```

Instead of a vector, lets assume the numbers 1-20 are present in a dataframe column called "chromosomes". 
Create the dataframe and modify the loop so it works with the dataframe
```{r}


```

Write a function called 'add_prefix' that takes as input a vector and adds the prefix "chr" to each value of the vector 
```{r}


```

Modify the function 'add_prefix' so that it takes two inputs; a vector and a custom prefix text (instead of "chr") that it adds to each value of the input vector 
```{r}


```

Recreate the R function 'mean()'. The function takes a vector as input and outputs its mean. We can call it mean_xdas
```{r}


```

# dpqr Functions

We can use the rnorm() function to simulate a sample taken from a normally distributed population:
```{r}


```

The dnorm() function gives the height of the probability density function at a given point (this is more useful for discrete rather than continuous distributions)
```{r}


```

We can also use the dnorm() function to calculate the probability density function (pdf) of the whole population.
Lets make a pdf for a standard normal curve by applying the dnorm function to a range of zscores
```{r}


```

We can use the pnorm function to calculate the area under the curve below a specific zscore
```{r}


```

We can use the qnorm function find the zscore below which a certain percentage of the pdf area lies (qnorm is the inverse of pnorm)
```{r}


```


# More practice questions

For a standard normal distribution >99% of the values lie between -3 and +3 standard deviations (or z-scores). 
Create a dataframe with a column called 'zscores' containing values from -3 to +3 in intervals of 0.1 (use seq function)
```{r}


```

Use a for loop to  apply the dnorm function on each row of the dataframe and store the results in a new column called 'probability'
```{r}


```

Plot the zscore on the x-axis and probabilities on the y-axis using ggplot to recreate the standard normal cure (use geom_line)
```{r}


```

Use the same strategy to recreate a normal curve with a mean of 10 and sd of 3 
(hint: you will first need to find the zscores enclosing 99% of the values using qnorm)
```{r}


```

Inspect the plot to find height of the curve at its highest point. Now calculate this exact height using the dnorm function 
```{r}


```
