---
title: "Week 5 Recitation"
author: "Chris"
date: "9/25/2019"
output: html_document
---

```{r setup, include=FALSE}
# Do not change any options in this setup block

knitr::opts_chunk$set(echo = TRUE, error=TRUE)

# Import libraries
require(MASS)
require(ggplot2)
```

# Sampling

Lets load a data set. This is the expression of three genes from a group of cells in a single-cell experiment.

```{r}
example.expression.data <- read.delim("Week_4_Example.tsv")
```

Lets do some basic analysis of this data

```{r}
# The ecdf() function calculates an EMPIRICAL CDF
# It returns an ECDF object
# All you need to know about that object is that plot() knows what it is and will handle it fine
ylr044c.data <- example.expression.data$YLR044C
ylr044c.ecdf <- ecdf(ylr044c.data)

# Lets plot the histogram against the ECDF
par(mfrow=c(1,2))
hist(ylr044c.data, main="YLR044C Histogram", breaks = 100, xlab="Expression (Count)")
plot(ylr044c.ecdf, main = "YLR044C ECDF", xlab="Expression (Count)", ylab="Probability")
```

Yep, that's data. Just like it was last week. This week we're going to fit the negative binomial to this distribution.

```{r}
nbinom.fit <- MASS::fitdistr(example.expression.data$YLR044C, densfun = "negative binomial")
nbinom.fit
```

What is mu? It's just size & probability combined into a new parameter. I hate it. Lets recalculate `prob` from mu. We can put it in the same data structure that holds everything else

```{r}
calc.prob <- function(fitdistr.result) {fitdistr.result$estimate['size']/(fitdistr.result$estimate['size']+fitdistr.result$estimate['mu'])}
nbinom.fit$estimate['prob'] <- calc.prob(nbinom.fit)
```

How stable is this estimate for probability? Lets figure it out with SAMPLING.

```{r}

p.samples <- list()
num.resampling <- 10
sample.size <- 500

for (i in 1:num.resampling) {
  loop.sample <- example.expression.data$YLR044C[sample(nrow(example.expression.data), replace=T)]
  loop.fit <- MASS::fitdistr(loop.sample, densfun = "negative binomial")
  p.samples <- c(p.samples, calc.prob(loop.fit))
}

```

Neat. Lets plot this as a histogram.

```{r}

hist(unlist(p.samples))

```

Well this plot is just garbage. Lets just use `ggplot`

```{r}

p.samples <- data.frame(prob = unlist(p.samples))

ggplot(p.samples, aes(x=prob)) +
  geom_histogram(binwidth = 0.001) +
  theme_classic()

```

That is instantly less bad as a plot. Still though this looks pretty bad - lets try adding more samples

```{r}
p.samples <- list()
num.resampling <- 100
sample.size <- 500

for (i in 1:num.resampling) {
  loop.sample <- example.expression.data$YLR044C[sample(nrow(example.expression.data), replace=T)]
  loop.fit <- MASS::fitdistr(loop.sample, densfun = "negative binomial")
  p.samples <- c(p.samples, calc.prob(loop.fit))
}


```

And the plot:

```{r}

p.samples <- data.frame(prob = unlist(p.samples))

ggplot(p.samples, aes(x=prob)) +
  geom_histogram(binwidth = 0.001) +
  theme_classic()

```

Break into 3 groups and use the data provided to:

1. Load and look breifly at the structure of your data
2. Fit the negative binomial to each of 3 genes
3. Resample and plot the estimates of size and probability