---
title: "Data Manipulation using `dplyr`"
date: "September 17, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Our World in Data: the COVID19 pandemic

[Our World in Data](https://ourworldindata.org/) publishes data, visualizations, and analyses of issues of global importance, such as poverty, climate change, and disease. Today, we will a data set on COVID19 compiled by them. It contains profiles of 207 countries or territories. The data include the **daily** number of cases, deaths, vaccinations etc., and they are accompanied by some useful country stats, e.g. their population. We downloaded the data set (version 44) [from kaggle](https://www.kaggle.com/tunguz/data-on-covid19-coronavirus) but only kept a few selected columns.

## Look at the data

First, import and examine the data.
```{r}
covid <- read.csv("owid-covid-data_filtered.csv")
head(covid)
tail(covid)
str(covid)
```

## Set an ambitious goal

**Determine whether European countries that have the highest total number of COVID19 cases normalized to population size also have the highest number of COVID19-related deaths.**

## Split it into steps

First, describe each step in words.
```
1. 
2. 
...

```

Next, write `dplyr` code for each individual step.
```{r}
# step 1

# step 2

# ...

```

## Combine the steps into a pipeline

Join all the individual steps with `%>%`, feed the input data.frame into the pipeline, and assign the output tibble to a variable.
```{r}

```

Finally, plot the data on a bar graph. Make sure that the bars representing number of cases and number of deaths for each country are plotted side-by-side.
```{r}
# force the levels of "location" to maintain the same order as in the output tibble (otherwise, countries will be sorted alphabetically)

# make the plot

```

