---
title: "Data Manipulation using `dplyr`"
date: "September 17, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Our World in Data: the COVID19 pandemic

[Our World in Data](https://ourworldindata.org/) publishes data, visualizations, and analyses of issues of global importance, such as poverty, climate change, and disease. Today, we will a data set on COVID19 compiled by them. It contains profiles of 207 countries or territories. The data include the **daily** number of cases, deaths, vaccinations etc., and they are accompanied by some useful country stats, e.g. their population. We downloaded the data set (version 44) [from kaggle](https://www.kaggle.com/tunguz/data-on-covid19-coronavirus) but only kept a few selected columns.

## Look at the data

First, import and examine the data.
```{r}
covid <- read.csv("owid-covid-data_filtered.csv")
head(covid)
tail(covid)
str(covid)
```

## Set an ambitious goal

**Determine whether European countries that have the highest total number of COVID19 cases normalized to population size also have the highest number of COVID19-related deaths.**

## Split it into steps

First, describe each step in words.
```
1. Subset countries located in Europe.
2. Normalize the number of cases and the number of deaths to the population size.
3. For each country, sum up normalized new cases and normalized new deaths across all recorded days.
4. Sort the data by the normalized total number of cases and plot them.
```

Next, write `dplyr` code for each individual step.
```{r}
# subset countries located in Europe
covid_europe <- filter(covid,
                       continent=="Europe")

head(covid_europe)

# the same, but using piping
covid_europe <-
  covid %>%
  filter(continent=="Europe")

head(covid_europe)

# normalize cases and deaths to the population size
covid_europe_norm <-
  covid_europe %>% 
  mutate(new_cases_per_thousand=1000*new_cases/population,
         new_deaths_per_100k=100000*new_deaths/population)

head(covid_europe_norm)
covid_europe_norm[100:105,]

# group by country
covid_europe_norm_grouped <-
  covid_europe_norm %>%
  group_by(location)

head(covid_europe_norm_grouped)

# sum cases and deaths across days
covid_europe_norm_grouped_totals <-
  covid_europe_norm_grouped %>%
  summarize(total_cases_per_thousand=sum(new_cases_per_thousand,na.rm=TRUE),
            total_deaths_per_100k=sum(new_deaths_per_100k,na.rm=TRUE))

head(covid_europe_norm_grouped_totals)

# sort in descending order
covid_europe_norm_grouped_totals_sorted <-
  covid_europe_norm_grouped_totals %>%
  arrange(desc(total_cases_per_thousand))

head(covid_europe_norm_grouped_totals_sorted)

# transform into the long format
covid_europe_norm_grouped_totals_sorted_long <-
  covid_europe_norm_grouped_totals_sorted %>%
  gather(key="metric",
         value="value",
         "total_cases_per_thousand",
         "total_deaths_per_100k")

head(covid_europe_norm_grouped_totals_sorted_long)
tail(covid_europe_norm_grouped_totals_sorted_long)
```

## Combine the steps into a pipeline

Join all the individual steps with `%>%`, feed the input data.frame into the pipeline, and assign the output tibble to a variable.
```{r}
covid_europe_summary <- 
  covid %>%
  filter(continent=="Europe") %>%
  mutate(new_cases_per_thousand=1000*new_cases/population,
         new_deaths_per_100k=100000*new_deaths/population) %>%
  group_by(location) %>%
  summarize(total_cases_per_thousand=sum(new_cases_per_thousand,na.rm=TRUE),
            total_deaths_per_100k=sum(new_deaths_per_100k,na.rm=TRUE)) %>%
  arrange(desc(total_cases_per_thousand)) %>%
  gather(key="metric",
         value="value",
         "total_cases_per_thousand",
         "total_deaths_per_100k")

covid_europe_summary
```

Finally, plot the data on a bar graph. Make sure that the bars representing number of cases and number of deaths for each country are plotted side-by-side.
```{r}
# force the levels of "location" to maintain the same order as in covid_europe_summary (otherwise, countries will be sorted alphabetically)
covid_europe_summary$location <- factor(covid_europe_summary$location,
                                        levels=unique(covid_europe_summary$location)) # we use unique() here because each country name is repeated twice in the tibble, and level names must be unique

# make the plot
ggplot(data = covid_europe_summary,
       mapping = aes(x = location,
                     y = value,
                     fill = metric)) +
  geom_col(position="dodge") +
  coord_flip()
```

