---
title: "Bayes"
author: "Chris"
date: "11/24/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(ggplot2)
ggplot2::theme_set(theme_classic())
require(bayestestR)
require(rstanarm)
```

## Measurements

```{r}
measurements <- c(0, 0, 1, 1, 2)
print(measurements)
```

## Likelihood Function

```{r}
# mu ~ Uniform(-1,2)
mu <- seq(-1, 2, by = 0.02)

# Calculate likelihoods for each mu value
likelihood.function.of.mu <- function(x) {prod(dnorm(measurements, mean=x, sd=1))}

likelihoods <- vapply(mu, likelihood.function.of.mu, double(1))


# Normalize to area under the curve = 1 with the trapezoid rule
likelihoods.density <- likelihoods / sum(mu * c(likelihoods[1], abs(likelihoods[2:151] - likelihoods[1:150])/2))

qplot(mu, likelihoods.density, geom="line")
```

## Maximum likelihood

```{r}
print(mu[which.max(likelihoods)])
print(mean(measurements))
```

## Bayes Factor

Let's load some data and do bayesian regression

```{r}
gdh2 <- read.delim("Bayes_Data.tsv")
qplot(time, GDH2_L2FC, data=gdh2, geom="point", color=Experiment)
```

Define a prior for time

```{r}
# X is the predictor data (time)
# y is the response data (log2FC expression)
# beta is the model parameter from regression
# sigma is the standard deviation of the response data

gdh2.prior <- rstanarm::normal(location = c(0.1), scale=c(0.1), autoscale = F)

```

Set prior knowledge and regress against the data (x)

```{r}
capture.output(rstanarm_results <- stan_glm(GDH2_L2FC ~ time, data=gdh2[gdh2$Experiment == "Treated",], prior = gdh2.prior), file="NULL")
prior_summary(rstanarm_results)

summary(rstanarm_results, probs=c(.025, .975), digits=3)

plot(rstanarm_results)
```

What is this saying?

```{r}
posterior.draws <- as.data.frame(as.matrix(rstanarm_results))
colnames(posterior.draws) <- c("Intercept", "time", "sigma")

qplot(time, GDH2_L2FC, data=gdh2, geom="point", color=Experiment) +
  geom_abline(data = posterior.draws, aes(intercept = Intercept, slope = time), color = "skyblue", size = 0.2, alpha = 0.1) + 
  geom_abline(intercept = coef(rstanarm_results)[1], slope = coef(rstanarm_results)[2], color = "skyblue4", size = 1)
```

Let's load some new data.

```{r}
gdh2.more.data <- read.delim("Bayes_Update.tsv")
qplot(time, GDH2_L2FC, data=gdh2.more.data, geom="point", color=Experiment)
```

How does it look if we just model this new data?

```{r}
capture.output(rstanarm_results_new <- stan_glm(GDH2_L2FC ~ time, data=gdh2.more.data[gdh2.more.data$Experiment == "Treated",], 
                                                prior = gdh2.prior), 
               file="NULL")
prior_summary(rstanarm_results)

summary(rstanarm_results_new, probs=c(.025, .975), digits=3)

plot(rstanarm_results_new)

posterior.draws <- as.data.frame(as.matrix(rstanarm_results_new))
colnames(posterior.draws) <- c("Intercept", "time", "sigma")

qplot(time, GDH2_L2FC, data=gdh2, geom="point", color=Experiment) +
  geom_abline(data = posterior.draws, aes(intercept = Intercept, slope = time), color = "skyblue", size = 0.2, alpha = 0.1) + 
  geom_abline(intercept = coef(rstanarm_results_new)[1], slope = coef(rstanarm_results_new)[2], color = "skyblue4", size = 1)
```

Now let's use the posterior prediction from the first set of data as the prior for this model

```{r}
updated.prior <- normal(location = coef(rstanarm_results)[2], scale = rstanarm_results$ses[[2]])
updated.intercept <- normal(location = coef(rstanarm_results)[1], scale = rstanarm_results$ses[[1]])

capture.output(rstanarm_results_updated <- stan_glm(GDH2_L2FC ~ time, 
                                                    data=gdh2.more.data[gdh2.more.data$Experiment == "Treated",], 
                                                    prior = updated.prior,
                                                    prior_intercept = updated.intercept), file="NULL")

summary(rstanarm_results_updated, probs=c(.025, .975), digits=3)

posterior.draws <- as.data.frame(as.matrix(rstanarm_results_updated))
colnames(posterior.draws) <- c("Intercept", "time", "sigma")

qplot(time, GDH2_L2FC, data=gdh2, geom="point", color=Experiment) +
  geom_abline(data = posterior.draws, aes(intercept = Intercept, slope = time), color = "skyblue", size = 0.2, alpha = 0.1) + 
  geom_abline(intercept = coef(rstanarm_results_updated)[1], slope = coef(rstanarm_results_updated)[2], color = "skyblue4", size = 1)

```
