
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loops and functions

Load libraries
```{r}
library(tidyverse)
```

create a vector called 'chromosomes' that contains numbers from 1:20 
```{r}
chromosomes=seq(1:20)
```

Lets assume we want to add the prefix prefix "chr" before each number to signify that it is a chromosome. 
We will write a simple loop to do this (this can be easily done without a loop but we are using this example for practice).

```{r}
temp=as.numeric()
for (i in seq(1:length(chromosomes))){
  temp[i]=paste("chr",chromosomes[i],sep = "")
    }
temp
```

Instead of a vector, lets assume the numbers 1-20 are present in a dataframe column called "chromosomes". 
Create the dataframe and modify the loop so it works with the dataframe
```{r}
d=data.frame(chromosomes=seq(1:20))

temp=as.numeric()
for (i in seq(1:length(d$chromosomes))){
  temp[i]=paste("chr",d[i,"chromosomes"],sep = "")
}
temp
```

Write a function called 'add_prefix' that takes as input a vector and adds the prefix "chr" to each value of the vector 
```{r}
add_prefix=function(inputdata){
  temp=as.numeric()
  for (i in seq(1:length(inputdata))){
    temp[i]=paste("chr",inputdata[i],sep = "")
  }
temp
}

add_prefix(inputdata = chromosomes)

```

Modify the function 'add_prefix' so that it takes two inputs; a vector and a custom prefix text (instead of "chr") that it adds to each value of the input vector 
```{r}
add_prefix=function(inputdata,customtext){
  temp=as.numeric()
  for (i in seq(1:length(inputdata))){
    temp[i]=paste(customtext,inputdata[i],sep = "")
  }
  temp
}

add_prefix(inputdata = chromosomes,customtext = "abcd")
```

Recreate the R function 'mean()'. The function takes a vector as input and outputs its mean. We can call it mean_xdas
```{r}
mean_xdas=function(inputdata){
  sum(inputdata)/length(inputdata)
}

mean(chromosomes)
mean_xdas(chromosomes)
```

# dpqr Functions

We can use the rnorm() function to simulate a sample taken from a normally distributed population:
```{r}
d=rnorm(100)
hist(d)

d=rnorm(100,10,2)
hist(d)
```

The dnorm() function gives the height of the probability density function at a given point (this is more useful for discrete rather than continuous distributions)
```{r}
dnorm(2,mean = 0,sd=1)
```

We can also use the dnorm() function to calculate the probability density function (pdf) of the whole population.
Lets make a pdf for a standard normal curve
```{r}
y=dnorm(seq(-3,3,0.1),mean=0,sd=1)
x=seq(-3,3,0.1)
pdf=data.frame(zscores=x,probability=y)
pdf_plot=ggplot(pdf,aes(x=zscores,y=probability))+geom_line()
pdf_plot
```

We can use the pnorm function to calculate the area under the curve below a specific zscore
```{r}
pnorm(2,mean = 0,sd=1)
```

We can use the qnorm function find the zscore below which a certain percentage of the pdf area lies (qnorm is the inverse of pnorm)
```{r}
qnorm(0.02275,mean = 0,sd=1)
```


# More practice questions

For a standard normal distribution >99% of the values lie between -3 and +3 standard deviations (or z-scores). 
Create a dataframe with a column called 'zscores' containing values from -3 to +3 in intervals of 0.1 (use seq function)
```{r}
d=data.frame(zscores=seq(-3,3,0.1))
```

Use a for loop to  apply the dnorm function on each row of the dataframe and store the results in a new column called 'probability'
```{r}
for (i in seq(1:length(d$zscores))){
  d[i,"probability"]=dnorm(d[i,1])
  } 
```

Plot the zscore on the x-axis and probabilities on the y-axis using ggplot to recreate the standard normal cure (use geom_line)
```{r}
ggplot(d,aes(x=zscores,y=probability))+geom_line()
```

Use the same strategy to recreate a normal curve with a mean of 10 and sd of 3 
(hint: you will first need to find the zscores enclosing 99% of the values using qnorm)
```{r}
qnorm(0.995,10,3)
qnorm(0.005,10,3)
d=data.frame(zscores=seq(2,18))

for (i in seq(1:length(d$zscores))){
  d[i,"probability"]=dnorm(d[i,1],10,3)
} 

ggplot(d,aes(x=zscores,y=probability))+geom_line()
```

Inspect the plot to find height of the curve at its highest point. Now calculate this exact height using the dnorm function 
```{r}
dnorm(10,10,3)
```
