```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This dataset contains the details of the covid cases reported during the first two months of the pandemic.


Import the data and remove any rows containing 'NA'
```{r}
covid=read.csv("covid.csv")
covid=na.omit(covid)


```

We suspect that most people who dont survive a covid infection are over the age of 60. Test the hypothesis that the mean age of covid patients who die is greater than 60.
```{r}
#What type of t.test is appropriate to test this hypothesis?
#one-sample, one-tailed


#Write the null and alternative hypothesis:
#Null: u<=60
#Alternative: u>60


#Calculate the mean age of the people who dont survive:
avg_death=mean(covid[covid$Survived=="No","Age"])


#Calculate the sem for people who dont survive:
sem=sd(covid[covid$Survived=="No","Age"])/sqrt(length(covid[covid$Survived=="No","Age"]))


#Calculate the t-statistic for the hypothesized mean age of 60:
t=(avg_death-60)/sem


#Calculate the p.value  and 95% confidence interval
pt(t,df=57,lower.tail = F)

t_crit=qt(p = 0.05,df=57,lower.tail = F)
avg_death-(t_crit*sem)


#Does the confidence interval contain the target value of 60 years and what does that tell us?
#No.It tells us that the result is significant


#Perform the test with the t.test function to confirm previous answers
t.test(covid[covid$Survived=="No","Age"],mu=60,alternative = "greater")


#Use bootstrapping with 10000 trials to find the previously calculated p.value and 95% C.I 
bs=as.numeric()
for (i in 1:10000){
  bs[i]=mean(sample(covid[covid$Survived=="No","Age"],58,replace = T))
  }

hist(bs)
sum(bs<60)/10000
(sort(bs))[500]




```

Test the hypothesis that the mean age of patients who survive the infection is different from those who do not survive
```{r}

#Calculate the mean, variance and no of observations in the surviving and non-surviving groups
mean_dead=mean(covid[covid$Survived=="No","Age"])
mean_alive=mean(covid[covid$Survived=="Yes","Age"])

n_dead=length(covid[covid$Survived=="No","Age"])
n_alive=length(covid[covid$Survived=="Yes","Age"])

var_dead=var(covid[covid$Survived=="No","Age"])
var_alive=var(covid[covid$Survived=="Yes","Age"])

var.test(covid[covid$Survived=="No","Age"],covid[covid$Survived=="Yes","Age"])


#Which type of t.test should be used to test this hypothesis?
#Two sample test with Welch df as variance is not equal


#Calculate the standard error 
se_welch=sqrt((var_dead/n_dead)+(var_alive/n_alive))


#Calculate the t.statistic 
t=(mean_dead-mean_alive)/se_welch


#Calculate the p.value (Use 72.076 as degrees of freedom)
(1-pt(q=10.674,df=72.076))*2

pt(q=t,df=72.076,lower.tail = F)*2 #doesn't match with t.test results possibly due to rounding error

#Calculate the 95% confidence interval (Use 72.076 as degrees of freedom)
t_crit=qt(0.025,df=72.076)
(mean_dead-mean_alive)+(t_crit*se_welch)
(mean_dead-mean_alive)-(t_crit*se_welch)

#Confirm your answers using the t.test function
t.test(x=covid[covid$Survived=="No","Age"],y=covid[covid$Survived=="Yes","Age"],mu=0,alternative = "t",paired = F,var.equal = F)


```

Does the *gender* of a patient affect their chances of *survival*? Use any appropriate test to test this hypothesis.
```{r}
tab=table(covid$Survived,covid$Gender)
chisq.test(tab)
fisher.test(tab)



```

Is one gender more likely to be *infected* by covid than the other? Use any appropriate test to test this hypothesis.
```{r}
binom.test(sum(covid$Gender=="male"),length(covid$Gender),p=0.5,alternative = "t")

```

