---
title: 'In-class exercise: Conditional Bayes'
subtitle: "XDASI Fall 2021"
date: "9/23/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


The following exercise is from Whitlock & Schluter Q36, 37.

Some people are hypersensitive to the smell of asparagus, and can even detect a strong odor in the urine of a person who has recently eaten asparagus. This trait turns out to have a simple genetic basis:

+ An individual with one or two copies of the $A$ allele at the gene ($AA$ or $Aa$ genotypes) can smell asparagus in urine, whereas a person with two copies of the alternative $a$ allele ($aa$ genotypes) cannot (Online Mendelian Inheritance in Man 2019). 

Assume that men and women in the population have the same allele frequencies at the asparagus-smelling gene and that marriage and child production are independent of the genotype at the gene. In the human population, 5% of alleles are $A$ and 95% are $a$:

+ $Freq(A) = 0.05$
+ $Freq(a) = 0.95$

**A) What is the probability that a randomly sampled individual from the population has two copies of the $a$ allele (that is, that it has an $aa$ genotype), assuming random mating?**

```{r}
Pr = list()  # set up a data structure to hold allele data

# allele frequencies in the population
Pr$a = 0.95
Pr$A = 0.05

# expected frequency of aa homozyotes in the population
# Pr[aa] = Pr[a] * Pr[a] = (0.95)(0.95) = 0.9025.
Pr$aa = Pr$a * Pr$a
Pr$aa
```


**B) What is the probability that both members of a randomly sampled married couple (man and woman) are $aa$ at the asparagus-smelling gene?** 

```{r}
# joint probability that both parents are homozygous for aa
# Pr[both aa] = Pr[aa] Pr[aa] = (0.9025)(0.9025) = 0.8145.
Pr$aa_aa = Pr$aa * Pr$aa
Pr$aa_aa
```


**C) What is the probability that both members of a randomly sampled married couple (man and woman) are $Aa$ heterozygotes at this locus?**

```{r}
# probability of being a heterozygote: M(A)F(a) + M(a)F(A)

# Pr[Aa] = Pr[A] Pr[a] + Pr[a] Pr[A] = 2(0.95)(0.05) = 0.095. 
Pr$Aa = ( Pr$A * Pr$a ) + ( Pr$a * Pr$A )
Pr$Aa

# probability that both parents are heterozygous

# Pr[both Aa] = Pr[Aa] Pr[Aa] = (0.095)(0.095) = 0.009025.
Pr$Aa_Aa = Pr$Aa * Pr$Aa
Pr$Aa_Aa        
```


**D) Consider the type of couple described in (C). What is the probability that the first child of such a couple is also an $Aa$ heterozygote? Remember that the child must receive exactly one allele from each parent.**

```{r}
# if both parents are Aa, than we can draw a table
knitr::kable(data.frame(A=c("AA","aA"),
                        a=c("Aa","aa"),
                        row.names = c("A","a")))
```

```{r}
# 4 possible combinations, of which 2 are “Aa” offspring: A is from mother and a is from father, or vice-versa. 
# Pr[child is Aa | both parents Aa] = 2(0.5)(0.5) = 0.5.
Pr$C.Aa = 0.5^2 + 0.5^2
Pr$C.Aa
```

So, $P(C.Aa | M.Aa,F.Aa) = 0.5$


***E) If a randomly sampled child has the $aa$ genotype, what is the probability that both its parents were also $aa$?***

<!-- ======================================================================= -->
<!-- <div class="qbox"> -->
<!-- $\Rightarrow$ ***Question: ?*** -->

<!-- <details closed markdown="block"> -->
<!--   <summary>Answer</summary> -->

<!-- + Answer. -->

<!-- </details> -->
<!-- </div> -->
<!-- ======================================================================= -->

Let's use Bayes theorem to frame the problem:

$$P(M.aa,F.aa|C.aa)=\frac{P(C.aa|M.aa,F.aa)*P(M.aa,F.aa)}{P(C.aa)}$$

Using the Law of Total Probability:

$$
\begin{aligned}
P(C.aa) =& \ P(C.aa | M.aa,F.aa) \ \ * P(M.aa,F.aa) \ \ + \\
         & \ P(C.aa | M.Aa,F.aa) \ * P(M.Aa,F.aa) \ + \\
         & \ P(C.aa | M.aa,F.Aa) \ * P(M.aa,F.Aa) \ + \\
         & \ P(C.aa | M.Aa,F.Aa) * P(M.Aa,F.Aa) \\
        =& \ (1)(0.9025)(0.9025) + (0.5)(0.095)(0.9025) + (0.5)(0.9025)(0.095) + (0.25)(0.095)(0.095) \\
        =& \ 0.9025.
\end{aligned}
$$ 

<!-- ======================================================================= -->
<!-- <div class="qbox"> -->
<!-- $\Rightarrow$ ***Question: ?*** -->

<!-- <details closed markdown="block"> -->
<!--   <summary>Answer</summary> -->

<!-- + Answer. -->

<!-- </details> -->
<!-- </div> -->
<!-- ======================================================================= -->

Since all other parental combinations will have zero probability of obtaining offspring with the $aa$ genotype.

Let's first recall some basics:

```{r}
# allele frequencies
Pr$a = 0.95
Pr$A = 0.05

# single parents
Pr$aa = Pr$a * Pr$a
Pr$Aa = ( Pr$A * Pr$a ) + ( Pr$a * Pr$A )
Pr$aa
Pr$Aa

# both parents aa
Pr$aa_aa = Pr$aa * Pr$aa
Pr$aa_aa

# both parents aA
Pr$Aa_Aa = Pr$Aa * Pr$Aa
Pr$Aa_Aa 

# one parent aa, other Aa
Pr$Aa_aa = Pr$Aa * Pr$aa
Pr$Aa_aa
```

Now we are ready to compute the total probability:

```{r}
# probabilities of child genotype based on Punnett square
Pr$Caa_MaaFaa = 1
Pr$Caa_MaaFAa = 0.5
Pr$Caa_MAaFaa = 0.5
Pr$Caa_MAaFAa = 0.25

Pr$Caa = Pr$Caa_MaaFaa * Pr$aa * Pr$aa +
         Pr$Caa_MaaFAa * Pr$aa * Pr$Aa + 
         Pr$Caa_MAaFaa * Pr$Aa * Pr$aa +
         Pr$Caa_MAaFAa * Pr$Aa * Pr$Aa
Pr$Caa
```

Finally:

<!-- ======================================================================= -->
<!-- <div class="qbox"> -->
<!-- $\Rightarrow$ ***Question: ?*** -->

<!-- <details closed markdown="block"> -->
<!--   <summary>Answer</summary> -->

<!-- + Answer. -->

<!-- </details> -->
<!-- </div> -->
<!-- ======================================================================= -->

$$P(M.aa,F.aa | C.aa) = \frac{P(C.aa | M.aa,F.aa)*P(M.aa,F.aa)}{P(C.aa)} = \frac{1 * (0.9025)^2}{0.9025} = 0.9025$$

```{r}
Pr$MaaFaa_Caa = ( Pr$Caa_MaaFaa * Pr$aa * Pr$aa ) / Pr$Caa
Pr$MaaFaa_Caa
```
