---
title: "Homework1 Worksheet"
author: "Manpreet S. Katari"
date: "September 5, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Q1: Create a matrix (call it transcriptome) with the values below. The experiments are column names and genes are the row names.
_Note: use a matrix object for now, don't worry about trying to create a data frame with factors. We will go over this again next week._

GeneName | Control | Nitrogen | Phosphate | Potassium
---------|---------|----------|-----------|-----------
GeneA    |   89    |    78    |     77    |    56
GeneB    |   90    |    99    |     85    |    97
GeneC    |   78    |    94    |     99    |    87
GeneD    |   81    |    83    |     80    |    79

```{r}

# Enter you answer here.
# Don't forget to add comments to explain your code.


```


### Q2: Use an R function to calculate the average expression for all genes and save it in a vector called ``expression_average``.
_Hint: Remember that you can always get help with commands by typing `?commandname`, `apropos("commandname")`, and `example("commandname")`._

```{r}

# Enter you answer here.
# Don't forget to add comments to explain your code.


```

### Q3: Sort the matrix such that the gene with the highest average gene expression is on top.
_Hint: the **sort()** function sorts the data and the **order()** function provides how to order the data to be sorted.

```{r}

# Enter you answer here.
# Don't forget to add comments to explain your code.


```
