---
title: "Anova"
author: "Raza"
date: "11/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
This exercise is based on the mousegenes dataset which contained the gc_content of all mouse genes. For this exercise I have retained just 3 chromosomes and simulated the values so the data is not real.

The data shows gc-content of randomly selected regions from chr 10,11 and 12. We will use anova to check if there is a significant difference in the gc-content of different chromosomes. 

1. Import the data
```{r}
genes=read.csv("gc.csv")

```

2. Make a boxplot of gc content by chromosome to check if the gc-content looks different among chromosomes 
```{r}
boxplot(data=genes,GC_Percent~Chr)

#chr 12 looks different
```

3. Is there a significant difference between the gc_content of chr11 and chr12 as compared to chr10 (do planned anova)
```{r}
summary(lm(data=genes,GC_Percent~Chr))

#chr 11 and 12 have significant difference, chr10 and 12 dont
```

4. Repeat this analysis using chr12 as the reference 
```{r}
fac=factor(genes$Chr,levels = c("chr12","chr10","chr11"))
summary(lm(data=genes,GC_Percent~fac))

```

5. Compare the GC_content among all 3 chromosomes using unplanned anova
```{r}
summary(aov(data=genes,GC_Percent~Chr))

```

6. Use the TukeyHSD test to find which chromosomes have different GC_content and compare the results with those obtained with the planned anova
```{r}
TukeyHSD(aov(data=genes,GC_Percent~Chr))

#chr12 is different from others. ch10 and chr 11 are not different. Results match with previous answer
```

7. Manually compute the SSE and SSB (should match the results from the aov function)
```{r}
# make 3 vectors containing gc percentage for chr 10, 11, and 12
chr10=(genes[genes$Chr=="chr10","GC_Percent"])
chr11=(genes[genes$Chr=="chr11","GC_Percent"])
chr12=(genes[genes$Chr=="chr12","GC_Percent"])

#calculate the mean of each chromosome and the grand mean
gr.mean=mean(c(chr10,chr11,chr12))
chr10.mean=mean(chr10)
chr11.mean=mean(chr11)
chr12.mean=mean(chr12)

#use the means and the vectors to find ssb and sse
ssb= sum(length(chr10)*(chr10.mean-gr.mean)^2)+sum(length(chr11)*(chr11.mean-gr.mean)^2)+sum(length(chr12)*(chr12.mean-gr.mean)^2)
sse=sum((chr10-chr10.mean)^2)  + sum((chr11-chr11.mean)^2)  + sum((chr12-chr12.mean)^2) 
```

8. Calculate dof for ssb and sse
```{r}
df_ssb=3-1
df_sse=(length(chr10)-1)+(length(chr11)-1)+(length(chr12)-1)
```

9. Calculate the mean square values for ssb and sse
```{r}
ssb_mean=ssb/df_ssb
sse_mean=sse/df_sse
```

10. Calculate the f statistic
```{r}
f=ssb_mean/sse_mean
```

11. Calculate the p.value  
```{r}
pf(f,df1 = df_ssb,df2 = df_sse,lower.tail = F)

```

12. Conserved coding regions may have higher gc-content than noncoding regions. Add an interaction term to the aov function to check if being a coding/noncoding region contributes to the difference in GC_content among chromosomes 
```{r}
summary(aov(data=genes,GC_Percent~Chr*Coding))

#interaction is significant
```

13. Does this interaction exist on all chromosomes or on a specific chromosome? Check with anova and with a boxplot
```{r}
TukeyHSD(aov(data=genes,GC_Percent~Chr*Coding))
boxplot(data=genes,GC_Percent~ Chr+Coding)

#interaction is only significant in pairwise comparisons involving chr12. boxplot also confirms interaction exists only on chr12

```



