---
title: "Anova"
author: "Raza"
date: "11/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
This exercise is based on the mousegenes dataset which contained the gc_content of all mouse genes. For this exercise I have retained just 3 chromosomes and simulated the values so the data is not real.

The data shows gc-content of randomly selected regions from chr 10,11 and 12. We will use anova to check if there is a significant difference in the gc-content of different chromosomes. 

1. Import the data
```{r}


```

2. Make a boxplot of gc content by chromosome to check if the gc-content looks different among chromosomes 
```{r}


```

3. Is there a significant difference between the gc_content of chr11 and chr12 as compared to chr10 (do planned anova)
```{r}

```

4. Repeat this analysis using chr12 as the reference 
```{r}

```

5. Compare the GC_content among all 3 chromosomes using unplanned anova
```{r}


```

6. Use the TukeyHSD test to find which chromosomes have different GC_content and compare the results with those obtained with the planned anova
```{r}

```

7. Manually compute the SSE and SSB (should match the results from the aov function)
```{r}
# make 3 vectors containing gc percentage for chr 10, 11, and 12
chr10=
chr11=
chr12=

#calculate the mean of each chromosome and the grand mean
gr.mean=
chr10.mean=
chr11.mean=
chr12.mean=

#use the means and the vectors to find ssb and sse
ssb= 
sse=
  
```

8. Calculate dof for ssb and sse
```{r}

df_ssb=    #no of groups -1
df_sse=    #sum of (n-1) of all groups  
  
```

9. Calculate the mean square values for ssb and sse
```{r}
ssb_mean=
sse_mean=
```

10. Calculate the f statistic
```{r}
f=
  
```

11. Calculate the p.value  
```{r}


```

12. Conserved coding regions may have higher gc-content than noncoding regions. Add an interaction term to the aov function to check if being a coding/noncoding region contributes to the difference in GC_content among chromosomes 
```{r}

```

13. Does this interaction exist on all chromosomes or on a specific chromosome? Check with anova and with a boxplot
```{r}


```



