---
title: "Distributions in-class exercise"
subtitle: "XDASI Fall 2021"
author: "YOUR NAMES HERE"
date: "11/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

## Scenario

You want to clone a gene and you have some reporter for success of transformation in *E. coli*. Positive colonies are blue because they express $\beta$-galactosidase when plated on medium with IPTG. The probability of a successful transformation is $0 \le P(success) \le 1$. For the following exercise, you may consider one plate, many plates, or a truly gargantuan number of plates.

## Exercise

The class has been split into groups and each has taken one or two of the following distributions:

- Bernoulli
- Binomial
- Hypergeometric
- Geometric
- Negative binomial
- Poisson
- Normal
- Exponential

For your assigned distribution, answer the following questions.

#### a. What is your function? Describe it in terms of the kind of problem that it applies to. Is it discrete or continuous?

<!-- Insert your answer below. -->
```{r}
"geometric:pdf=(1-p)^x*p . In a series of bernulli trails, the probability of have x number of failures before 1 success. Discrete.

negative binomial:combine(r+x-1,r)*(1-p)^x*p^r. In a series of bernulli trails, the probability of have x number of failures before r successes. Discrete"

```

#### b. Describe a question that you can ask with your distribution as applied to the transformation scenario given above.

<!-- Insert your answer below. -->
```{r}
"geometric:Find the probability of some number of non-transformed colonies per one transformed colony

negative binomial:How many trials do we need before x number of transformed colonies per plate. "

```


#### c. Describe the parameterization of the equation. What parameters does it take, and what do they mean?

<!-- Insert your answer below. -->
```{r}
"geometric:p, probability of success

negative binomial:p, probability of success;r, number of successes "

```


#### d. Pick some appropriate parameters for your distribution as applied to the transformation scenario and graph the PDF and CDF.

```{r}
xrange=1:30
pgeom=data.frame(x=rep(xrange,3),y=c(dgeom(xrange, 0.2),dgeom(xrange, 0.4),dgeom(xrange, 0.6)),
                p=c(rev(rep("p=0.2",30)),rev(rep("p=0.4",30)),rev(rep("p=0.6",30))))
ggplot(pgeom,
       aes(x,y,col=p))+
  geom_line()+
  labs(title = "geom_pdf")
cgeom=data.frame(x=rep(xrange,3),y=c(pgeom(xrange, 0.2),pgeom(xrange, 0.4),pgeom(xrange, 0.6)),
                p=c(rev(rep("p=0.2",30)),rev(rep("p=0.4",30)),rev(rep("p=0.6",30))))
ggplot(cgeom,
       aes(x,y,col=p))+
  geom_line()+
  labs(title = "geom_cdf")

pnb=data.frame(x=rep(xrange,3),y=c(dnbinom(xrange, 3,0.4),dnbinom(xrange,6, 0.4),dnbinom(xrange,9, 0.4)),
                p=c(rev(rep("r=3",30)),rev(rep("r=6",30)),rev(rep("r=9",30))))
ggplot(pnb,
       aes(x,y,col=p))+
  geom_line()+
  labs(title = "nbinom_pdf,p=0.4")
cnb=data.frame(x=rep(xrange,3),y=c(pnbinom(xrange, 3,0.4),pnbinom(xrange,6, 0.4),pnbinom(xrange,9, 0.4)),
                p=c(rev(rep("r=3",30)),rev(rep("r=6",30)),rev(rep("r=9",30))))
ggplot(cnb,
       aes(x,y,col=p))+
  geom_line()+
  labs(title = "nbinom_cdf,p=0.4")
```


#### e. Compute a cumulative probability for some appropriate value and describe in words what it means.

```{r}
pgeom(5,0.5)
pnbinom(5,5,0.5)
"The probability of getting a success with no more than 5 number of failures when probability for success is 0.5
The probability of getting 5 successes with no more than 5 failures before when probability for success is 0.5"
```

<!-- Insert your answer below. -->


#### f. Identify at least one other distribution that yours relates to and be prepared to discuss the relationship between them. This could be a special case, a limiting case, an approximation, or an inverse CDF (survival) relationship.

<!-- Insert your answer below. -->
```{r}
"geometric distribution is a limiting case of negative binomial when we set the size of negative binomial to 1"
```

