---
title: 'Homework 10: Hypothesis Testing'
author: "Shaili Regmi"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
  pdf_document: default
subtitle: XDASI Fall 2021
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
library(mosaic)
```

<!-- ======================================================================= -->
## COVID-19 cases in NYC

"NYC_covid_cases_by_age.csv" contains weekly data of cases in NYC over the course of the pandemic. 

### 1) Import, tidy, and plot the data

#### a) Convert to long format

Read in the data file and convert the data to long format (you will need this in order to make the plot below).

  + You should preserve the `week_ending` column.
  + Get rid of the `age_all_ages` column.
  + Convert the rest of the columns to two new columns containing the `age_group` and `cases` information.
  + Turn the `age_groups` column into a factor (watch out, R likes to put everything in alphabetical order! you can set the levels in the correct order of age group by making a vector of the column names from the original data table and then selecting just columns 3:12)
  + Make sure the `week_ending` column is properly formatted as a Date type.
  
```{r}
# import file
covid = read.csv("NYC_covid_cases_by_age.csv")

# check structure
str(covid)

# make a data frame in long format
covid_long = covid %>%
  gather(key = "age_group",
         value = "cases",
         colnames(covid[3:12]))

# get rid of what used to be the age_all_ages column
covid_long = subset(covid_long,
                    select = -age_all_ages)

# format age_group as factor in the correct order of age
covid_long$age_group = factor(covid_long$age_group, levels = colnames(covid[3:12]))

# format week_ending as a Date type (hint: look up `as.Date()`)
covid_long$week_ending = as.Date(covid_long$week_ending, "%m/%d/%y")
str(covid_long)

```

#### b) Percentage of cases in each age group

As a convenience for questions to come, also add a column containing the percentage of cases observed in each age group out of the total cases per week

Hint: dplyr will be really useful for this. It's a good idea to drop rows with `NA` at this point (look up `drop_na`).

```{r}
# add pct_cases column
covid_long = covid_long %>% 
  drop_na() %>% 
  group_by(week_ending) %>% 
  mutate(pct_cases = cases / sum(cases) * 100)

# check weekly totals (should = 100)
totals = c()
for (i in covid_long$week_ending[1:84]){
  totals = c(totals, sum(covid_long$pct_cases[covid_long$week_ending == i]))
}
totals

```

#### c) Plot cases over time by age group

Create a plot that shows the trajectory of cases over the course of the pandemic broken down by age group.

```{r}
# ggplot line plot

ggplot(data = covid_long,
       mapping = aes(x = week_ending,
                     y = cases,
                     col = age_group)) +
  geom_line() +
  scale_x_continuous(breaks = covid_long$week_ending[c(1, 10, 20, 30, 40, 50, 60, 70, 80)]) +
  theme(axis.text.x = element_text(angle = 45))

```


<!-- ======================================================================= -->
### 2) First and second waves

The plot above shows that the number of cases over time forms "waves". Did the first and the second wave have a similar number of weekly recorded cases in all age groups?

#### a) Total cases in each wave

Subset the data that corresponds to the first wave. You have to decide for yourself where to draw the cutoffs on the time axis. Make a new data frame containing the total weekly number of cases in the first wave (sum them up across all age groups).

Then, do the same for the second wave.

```{r}
# ============================================================================ #
# First wave data
# ============================================================================ #
# filtered data
first_wave = covid_long %>% 
  filter(week_ending >= "2020-03-21" & week_ending <= "2020-05-23") %>% 
  group_by(week_ending) %>% 
  mutate(weekly_total = sum(cases))

# summary table (total cases per week in 1st wave)
first_wave_summary = data.frame(week_ending = first_wave$week_ending[1:10],
                                weekly_total = first_wave$weekly_total[1:10])
first_wave_summary

# ============================================================================ #
# Second wave data
# ============================================================================ #
# filtered data
second_wave = covid_long %>% 
  filter(week_ending >= "2020-12-19" & week_ending <= "2021-02-20") %>% 
  group_by(week_ending) %>% 
  mutate(weekly_total = sum(cases)) ## Although these dates only account for the start of the second wave and not the entire duration, only 10 weeks were chosen to make the sample size the same as the first wave, so that the two waves can be compared more accurately.

# summary table (total cases per week in 2nd wave)
second_wave_summary = data.frame(week_ending = second_wave$week_ending[1:10],
                                weekly_total = second_wave$weekly_total[1:10])
second_wave_summary
```


#### b) Exploratory analysis

Explore the weekly number of cases for the two waves visually using a boxplot, histograms, and qq plots, and perform a test for normality.

```{r}
combined = rbind(first_wave_summary, second_wave_summary)
combined$wave = c(rep("first_wave", 10), rep("second_wave", 10))

# boxplot
ggplot(data = combined,
       mapping = aes(x = week_ending,
                     y = weekly_total,
                     fill = wave)) +
  geom_boxplot() +
  geom_jitter(position = position_jitter(0.1))
  
# histograms and qqnorm plots
hist(first_wave_summary$weekly_total, breaks = 10)
hist(second_wave_summary$weekly_total, breaks = 10)
hist(combined$weekly_total, breaks = 10)
 
qqnorm(first_wave_summary$weekly_total)
qqline(first_wave_summary$weekly_total)

qqnorm(second_wave_summary$weekly_total)
qqline(second_wave_summary$weekly_total)

qqnorm(combined$weekly_total)
qqline(combined$weekly_total)

# normality test
shapiro.test(first_wave_summary$weekly_total)
shapiro.test(second_wave_summary$weekly_total)
shapiro.test(combined$weekly_total)
```

#### c) Data transformations

If the data look highly skewed, it might be useful to transform the data. Try some different transformations and see if they help make the data look more normal.

```{r}

# Although the histograms do not look like normal distributions due to the small sample size, the qqplot and Shapiro normality test show that the data are normally distributed. So, data does not need to be transformed. I discussed this with Kris and was asked to do the following instead as an alternative to the data transformations.

# When would a log-transformation help make your data look more normal, and when would this not help?
# log-transformation would help if the frequency distribution of the data are skewed to the right, and usually when the measurements are ratios or products of variables, the group having the larger mean also has the higher standard deviation, and the data spans several orders of magnitude. log-transformations would not help if the frequency distribution is left-skewed because of the differences in the nature of log values at small and large values.  

```

Compare the results of a parametric and a non-parametric test and discuss why a parametric test is preferred to a non-parametric test when the data are normal.

```{r eval=F}
t.test(first_wave_summary$weekly_total, second_wave_summary$weekly_total)
wilcox.test(first_wave_summary$weekly_total, second_wave_summary$weekly_total)

# These tests show that the p-values are different from the two tests, specifically the non-parametric test overestimates the p-value (although both show significant difference in this particular case). In general, non-parametric tests have lower power than parametric tests as usually non-parametric tests use ranks which discard most of the information in the data such as magnitude of the differences. With less power, there is a lower probability of rejecting a false null hypothesis and therefore there is a higher Type II error rate. So, when the data are normal, parametric tests work better than non-parametric tests.
```

#### d) Statistical test

Run an appropriate statistical test to compare the number of cases in the first and second wave. 

```{r}
# statistical test
t.test(first_wave_summary$weekly_total, second_wave_summary$weekly_total)
```

Which wave was bigger?

```{r eval=F}
# Since the p-value is 0.003714 which is smaller than the significance level of 0.05, the mean number of cases in the first and second wave are indeed different (null hypothesis rejected). The mean number of cases in the second wave is significantly greater than that in the first wave.

```


<!-- ======================================================================= -->
### 3) Changes in case rates over time

We cannot directly compare the number of cases in each age category because we do not know the total number of population in each group (e.g. are there more cases among 75+ year olds than among 0-4 year olds because they get sick more or because there are more of them to begin with?) 

But we can look whether the **proportion** of cases that fall within each age group (out of the total number of weekly cases summed up across all age groups) differs with time. 

For example, is there a significant difference between the proportion of infections among younger adults vs. older adults in the first and second waves? What about before and after the appearance of the Delta variant?

#### a) Younger vs. older adults in waves 1 and 2

Did the age distribution of cases differ between the first two big waves of cases?

First, subset the data to get the percentage of cases in adults under 35 in each wave. Make a box plot to compare them and perform an appropriate statistical test for significance between the rates in the two waves.

Then, do the same thing for adults 65 and up.

```{r}
# ============================================================================ #
# young adults (18-34) - 1st vs. 2nd wave
# ============================================================================ #
# first wave

first_young_adults = filter(first_wave, age_group == "age_18_24" | age_group == "age_25_34")

# second wave
second_young_adults = filter(second_wave, age_group == "age_18_24" | age_group == "age_25_34")

# boxplot
boxplot(first_young_adults$pct_cases, second_young_adults$pct_cases)

# statistical test
shapiro.test(first_young_adults$pct_cases)
shapiro.test(second_young_adults$pct_cases)

t.test(first_young_adults$pct_cases, second_young_adults$pct_cases)
# ============================================================================ #

# older adults (65 and up) - 1st vs. 2nd wave
# ============================================================================ #
# first wave
first_older_adults = filter(first_wave, age_group == "age_65_74" | age_group == "age_75up")

# second wave
second_older_adults = filter(second_wave, age_group == "age_65_74" | age_group == "age_75up")

# boxplot
boxplot(first_older_adults$pct_cases, second_older_adults$pct_cases)

# statistical test
shapiro.test(first_older_adults$pct_cases)
shapiro.test(second_older_adults$pct_cases)

t.test(first_older_adults$pct_cases, second_older_adults$pct_cases)
```

How did the rates of infection in these two groups differ between the first and second waves? Can you speculate on the reason for any observed differences?

```{r eval=FALSE}
# For the young adults, the rate of infection is significantly higher in the second wave as compared to the first wave. Whereas, for the older adults, the rate of infection is significantly higher is the first wave as compared to the second wave. This may be because there was already a greater mortality in the older adult population in the 1st wave, so incidence is lower in this group in the 2nd wave. Another reason may be that older adults, deemed most vulnerable, mostly stayed indoors during the 2nd wave, while younger adults went out more.

```

#### b) Bootstrap confidence intervals

Perform a bootstrap analysis to get estimates for the median and 95% confidence intervals for young adults and older adults in each of the two waves.

```{r}
# ============================================================================ #
# young adults (18-34) - 1st vs. 2nd wave
# ============================================================================ #
# resampling for first wave

young_first_medians = c()
for (i in 1:1000){
  sample_i = sample(x = first_young_adults$pct_cases,
                     size = length(first_young_adults$pct_cases),
                     replace = TRUE)

  sample_i_median = median(sample_i)
  young_first_medians = c(young_first_medians, sample_i_median)
}

# quantiles for first wave (2.5%, 50% (median), 95.7%)
first_young_quant = quantile(x = young_first_medians,
                             probs = c(0.025, 0.5, 0.975))

# resampling for second wave

young_second_medians = c()
for (i in 1:1000){
  sample_i = sample(x = second_young_adults$pct_cases,
                     size = length(second_young_adults$pct_cases),
                     replace = TRUE)

  sample_i_median = median(sample_i)
  young_second_medians = c(young_second_medians, sample_i_median)
}

# quantiles for second wave (2.5%, 50% (median), 95.7%)
second_young_quant = quantile(x = young_second_medians,
                             probs = c(0.025, 0.5, 0.975))

# ============================================================================ #
# older adults (65 and up) - 1st vs. 2nd wave
# ============================================================================ #
# resampling for first wave

older_first_medians = c()
for (i in 1:1000){
  sample_i = sample(x = first_older_adults$pct_cases,
                     size = length(first_older_adults$pct_cases),
                     replace = TRUE)

  sample_i_median = median(sample_i)
  older_first_medians = c(older_first_medians, sample_i_median)
}

# quantiles for first wave (2.5%, 50% (median), 95.7%)
first_older_quant = quantile(x = older_first_medians,
                             probs = c(0.025, 0.5, 0.975))

# resampling for second wave

older_second_medians = c()
for (i in 1:1000){
  sample_i = sample(x = second_older_adults$pct_cases,
                     size = length(second_older_adults$pct_cases),
                     replace = TRUE)

  sample_i_median = median(sample_i)
  older_second_medians = c(older_second_medians, sample_i_median)
}

# quantiles for second wave (2.5%, 50% (median), 95.7%)
second_older_quant = quantile(x = older_second_medians,
                             probs = c(0.025, 0.5, 0.975))

print("Younger Adults First Wave Quantiles:")
first_young_quant
print("Younger Adults Second Wave Quantiles:")
second_young_quant
print("Older Adults First Wave Quantiles:")
first_older_quant
print("Older Adults Second Wave Quantiles:")
second_older_quant

```


<!-- ======================================================================= -->
### 4) Delta strain

The Delta strain became prevalent - i.e. started accounting for >50% all cases - on 6/23/21. Is there a significant difference between the proportion of infections among 25-34 year-olds before and after the rise of the Delta strain? 

#### a) Before and after Delta

Extract the data for the 25-34 age group from your long table and add a column that labels the data for each week as "Before_Delta" or "After_Delta" based on the date of appearance of the Delta variant.

```{r}
# filter by age and label before and after groups
delta_25_34 = filter(covid_long, age_group == "age_25_34")
delta_25_34$delta = c(NA)
delta_cutoff = which(delta_25_34$week_ending == "2021-06-26")

delta_25_34$delta[1:(delta_cutoff - 1)] = "Before_Delta"
delta_25_34$delta[delta_cutoff:length(delta_25_34$week_ending)] = "After_Delta"

# change factor order (optional)
delta_25_34$delta = factor(delta_25_34$delta, levels = c("Before_Delta", "After_Delta"))

delta_25_34
str(delta_25_34)
```

#### b) Violin plot

Make a violin plot of the percentage of cases observed per week in the 25-34-year old group before and after the appearance of the Delta variant.

```{r}
# ggplot violin plot (optional: put a narrow boxplot inside)
ggplot(data = delta_25_34,
       mapping = aes(x = delta,
                     y = pct_cases)) +
  geom_violin(aes(fill = delta),
              alpha = 0.5) +
  geom_boxplot(width = 0.2) 
```

#### c) Permutation test

Perform a permutation ("shuffle") test for the observed median difference in the percentage of total cases in the 25-34 year-old age group before and after Delta. 

```{r}
# observed difference in rates
obs_diff = diff(median(pct_cases ~ delta, data = delta_25_34))
obs_diff


# shuffle test
shuffle_test = do(2000) * diff(median(pct_cases ~ shuffle(delta), data = delta_25_34))

# check column names in permuted data structure and rename (optional)
# this is because the permutation may change the name for the percent cases column
colnames(shuffle_test)
colnames(shuffle_test) = "median_differences_in_pct_cases"

# draw a histogram (optional)
gf_histogram(gformula = ~ median_differences_in_pct_cases,
             fill = ~ (median_differences_in_pct_cases >= obs_diff),
             data = shuffle_test,
             binwidth = 0.4,
             xlab = "Distribution of difference in medians under the null hypothesis")
```

What's the $p$-value you get from the shuffle test? Compare this with with the result of another appropriate statistical test.

```{r}
# p-value (permutation)
sum(shuffle_test >= obs_diff) / 2000

# p-value (statistical test)
shapiro.test(delta_25_34$pct_cases[delta_25_34$delta == "Before_Delta"])
shapiro.test(delta_25_34$pct_cases[delta_25_34$delta == "After_Delta"])

t.test(delta_25_34$pct_cases[delta_25_34$delta == "Before_Delta"],
       delta_25_34$pct_cases[delta_25_34$delta == "After_Delta"])

# The p-value from the permutation test is 0 and that from the t-test is also close to 0 (0.0002). So, the percentage of cases after delta is significantly greater in this age group as compared to that before delta.

```


<!-- ======================================================================= -->
### 5) Breakthrough cases

Consider the following data of the number of NYC documented Covid cases, hospitalizations, and deaths among vaccinated and unvaccinated individuals:

+ Cases:
  + Vaccinated: 820
  + Unvaccinated: 4140  

+ Hospitalizations:
  + Vaccinated: 32
  + Unvaccinated: 349

+ Deaths:
  + Vaccinated: 4
  + Unvaccinated: 39


What is an appropriate test you could use to analyze these data? Explain.

```{r eval=FALSE}
# Since these are categorical variables with numerical data, and because the contingency table will be greater than 2X2, chi-square test can be used.

```

Use this data to create a contingency table, and then perform the appropriate test to determine whether there is a significant difference between the vaccinated and unvaccinated groups.

```{r}
# data frame for given values
vaccinated_cases = 820
unvaccinated_cases = 4140
vaccinated_hospitalizations = 32
unvaccinated_hospitalizations = 349
vaccinated_deaths = 4
unvaccinated_deaths = 39

breakthrough_cases = rbind(c(vaccinated_cases, unvaccinated_cases),
                           c(vaccinated_hospitalizations, unvaccinated_hospitalizations),
                           c(vaccinated_deaths, unvaccinated_deaths))
rownames(breakthrough_cases) = c("Cases", "Hospitalizations", "Deaths")
colnames(breakthrough_cases) = c("Vaccinated", "Unvaccinated")

# pretty output formatting
knitr::kable(breakthrough_cases)

# statistical test(s)
chisq.test(breakthrough_cases)

# p-value is smaller than 0.05, so there is a significant difference between the vaccinated and unvaccinated groups.
```

<!-- THE END! :) =========================================================== -->







