---
title: 'In-class exercises: Permutations, Combinations, Binomial Proportions'
subtitle: "XDASI Fall 2021"
date: "9/30/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = FALSE)
```

## Practice problems

### 1. Nucleic acids: Sampling, permutations and combinations

a. Restriction enzymes recognize specific sequences in DNA and cut the DNA within or near those sites. How many possible restriction sites of length 6 are there?
```{r}

```

b. At what frequency would you expect to find a binding site for a particular restriction enzyme that recognizes a 6 bp sequence? _Note: Restriction enzymes usually recognize a palindromic sequence, so you don't need to worry about looking on both strands._
```{r}

```

c. What is the average length of a fragment produced by a 6-cutter?
```{r}

```

d. The EcoR1 enzyme cuts the sequence "GAATTC". How many different sequences could be made out of this set of nucleotides (assume that you treat each A and T as distinct individuals)?
```{r}

```

d. You are interested in synthesizing a bunch of random oligonucleotides for a SELEX experiment. How many possible oligos of length 22 are there?
```{r}

```


e. How many ways could you make a sequence of 6 nt by grabbing them at random from a bag of 12 nucleotides (assuming each base is treated as a distinct unique entity, and you can only pick each nt once)? This is an example of "random sampling without replacement".
```{r}

```

f. How many combinations of 6nt are there in a set of 12 random nucleotides? (That is, pick any group of 6 nt, where each one is treated as unique, and you don't care about the sequence in which you pick them.)
```{r}

```

g. How are your answers in parts e and f related?

```{r eval=FALSE}

```


### 4. Wnt signaling: Binomial proportions

Proliferation of embryonic stem cells is important for early development and is promoted by Wnt signaling, which promotes cell cycle progression through the transcriptional activator beta-catenin. In a study of the developing mouse, it was found that only 11% of cells in a region of the brain that is rich in neuronal progenitors was responsive to Wnt (even though Wnt is required for stem cell expansion).

***Note: You may find the `choose()` function to be useful for parts of this question.***

a. If you were to perform FACS on cells from this region that are labeled with a fluorescent antibody against the Wnt receptor, what is the probability of observing the following outcomes among a sample of 6 cells? Below, let's call Wnt-responsive cells = W and Wnt-insensitive cells = I.

```{r}
W = 0.11
I = 0.89

# 1. A sequence of WWIIII

# 2. A sequence of IIWWII

# 3. A string of WWWWWW

# 4. At least one W cell out of 6

```

b. How many ways are there to get four Wnt-insensitive cells out of 6 total? (This is the same as the number of ways to get two W cells out of 6.)
```{r}

```

c. What is the total probability of getting exactly 4 I cells and 2 W cells in your sample, taking into account all the different ways that this outcome can be obtained?
```{r}

```

d. What would be the probability of getting 2 or less W cells in a sample? (Consider the probability of getting either 0, 1, or 2 W cells.)
```{r}

```

---