---
title: "Exercise: Exponential and Poisson"
subtitle: "XDASI Fall 2021"
date: "10/21/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Example: Lizard mortality

For some animals, like lizards (or hydra), their mortality rate is independent of their age. Over an extended trip to Costa Rica, you studied a large population of lizards and found that their rate of mortality is $\lambda$ = 0.05 deaths per month.

$\Rightarrow$ ***What is the expected number of deaths per year?***

```{r}

```


$\Rightarrow$ ***What is the expected lifespan of one of these lizards?*** (In other words, this is the typical "wait time" until one death occurs?)

```{r}

```


$\Rightarrow$ ***What is the probability an animal will survive more than 20 months?*** 

Hint: 

+ This is given by the the exponential survivorship function.
+ Alternatively, it is given by the lower-tail Poisson probability (use $\mu$ formulation).

```{r, collapse = T}
# with exponential

# with Poisson

```

$\Rightarrow$ ***What is the probability an animal will survive 20 months or less?***

Hint: 

+ This is given by the lower-tail CDF of one distribution, and the upper-tail CDF of the other distribution.

```{r, collapse = T}
# using exponential

# using poisson

```


$\Rightarrow$ ***What is the probability an animal will survive between 10 and 15 months?*** 

Hint: 

+ To get the total probability across an interval, we subtract the total probability of the lower value from that of the higher value.
+ We can't answer this with Poisson, since can only vary # events (not time)!

```{r, collapse = T}

```

