---
title: "Data Manipulation using `dplyr`"
date: "September 17, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(ggplot2)
```

# Our World in Data: the COVID19 pandemic

[Our World in Data](https://ourworldindata.org/) publishes data, visualizations, and analyses of issues of global importance, such as poverty, climate change, and disease. Today, we will analyze a data set on COVID19 compiled by them. It contains profiles of 207 countries or territories. The data include the **daily** number of cases, deaths, vaccinations etc., and they are accompanied by some useful country stats, e.g. their population. We downloaded the data set (version 44) [from kaggle](https://www.kaggle.com/tunguz/data-on-covid19-coronavirus) but only kept a few selected columns.

## Look at the data

First, import and examine the data.
```{r}
covid <- read.csv("owid-covid-data_filtered.csv")
head(covid)
tail(covid)
str(covid)
```

## Set an ambitious goal

**Determine whether European countries that have the highest total number of COVID19 cases normalized to population size also have the highest number of COVID19-related deaths.**

## Split it into steps

First, describe each step in words.
```
1. Subset countries located in Europe.
2. For each country, sum up new cases and deaths across all recorded days to get their total numbers.
3. Normalize the number of cases and deaths to the population size.
4. Sort the data by the normalized total number of cases and plot them.
```

Next, write `dplyr` code for each individual step.
```{r}
# 1.1. subset countries located in Europe
covid_europe <- filter(covid,
                       continent == "Europe")

head(covid_europe)

# 1.2. the same, but using piping
covid_europe <-
  covid %>%
  filter(continent == "Europe") # no need to have covid inside parenthesis because it has already been piped into the function

head(covid_europe)

# 2.1. group by country - the data won't change but this step is necessary before performing any sort of calculations that need to be done for each group separately
covid_europe_grouped <-
  covid_europe %>%
  group_by(location)

head(covid_europe_grouped) # the data are now a tibble (think of it as an enhanced data.frame) - note the new attribute "Groups"

# 2.2. sum up cases and deaths across days for each country (=for each group)
covid_europe_grouped_totals <-
  covid_europe_grouped %>%
  summarize(total_cases = sum(new_cases, na.rm = TRUE), # na.rm=TRUE makes sum ignore NA values
            total_deaths = sum(new_deaths, na.rm = TRUE), # na.rm=TRUE makes sum ignore NA values
            population = mean(population)) # could have been max or min - population is likely the same for each day

head(covid_europe_grouped_totals)

# 3. normalize cases and deaths to the population size
covid_europe_grouped_totals_norm <-
  covid_europe_grouped_totals %>% 
  mutate(total_cases_per_thousand = 1000*total_cases/population,
         total_deaths_per_100k = 100000*total_deaths/population) # we arbitrarily normalize by 100k instead of 1k people to make the number of cases and death easily comparable when plotted on the same scale

head(covid_europe_grouped_totals_norm)

# 4.1. sort in descending order
covid_europe_grouped_totals_norm_sorted <-
  covid_europe_grouped_totals_norm %>%
  arrange(desc(total_cases_per_thousand)) # use desc() to sort in descending order

head(covid_europe_grouped_totals_norm_sorted)

# 4.2. only keep those columns that are relevant
covid_europe_grouped_totals_norm_sorted_selected <-
  covid_europe_grouped_totals_norm_sorted %>%
  select(location, total_cases_per_thousand, total_deaths_per_100k)

head(covid_europe_grouped_totals_norm_sorted_selected)

# 4.3. transform the data into the long format
covid_europe_grouped_totals_norm_sorted_selected_long <-
  covid_europe_grouped_totals_norm_sorted_selected %>%
  gather(key = "metric",
         value = "value",
         "total_cases_per_thousand",
         "total_deaths_per_100k") # gather() is actually from the package tidyr, but it is compatible with dplyr

head(covid_europe_grouped_totals_norm_sorted_selected_long)
tail(covid_europe_grouped_totals_norm_sorted_selected_long)

# 4.4. convert "location" to a factor and force its levels to maintain the same order as in the sorted data frame (otherwise, countries will be sorted alphabetically by ggplot2)
covid_europe_grouped_totals_norm_sorted_selected_long$location <- 
  factor(covid_europe_grouped_totals_norm_sorted_selected_long$location,
         levels = unique(covid_europe_grouped_totals_norm_sorted_selected_long$location)) # we use unique() here because each country name is repeated twice in the tibble, and level names must be unique. this trick works here but won't necessarily work elsewhere
```

Finally, plot the data on a `ggplot2` bar graph. Make sure that the bars representing number of cases and number of deaths for each country are plotted side-by-side.
```{r}
ggplot(data = covid_europe_grouped_totals_norm_sorted_selected_long,
       mapping = aes(x = location,
                     y = value,
                     fill = metric)) +
  geom_col(position = "dodge") +
  coord_flip()
```


## Combine all `dplyr` steps into a single pipeline

Join all the individual steps with `%>%`, feed the input data.frame into the pipeline, and assign the output tibble to a variable.
```{r}
covid_dplyr <- 
  covid %>%
  filter(continent == "Europe") %>%
  group_by(location) %>% 
  summarize(total_cases = sum(new_cases, na.rm = TRUE),
            total_deaths = sum(new_deaths, na.rm = TRUE),
            population = mean(population)) %>% 
  mutate(total_cases_per_thousand = 1000*total_cases/population,
         total_deaths_per_100k = 100000*total_deaths/population) %>% 
  arrange(desc(total_cases_per_thousand)) %>%
  select(location, total_cases_per_thousand, total_deaths_per_100k) %>%
  gather(key = "metric",
         value = "value",
         "total_cases_per_thousand",
         "total_deaths_per_100k") %>%
  mutate(location = factor(location, levels = unique(location)))

covid_dplyr
```

Plot the data again to make sure the plot looks the same.
```{r}
ggplot(data = covid_dplyr,
       mapping = aes(x = location,
                     y = value,
                     fill = metric)) +
  geom_col(position = "dodge") +
  coord_flip()
```

