```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Scenario

You want to clone a gene and you have some reporter for success of transformation in *E. coli*. Positive colonies are blue because they express $\beta$-galactosidase when plated on medium with IPTG. The probability of a successful transformation is $0 \le P(success) \le 1$. For the following exercise, you may consider one plate, many plates, or a truly gargantuan number of plates.

## Exercise

The class will split into 8 groups and each one will take one of the following distributions:

- Bernoulli
- Binomial
- Geometric
- Hypergeometric
- Negative binomial
- Poisson
- Normal
- Exponential

For your assigned distribution, answer the following questions. At the end of the class, each group will present their distribution and how it applies to the scenario above to the rest of the class.

#### a. What is your function? Describe it in terms of the kind of problem that it applies to. Is it discrete or continuous?

<!-- Insert your answer below. -->
Geometric function
It gives the probability of getting x number of failures before a success 

#### b. Describe a question that you can ask with your distribution as applied to the transformation scenario given above.

<!-- Insert your answer below. -->
When randomly picking colonies from yor plate, what is the probability of getting 6 white colonies before getting a green one?

#### c. Describe the parameterization of the equation. What parameters does it take, and what do they mean?

<!-- Insert your answer below. -->
The equation takes two parameters:
x - number of failures
p - probability of success

#### d. Pick some appropriate parameters for your distribution as applied to the transformation scenario and graph the PDF and CDF.

```{r}
p = 0.2
par(mfrow=c(1,2))
plot(0:20, dgeom(0:20,prob = p), type = "l", main = "PDF")
plot(0:20, pgeom(0:20,prob = p), type = "l", main = "CDF")

```


#### e. Compute a cumulative probability for some appropriate value and describe in words what it means. Be prepared to explain your answer to the class.

```{r}
pgeom(6,prob = p)

```

<!-- Insert your answer below. -->
The probability of getting 6 and less white colonies before getting a green colony is 0.79

#### f. Identify at least one other distribution that yours relates to and be prepared to discuss the relationship between them. This could be a special case, a limiting case, an approximation, or an inverse CDF (survival) relationship.

<!-- Insert your answer below. -->
Negative binomial. The geometric function is a specific case of the negative binomial where r;number of successes is 1.