```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Scenario

You want to clone a gene and you have some reporter for success of transformation in *E. coli*. Positive colonies are blue because they express $\beta$-galactosidase when plated on medium with IPTG. The probability of a successful transformation is $0 \le P(success) \le 1$. For the following exercise, you may consider one plate, many plates, or a truly gargantuan number of plates.

## Exercise

The class will split into 8 groups and each one will take one of the following distributions:

- Bernoulli
- Binomial
- Geometric
- Hypergeometric
- Negative binomial
- Poisson
- Normal
- Exponential

For your assigned distribution, answer the following questions. At the end of the class, each group will present their distribution and how it applies to the scenario above to the rest of the class.

#### a. What is your function? Describe it in terms of the kind of problem that it applies to. Is it discrete or continuous?

The negative binomial distribution:

$$P(X=x)=\binom{x+r-1}{r-1}p^r(1-p)^x$$
It is discrete. It's often applied to situations where you have binary outcomes, and you want to stop after getting a certain number of one of those outcomes.

#### b. Describe a question that you can ask with your distribution as applied to the transformation scenario given above.

What is the probability of getting a certain number of transformants after transforming and plating a certain number of cells?


#### c. Describe the parameterization of the equation. What parameters does it take, and what do they mean?

x is the number of failures. r is the number of successes. p is the probability of success. The equation multiplies the probabilities of all successes by the probabilities of all failures.


#### d. Pick some appropriate parameters for your distribution as applied to the transformation scenario and graph the PDF and CDF.

```{r}

require(ggplot2)

r=100
p=0.0000025
nmax=10^8
nmin=10^7
xmax=nmax-r
xmin=nmin-r
xrange=seq(xmin,xmax,by=100000)

pmf <- dnbinom(xrange,r,p)
pmf.data <- as.data.frame(pmf)
ggplot(pmf.data,aes(x=xrange,y=pmf))+
  geom_line(color="purple")+
  xlab("X")+
  ylab("Probability")

cdf <- pnbinom(xrange,r,p)
cdf.data <- as.data.frame(cdf)
ggplot(cdf.data,aes(x=xrange,y=cdf))+
  geom_line(color="purple")+
  xlab("X")+
  ylab("P(X<=x)")



#There is another useful way to use this function. Like what is the #probability of getting 100 or less transformants. This could be #calculated by summing over the range 1:100 in dbinom's size #parameter (r).


```


#### e. Compute a cumulative probability for some appropriate value and describe in words what it means. Be prepared to explain your answer to the class.

```{r}

cdf2 <- pnbinom(5*10^7,r,p)
cdf2


```

It means that in the range from 0 to the specified number of cells, the total probability of getting 100 transformants is very close to 1. 


#### f. Identify at least one other distribution that yours relates to and be prepared to discuss the relationship between them. This could be a special case, a limiting case, an approximation, or an inverse CDF (survival) relationship.

The geometric distribution is identical to the negative binomial distribution when r=1.
