---
title: "Data Visualization"
date: "September 9-10, 2021"
output: 
  html_document:
    toc: TRUE
    toc_depth: 2
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```


# Output devices

To make any kind of plot, you need some sort of "canvas" to draw on. The ***device*** directive controls the output stream, i.e. where your plot is drawn. This may be somewhere on your screen, or else some type of output file (e.g. PDF, PNG, Postcript, etc.)

## Default device

If you don't specify a particular device as the output destination, then by **default**, any plots you make will be displayed directly on your screen. This default output device is called the `null device`.

Plots will be printed to your screen in one of three ways within RStudio:

+ in the **Plots** window
  + if you execute a command from the console or an R script using the `null device`
+ **inline**, if the command is executed from within an R markdown
+ in a **new window** on your screen
  + if you open a new screen device (see below)

For these exercises, we will use the built-in `iris` dataset, which contains measurements on sepals and petals of three different species of iris. Its class is `data.frame`.

Try issuing the following commands, first within this .Rmd file, and then from the console:

```{r}
# plot iris petal length vs. width
plot(iris$Petal.Length, iris$Petal.Width)
plot(iris$Petal.Length, iris$Petal.Width, col="blue")
plot(iris$Petal.Length, iris$Petal.Width, col="purple")
```

You've just written three different plots to the **Plots** window. You can scroll through them using the left and right arrows, and you may pop them out into a new window or save them to a file (read on).

# Checking devices

You can easily find out what your current device is, or list all the devices you have open, using the following commands. 

```{r}
dev.cur()   # current device (a.k.a. device 1)
dev.list()  # list all open devices (only one should be listed right now)
```

# Opening a new device

You can make a new canvas by opening a new output stream. 

When you do this using `windows()` (on a PC), or `quartz()` (on a Mac), a new window should open somewhere on your desktop (sometimes it could appear in front of the RStudio, but sometimes it might open behind RStudio and then you'll have to look for it).

Try the following commands from the **console** (nothing will happen if you do this from within this Markdown):

```{r, eval=FALSE}
# ================================================== #
# open a new window
# windows()  # on a pc
quartz()     # on a mac
dev.cur()    # now "quartz 2" instead of "null device 1"

# ================================================== #
# a scatter plot of petal width vs. length
# should appear in the new window
plot(iris$Petal.Length, iris$Petal.Width)
plot(iris$Petal.Length, iris$Petal.Width, col="blue")

# ================================================== #
# close the device
dev.off()  # window will disappear!
dev.cur()  # back to the null device

# plot again is directed to the Plots window
plot(iris$Petal.Length, iris$Petal.Width)
```

New plots will just overwrite old plots in the same device, unless you open a new device to create a new canvas (or you are working in an R Markdown file). Check that this happened above. 

As noted above, in RStudio the behavior is different for the `null device` because it will show you all your plots in the Plots window.

## Writing to files

You can also redirect your output to a file. Different file devices include: PDF, jpeg, png, PostScript, bitmap, and LaTeX.

PDF output is really useful. Let us write a plot to a pdf file. You may do this either from within this .Rmd file, or on the console.

```{r, collapse=TRUE}
# ================================================== #
# plot to PDF file in RMD (execute all at once)
pdf("iris_plot.pdf")
dev.cur()
plot(iris$Petal.Length, iris$Petal.Width, col="purple")
dev.off()

# ================================================== #
# repeat in console
pdf("iris_plot.pdf")
plot(iris$Petal.Length, iris$Petal.Width, col="magenta")
dev.off()

dev.cur()
```

Note that since you wrote the output to the same file name, only the second plot remains. You can make multi-page PDFs instead by sending both plots to the open PDF device before you close it:

```{r}
# ================================================== #
# two plots in the same PDF file, each gets its own page
pdf("iris_plot.pdf")
dev.cur()
plot(iris$Petal.Length, iris$Petal.Width, col="purple")
plot(iris$Petal.Length, iris$Petal.Width, col="magenta")
dev.off()
```


# Essentials of base R graphics

Here we introduce the basic elements of the <a href="https://www.rdocumentation.org/packages/graphics/versions/3.6.2"

target="_blank">**graphics**</a> package, which is part of the base R distribution. 

The **graphics** package provides ***functions*** for creating all kinds of plots. Many of these can be used to conveniently get a quick idea of what your data look like. They are simple yet powerful.

There are also many graphical ***parameters*** that can be set in order to control the appearance of points and lines, axis ticks, plot labels, text, and arrangement of plots when multiple plots are generated at the same time. Other packages will come with their own plotting functions, but generally they access the same range of parameters.

Learning to fine-tune the appearance of plots using the base graphics package can be very tedious and unrewarding. This is why we will transition to mainly using the **ggplot2** package to take our graphics to the next level. Nevertheless, we think it is important for you to know that these base functions exist and are available if you want to use them.


## The `plot()` function: automatic plot

The generic function for plotting R objects is aptly called `plot()`. The default output
of the function depends on the object it is passed.

`ldeaths` is an object of class `ts` (time series) that is provided in R.
This dataset provides the number of deaths in UK from 1974-1979 due to lung disease.
Since it is a numerical vector, the plot function plots a line.

```{r}
plot(ldeaths)
```

For the `iris` data frame, `plot()` shows all comparisons between the 5 columns in the data frame.

```{r}
head(iris)
str(iris)
plot(iris)
```


## The `plot()` function: scatter plot

Rather than relying on the ability of `plot()` to automatically recognize the structure of your R object, you can use this function to create simple scatter plots by explicitly specifying `x` and `y` parameters. We did this above already.

```{r}
plot(x=iris$Sepal.Length,
     y=iris$Sepal.Width)
```


# Customizing parameters

Above, we also drew the points in different colors. You may also want to tweak some other things, like the shape and size of the points, and the axis labels, or overlay additional information, such as a regression line.

Somewhat confusingly, the `plot()` help file (see `?plot`) does not describe any parameters to specify color. Instead, it says:
`... Arguments to be passed to methods, such as graphical parameters (see par).`

So you need to look up `par` in the Help to get a list of these. You can set parameters globally, or within a plot function. See <a href="https://www.statmethods.net/advgraphs/parameters.html" 

target="_blank">**QuickR: Graphical Parameters**</a> by Datacamp for more info on these parameters.

```{r}
# ================================================== #
# fancy plot
x=iris$Sepal.Length
y=iris$Sepal.Width

plot(x, y,
     pch = 15, # 23            # shape of data points
     col = "blue",             # color of data points
     cex = 0.5,                # size  of data points
     xlab='Sepal length, mm',  # nicer axis labels
     ylab='Sepal width, mm')

# ================================================== #
# add a regression line
# this throws an error if try to use dataframe cols directly
# this is because the function wants a matrix or vector object
abline(lm(y ~ x))
```


# Coloring by dataset

Above we used the `col` parameter to change the color of *all* the datapoints.

Even better, we can also color the points by **factor groupings**. 

## Map colors to factor columns in a data frame

Below we color the points from the three *Species* according to the three *factors* defined for this data column. 

```{r}
plot(x=iris$Sepal.Length,
     y=iris$Sepal.Width,
     col=iris$Species)   # color the points by their species group/factor
```


## Map data with a different palette

Don't like the default colors? Change the palette!

```{r}
# ================================================== #
# new palette with 3 weird colors
# BEWARE: these will just get recycled if there are more factors than colors!
palette(c("deepskyblue", "darkorchid1", "darkorange2"))
plot(x=iris$Sepal.Length,
     y=iris$Sepal.Width,
     col=iris$Species)

# ================================================== #
# change back to the (ugly) R default palette!
palette("default")
plot(x=iris$Sepal.Length,
     y=iris$Sepal.Width,
     col=iris$Species)
```

Note that R seems to know how to map `palette()` values onto factors automatically. 

  + *HOWEVER, if the number of factors exceeds the number of colors in the palette, colors will just get recycled! (You can try this with the iris dataset by removing one of the colors above and remaking the plot.)*

What if instead we just tried to use a random vector of three colors? Points will just get randomly assigned to colors! 

  + *Just like with the palette colors, these colors will get recycled until all of the points are colored; however there will be no relationship between factors and the color of each data point.*

Compare the graph below with the ones we just made above. Can you see the difference?

```{r}
# use 3 manually assigned colors
plot(x=iris$Sepal.Length,
     y=iris$Sepal.Width,
     col=c("red","green","blue"))
```

For more on palettes, see the <a href="https://kriscgun.github.io/xdasi-bio-2021/r_resources/4_graphics/4_color.html#palettes"

target="_blank">**Palettes**</a> 
section of the course website ("R Resources=>Data Visualization=> Working with Color").


## Map data with a character column

Abother way to color the data by groups/factors is using a *character vector* of the same length as the factor vector in the dataframe. This allows us to change the color mappings, but it is more complicated!

```{r}
# ================================================== #
# make a copy of the Species column and check the levels
species_colors = iris$Species  # a factor column
levels(species_colors)         # also a factor vector

# ================================================== #
# set color names as levels
levels(species_colors) = c("orange","purple","aquamarine")

# what is this vector? is it still a factor?
head(species_colors)
tail(species_colors)
str(species_colors)

# try this
plot(iris$Sepal.Length,
     y=iris$Sepal.Width,
     col=species_colors)
```

Huh? what happened? The colors were not applied!

We need to convert the factor vector to a character vector first. Why? perhaps because the vector is not part of the same data frame, or because the factors don't match? (I don't fully understand this.)

```{r}
# ================================================== #
# convert to character vector and replot
species_colors = as.character(species_colors)
plot(x=iris$Sepal.Length,
     y=iris$Sepal.Width,
     col=species_colors)
```


# Essentials of ggplot2

The following brief overview is adapted from an <a href="http://www.sthda.com/english/wiki/ggplot2-essentials" 

target="_blank">**STHDA tutorial**</a>
that covers all the basic types of plots you can make with ggplot. Also see the **ggplot2** pages on the course website ("R Resources=>Data Visualization=>ggplot2")

The concept behind ggplot2 divides plot into **three different fundamental** parts: 

**Plot = Data + Aesthetics + Geometry**

The principal elements of every plot can be defined as follows:

+ **Data** is a data frame (variables) to be plotted.
+ **Aesthetics**: the `aes()` function is used to indicate ***how*** to display the data:
  + which categories or measurements to map to x and y coordinates; or color, size or shape of points, etc.
+ **Geometry** defines the ***type of graphics*** 
  + e.g. histogram, box plot, line plot, density plot, dot plot, etc.

## Plotting functions

There are **two major functions** in the ggplot2 package:

+ `qplot()` stands for quick plot, which can be used to produce easily simple plots.
+ `ggplot()` function is more flexible and robust than `qplot()` for building a plot piece by piece.

Plots are constructed by **layering** geometries, additional aesthetics, and themes on top of the primary aesthetic mapping.

The **basic syntax** is:

```{r, eval=FALSE, echo=TRUE}
ggplot(data = <data.frame>,
       mapping = aes(x = <column of data.frame>, y = <column of data.frame>)) +
  geom_<type of geometry>()
```

Notice that as layers are added, a `+` symbol is added at the end of the previous layer. This signals that the plot is not finished. 

***The `+` symbol must always appear at the end of a line.***


## Aesthetics

If your data is **tidy**, then the columns of your data frame will contain the variables that you want to display. Let's take a minute to review tidy data by reviewing the page on the course website. 

Remember how we said that R is a **vectorized** language? Vectors are the basic units of all data structures in R.

So, each **column** of a data frame can be **mapped** to different **aesthetics** of the graph (e.g. axis, colors, shapes, etc.). A few of the examples below are based on Chapter 3 from ***R for Data Science***.

### Specifying aesthetics

There are **two ways** to specify aesthetics:

+ **Mapping**: This maps specific types of **data** to different visual elements
  + Mappings are are always included ***inside*** of an `aes()` directive.
    + In the **data** layer, mapping is applied to **all geometries**.
    + In the **geom** layer (or other layers), the mapping will be applied only to a particular geometry or feature.

+ **Setting**: Manually set an aesthetic **independently** of data
  + Settings are placed ***outside*** of an `aes()` directive.
    
Aesthetic elements include things such as:

+ x- and y- axes
+ Colors
+ The size of points (in mm)
+ The shape of points (a number; see below)
+ Transparency (range 0-1)

## Geometries

**Geometries** control the ***type of visual paradigm*** you want to use to display your data, for example:

+ `geom_bar()`       - barchart
+ `geom_histogram()` - histogram
+ `geom_dotplot()`   - dot plot, a.k.a. strip chart
+ `geom_boxplot()`   - boxplot
+ `geom_violin()`    - violin plot
+ `geom_point()`     - scatterplot

Geom functions also allow you to add additional features to a graph, for example:

+ `geom_jitter()` - spread points out (e.g. on strip charts) to make the data more visible
+ `geom_vline()`  - add a vertical line (can also add other kinds of lines)

Statistical features can also be layered onto graphs, e.g.:

+ `geom_smooth()` - a regression line (according to a global or local model)
+ `stat_summary()` - add some kind of statistical function to a graph
  + This can also be done by adding `stat = "something"` inside another geometry (some examples below)


### Themes

<a href="https://ggplot2.tidyverse.org/reference/theme.html"

target="_blank">**Themes**</a>
are used to customize the **non-data components** of your graphs, such as titles, labels, fonts, background, gridlines, and legends. 

+ The **default** appearance of ggplot graphs produces graphs with a gray background and white gridlines. This can be changed to almost any look and feel by customizing their themes, which can also be used to give plots a consistent look for presentation.

+ The <a href="https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/" target="_blank">**ggthemes**</a>
package provides a variety of defined themes that replicate the look and feel for different visual paradigms and applications. 

+ `theme()` components can also be set manually.


## Plotting functions

### The `qplot()` uses simple syntax to generate plots

The "quick plot" method uses simplified syntax resembling that of Base R. The biggest difference is that we explicitly add the **geometry**.

Let us recreate the `Sepal.Length` vs `Sepal.Width` scatterplot using `qplot()`. We'll also color the points by group, and add some axis labels.

```{r}
library(ggplot2)
head(iris)

# ================================================== #
# quick plot with defaults
qplot(x=Sepal.Length,
      y=Sepal.Width,
      data=iris,
      geom=c("point"),
      color=Species,
      xlab='Sepal length, mm',
      ylab='Sepal width, mm')
```


### The `ggplot()` function builds plots layer by layer

Let us recreate the Sepal.Length vs Sepal.Width scatterplot using `ggplot()`.

```{r}
ggplot(data=iris,                    # specify the data frame
       mapping=aes(x=Sepal.Length,   # add aesthetics
                   y=Sepal.Width)) +
  geom_point() +                     # add geometry
  theme_grey() +                     # add a theme (this is actually the default theme)
  xlab('Sepal length, mm') +         # add axis labels
  ylab('Sepal width, mm')
```

Again, you can set a **color** for all points, or by groups (factors) in the dataset.

```{r}
# ================================================== #
# global aes mapping by factor group
ggplot(data=iris, 
       mapping=aes(x=Sepal.Length,
                   y=Sepal.Width,
                   color=Species)) + # global color
  geom_point() +
  xlab('Sepal length, mm') +
  ylab('Sepal width, mm')

# ================================================== #
# mapping to data to aes can also go inside a geometry
ggplot(data=iris, 
       mapping=aes(x=Sepal.Length,
                   y=Sepal.Width)) +
  geom_point(mapping=aes(color=Species)) # geom color

# ================================================== #
# manual mapping of ALL data to the same color

# can do this globally ...
ggplot(data=iris,
       mapping=aes(x=Sepal.Length,
                   y=Sepal.Width, color="red")) +
  geom_point()

# ... or in the geom layer
ggplot(data=iris,
       mapping=aes(x=Sepal.Length,
                   y=Sepal.Width)) +
  geom_point(color="red")
```


# Histograms

Histograms are good for showing the **distribution** of a **single quantitative variable**. Two or more distributions can be shown together on one histogram, though showing more than two or three gets really confusing. 

## `hist()` in base R

Base R has a special function called `hist()`. It's great for very quick visualization.

```{r}
hist(iris$Sepal.Length)
```

Keep in mind that `iris` contains data for three species, so the histogram above makes little biological sense because it is a mixture of measurements of three different species! There is no way to separate them inside `hist()`, so you have to split the data.frame manually.

Plot labels can get kind of ugly, so it's nice to add labels if you're going to share your plots with someone else.

```{r, collapse=TRUE}
# ================================================== #
# do it in steps

# virgnica_rows = which(iris$Species=='virginica')
# iris_virginica = iris[virgnica_rows,]
# hist(iris_virginica$Sepal.Length)

# ================================================== #
# do it in one go
hist(iris[which(iris$Species=='virginica'), "Sepal.Length"])

# ================================================== #
# prettier
hist(iris[which(iris$Species=='virginica'), "Sepal.Length"],
     xlab='Sepal length, mm',
     main='Sepal length for Virginica')
```


## `qplot()` and `geom_histogram()` in ggplot2

`qplot()` provides an easy way to display the data for the three species overlayed.

```{r}
qplot(x=Sepal.Length,
      data=iris,
      fill=Species,
      color=Species,   # here color applies to the borders!
      alpha=0.2)       # make the bars semi-transparent
```

The same with `ggplot()`:

```{r}
ggplot(data=iris,
       mapping=aes(x=Sepal.Length,
                   fill=Species,
                   color=Species,
                   alpha=0.2))  +
  geom_histogram()
```

Notice that both of these plots do something weird: they include the transparency parameter, `alpha`, in the figure legend. This is because both of these functions are trying to "map" data to an aesthetic, so they automatically include this in the legend!

However, transparency has no meaningful relationship with the data - it's only there to improve the appearance of the histogram. 

To fix this, we need to move the `alpha` parameter to the **geom** layer:

```{r}
ggplot(data=iris,
       mapping=aes(x=Sepal.Length,
                   fill=Species,
                   color=Species))  +
  geom_histogram(alpha=0.2)
```

Sometimes it's really important to include the aesthetic mapping in the data layer. In this case, however, we could also move the `fill` and `color` parameters to the **geom** layer. Since it's still a mapping of data to an aesthetic, we still need to wrap it inside an `aes` clause:

```{r}
ggplot(data=iris,
       mapping=aes(x=Sepal.Length))  +
  geom_histogram(aes(fill=Species, color=Species),
                 alpha=0.2)
```

There is an extension of ggplot2 that makes histograms look a little nicer (I'll find this and add to this worksheet later).


# Bar plots

Bar plots are among the most common plots and are useful for comparing counts among individuals or groups.

## `barplot()` in base R

Base R has a built-in function `barplot()`.

+ The first argument to `barplot()` is `height`, which is expected to be a vector or a matrix.
+ The arrangement of bars is given by the `beside` argument, which is `FALSE` by default.
+ If `height` is a matrix, the default setting makes a **stacked** barplot.
+ Set `beside=TRUE` to make a **side-by-side** barplot.

Again, there are a bunch of parameters that can be used to control other aspects of the plot's appearance.

Let's make a barplot for a toy dataset where we compare gene expression levels for five different *Arabidopsis* genes under different nutrient conditions.

```{r}
# ================================================== #
# create sample data
Control = sample(x=10,
                 size=5) # draw 5 values from a 1:x sequence of integers
Nitrogen = rnorm(n=5,
                 mean=5,
                 sd=2) # draw 5 values from a normal distribution with a given mean and SD
Phospate = runif(n=5,
                 min=1,
                 max=10) # draw 5 values from a uniform distribution with a given range

sampleData = cbind(Control,Nitrogen,Phospate)
rownames(sampleData) = c("GeneA","GeneB","GeneC", "GeneD", "GeneE")
sampleData # take a look

# ================================================== #
# genes in Control condition (a single vector)
barplot(height=Control)
# or
barplot(height=sampleData[,"Control"])

# ================================================== #
# matrix with stacked bars
barplot(height=sampleData)

# ================================================== #
# side-by-side bars in the rainbow palette, horizontal orientation
barplot(height=sampleData,
        beside=TRUE,
        horiz=TRUE,
        col=rainbow(4))
```

Oops! What happened? `rainbow(4)` makes a palette that just recycles four different colors. Since we have 5 genes, and we want the colors to match across groups, we need the right number of colors. 

Let's fix this up, and also add a legend. We will place the legend somewhere on the graph by specifying x- and y-coordinates.

```{r}
barplot(height=sampleData,
        beside=TRUE,
        horiz=TRUE,
        col=rainbow(5)) # use the right number of colors

# add a legend
legend(x=5,
       y=5, # horizontal and vertical position
       legend = rownames(sampleData), 
       fill = rainbow(5))
```

Well, we are getting there. To fix this up so the legend doesn't cover up any of the data, we also need to set the limits of the plot.

```{r}
barplot(height=sampleData,
        beside=TRUE,
        horiz=TRUE,
        xlim = c(0,12),
        col=rainbow(5)) # use the right number of colors

# add a legend with colors
legend(x=9.5,
       y=15, 
       legend = rownames(sampleData), 
       fill = rainbow(5))
```


## Barplots in ggplot2

To make a barplot grouped by **Condition** in ggplot2, we will need to transform `sampleData.df` into a different format. 

Let's review what **tidy** data is first by looking at the
<a href="https://kriscgun.github.io/xdasi-bio-2021/best_practices/tidy_data.html" 

target="_blank">**Tidy Data**</a>
page on the course website (see "Best Practices=>Tidy Data").

Right now the data are in what's balled "wide" format. Each of the columns in the dataset actually represent **variables** of a global attribute, **Condition**.

To plot the numerical data grouped by both the **Gene** and **Condition** attributes, we need to put each of these three "dimensions" into its own **column**. 

This means we need to **transform** our  data from a ***wide*** format to a ***long*** format (an example of this is illustrated on the **Tidy Data** page).

To do this we can use the `stack()` command. There are other commands to do this in the `tidyverse` package, which we will learn about later.

```{r}
# ================================================== #
# transform wide to long format
sampleData.df.long = stack(data.frame(sampleData))
head(sampleData.df.long) # check

# ================================================== #
# Let's rename "ind" to "Condition"
colnames(sampleData.df.long)[2] = "Condition"

# ================================================== #
# Let's also add a "Gene" column
# Since there are 5 genes and 3 conditions, we will just replicate the 5 gene names 3 times.
sampleData.df.long$Gene = rep(rownames(sampleData),3)
head(sampleData.df.long) # check
```

Now we can make our stacked and side-by-side bar plots.

```{r, out.width = "60%"}
# ================================================== #
# stacked
ggplot(data=sampleData.df.long,
       aes(x=Condition,
           y=values,
           fill=Gene)) +
  # geom_bar(stat="identity") # this works, but ...
    geom_col()                # this is simpler
  
# ================================================== #
# side-by-side
ggplot(data=sampleData.df.long,
       aes(x=Condition,
           y=values,
           fill=Gene)) +

  # geom_bar(stat="identity",
  #          color="black",
  #          position=position_dodge())
  
geom_col(position = "dodge",  # still simpler than geom_bar()
         color="black")
```


Next time, we will look at boxplots, violin plots, faceting, and placing multiple graphs on the same plot in the style of a figure in a publication.
