---
title: "Linear regression exercise"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

A dataset called ``trees`` contains information about ``Girth`` ( diameter of the tree ), ``Height`` and ``Volume`` of 31 cherry trees. Let's determine which variable is a better predictor of Volume.

- 1) create linear model volume ~ height

```{r}
# load dataset
trees

# create linear model -- the two expressions below are equivalent
# tree_lm = lm(trees$Volume ~ trees$Height)
tree_lm = lm(Volume ~ Height, data = trees)
summary(tree_lm)
```

- 2) inspect plots and determine if lm is of good quality
    - Is there a transformation that could help the model?
    - Are there any outliers?
    - Is the variance/residual equal around the model?

```{r}
# plot height by volume and the regression line
plot(trees$Height, trees$Volume)
lines(trees$Height, predict.lm(tree_lm))

# look at various plots for the model
plot(tree_lm)
```


- 3) Do the same for Girth. 
    - Is there a transformation that could help the model?
    - Are there any outliers?
    - Is the variance/residual equal around the model?

```{r}
# create linear model -- the two expressions below are equivalent
# tree_lm = lm(trees$Volume ~ trees$Height)
tree_lm = lm(Volume ~ Girth, data = trees)
summary(tree_lm)
```

```{r}
# plot height by volume and the regression line
plot(trees$Girth, trees$Volume)
lines(trees$Girth, predict.lm(tree_lm))

# look at various plots for the model
plot(tree_lm)
```


```{r}
## ============================================================= ##
# try a log transformation of Height
tree_lm_sqr = lm(sqrt(Volume) ~ Girth, data = trees)
summary(tree_lm_sqr)
plot(tree_lm_sqr)
plot(trees$Girth, sqrt(trees$Volume))
lines(trees$Girth, predict.lm(tree_lm_sqr))
```

- 4) Pick a range of values (0:25) and then use the best linear model for Girth to predict the Volume, and the lower and upper intervals.  

```{r}
range_predict = seq(from=0, to=25, by=1)

tree_lm_sqr_predict = predict.lm(tree_lm_sqr, newdata = data.frame(Girth=range_predict), interval="confidence")

```

- 5) use ggplot2 to plot the predicted values and the lower and upper intervals for the values 0:25. 

```{r}
library(ggplot2)
library(reshape2)

tree_predict.melt = melt(tree_lm_sqr_predict)

ggplot(data = tree_predict.melt, 
       aes(x=Var1, y=value, group=Var2)) + geom_line(aes(color=Var2))

```


