---
title: "Linear regression exercise"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

A dataset called ``trees`` contains information about ``Girth`` ( diameter of the tree ), ``Height`` and ``Volume`` of 31 cherry trees. Let's determine which variable is a better predictor of Volume.

- 1) create linear model volume ~ height

```{r}


```

- 2) inspect plots and determine if lm is of good quality
    - Is there a transformation that could help the model?
    - Are there any outliers?
    - Is the variance/residual equal around the model?

```{r}

```


- 3) Do the same for Girth. 
    - Is there a transformation that could help the model?
    - Are there any outliers?
    - Is the variance/residual equal around the model?

```{r}

```

```{r}


```


```{r}

```

- 4) Pick a range of values (0:25) and then use the best linear model for Girth to predict the Volume, and the lower and upper intervals.  

```{r}

```

- 5) use ggplot2 to plot the predicted values and the lower and upper intervals for the values 0:25. 

```{r}

```


