---
title: "Distributions in-class exercise"
subtitle: "XDASI Fall 2021"
author: "YOUR NAMES HERE"
date: "11/1/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Scenario

You want to clone a gene and you have some reporter for success of transformation in *E. coli*. Positive colonies are blue because they express $\beta$-galactosidase when plated on medium with IPTG. The probability of a successful transformation is $0 \le P(success) \le 1$. For the following exercise, you may consider one plate, many plates, or a truly gargantuan number of plates.

## Exercise

The class has been split into groups and each has taken one or two of the following distributions:

- Bernoulli
- Binomial
- Hypergeometric
- Geometric
- Negative binomial
- Poisson
- Normal
- Exponential

For your assigned distribution, answer the following questions.

#### a. What is your function? Describe it in terms of the kind of problem that it applies to. Is it discrete or continuous?

<!-- Insert your answer below. -->


#### b. Describe a question that you can ask with your distribution as applied to the transformation scenario given above.

<!-- Insert your answer below. -->


#### c. Describe the parameterization of the equation. What parameters does it take, and what do they mean?

<!-- Insert your answer below. -->


#### d. Pick some appropriate parameters for your distribution as applied to the transformation scenario and graph the PDF and CDF.

```{r}
```


#### e. Compute a cumulative probability for some appropriate value and describe in words what it means.

```{r}
```

<!-- Insert your answer below. -->


#### f. Identify at least one other distribution that yours relates to and be prepared to discuss the relationship between them. This could be a special case, a limiting case, an approximation, or an inverse CDF (survival) relationship.

<!-- Insert your answer below. -->
