---
title: "covid"
author: "Raza"
date: "9/03/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,eval = TRUE, warning = FALSE)
```

Load required packages
```{r}


```

Import the coronavirus dataset into R as a dataframe called 'covid' (hint:use the `read.csv` command) 
```{r}


```

Look at the first few entries in the dataset to check that the data has been imported correctly (hint:use the `head` command) 
```{r}


```

Check which format of data each column in the dataframe stores (hint:use `str` command)
```{r}


``` 

The dataset contains multiple columns such as the longitude, latitude provinces etc. However, we are only interested in the date,country,type and cases columns. Subset the data so that only these four columns are retained in the covid dataframe  
```{r}


```

We first want to focus on United States data only. Filter the dataset to retain rows where the "country" column contains the text "US" and store in a new variable called "usa" (hint: retain rows where covid$country=="US")
```{r}


```

Further filter the dataframe 'usa' and only retain days (rows) on which confirmed cases were recorded. Store in a variable called us_confirmed (hint:usa$type=="confirmed")
```{r}


```

Sum the "cases" column of the us_confirmed dataframe  to calculate the total of no of confirmed cases recorded in the US (hint:use the `sum()` function on the cases column)
```{r}


```

Calculate the mean of the "cases" column to calculate the avg number of confirmed cases recorded per day in the US  (hint:use the `mean` function on the cases column)
```{r}


```

We could also have calculated the total no of cases in the US by categories (confirmed, recovered, death) with the apply function and the dataframe usa (hint: use `tapply` and `sum` function)

```{r}


```

What was the highest number of confirmed cases recorded in the US on any single day since the study started (hint: use the `max` function on the cases column) 
```{r}


```

Find the index (row number) of the row containing the highest number of cases (hint:use the `which` command: which(us_confirmed$cases==???))
```{r}


```

Use this index to determine the date on which the highest number of cases were recorded
```{r}


```

We could also have used a single command to find the date on which max cases occured as follows: `us_confirmed[us_confirmed$cases==max(us_confirmed$cases),"date"]`

Modify this command to find the dates on which the number of confirmed cases recorded was higher than the daily average rate (hint:use `mean` instead of `max`) 
```{r}


```

Lets assume that 20% of confirmed cases usually exhibit covid19 symptoms. Add a new column to the us_confirmed dataframe showing the estimated number of symptomatic cases for each day. Name this column "symptomatic" and check that the column has been added correctly with the `head` command. (hint:this new column should contain the value us_confirmed$cases*0.2 )
```{r}


```

**dplyr**

We can usually manipulate data frames using the dplyr package more easily than using base R. Reimport the data and subset to retain the date,country,type and cases column using the dplyr package (hint:use the `select` command)   
```{r}


```

Filter the covid dataframe to retain rows where country is US (hint:use the `filter` command)   
```{r}


```

Filter the usa dataframe to retain rows where type is "confirmed" (hint:use the `filter` command)   
```{r}


```

Instead of first filtering by country and then by type, filter the covid dataset by both variables at the same time to obtain the us_confirmed dataset (hint:use the `filter` command)   
```{r}


```

DPLYR uses the `mutate` command to add new columns to a dataframe. Add the a column called "symptomatic" to the us_confirmed dataframe. As before, this column should contain the value of the cases column multiplied by 0.2  
```{r}


```

DPLYR uses the pipe operator `%>%`  (ctrl-shift-m is a shortcut to type the pipe operator in Rstudio) to pass the ouput of one command directly into another without creating an intermediate variable. Filter the 'covid' dataframe by country and type as before and directly pipe the output into the mutate function for adding the symptomatic column. Store the output in a variable called "pipe_data"   
```{r}


```

We can use the `summarise` function from dplyr to apply specific functions like mean, median, sum to data of interest. Filter the covid dataset to only retain confirmed cases and pipe into the `summarise` function to calculate the avg no of confirmed cases recorded (hint:covid %>% filter(???) %>% summarise(avg=???). 
```{r}


```

We can combine the `summarise` command with the `group_by` command to apply a function to specific groups of categorical variables. Find the avg no of confirmed cases for each country by filtering to retain confirmed cases, and piping the output into the `groupby` and `summarise` commands (hint:use group_by(country))  
```{r}


```
