---
title: "Exercise: RR, OR, Fisher's Exact Test"
subtitle: "XDASI Fall 2021"
date: "10/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
```

## Is high blood pressure a risk factor for stroke?

Consider two different experimental designs for determining whether high blood pressure is a risk factor for strokes:

+ **Design 1** 
  + 2000 men between the ages of 70-80 are chosen at random from the population.
  + These are divided into groups with and without a history of high blood pressure.
  + The number of men in each group who experienced a stroke sometime within the following 5 years is recorded.
 
+ **Design 2**
  + 2000 men between the ages of 70-80 are recruited for a study, of which 1000 have never had a stroke and 1000 have had a stroke in the last 5 years.
  + The proportion of each group who have a history of high blood pressure is recorded.

The data might look something like this:

**Design 1**
```{r, echo=F}
## Design 1 ========================= #
Stroke    = c(100,50)   # total =  200
NoStroke = c(600,1250)  # total = 1800
data_matrix1 = rbind(Stroke, NoStroke)
colnames(data_matrix1) = c("HiBP","LoBP")

# kable is a table generator from the knitr package
kable(data_matrix1)
```

**Design 2**
```{r, echo=F}
## Design 2 ========================= #
Stroke   = c(200,800)  # total =  1000
NoStroke = c(50,950)   # total = 1000
data_matrix2 = rbind(Stroke, NoStroke)
colnames(data_matrix2) = c("HiBP","LoBP")
kable(data_matrix2)
```


## 1) Compute RR and OR manually

Let's compute both $RR = \frac{p_{Stroke|HiBP}}{p_{Stroke|LoBP}}$ and 
$OR = \frac{p_{stroke|HiBP}/(1-p_{Stroke|HiBP})}{p_{Stroke|LoBP}/(1-p_{Stroke|LoBP}}$ for both scenarios:

```{r, collapse=T}
## Design 1 ================= #



## Design 2 ================= #



```

## 2) Use R commands to compute RR and OR

We can use the `riskratio()` and `oddsratio()` functions from the "epitools" package to compute the risk ratio automatically.

***Note:*** Look at the documentation for this command. It expects a table in a different format than what we have here! We can specify to change this by setting the `rev` parameter. We also need to transpose the table!!!

```{r}
library(epitools)


```


## 3) How do the RR and OR compare for these two scenarios?

```{r eval=FALSE}
# your answer here
```

## 4) Use Fisher's Exact Test 

Now use Fisher's test to get a p-value and OR estimate for these data. How do the results compare with your results above?

```{r}

```

