---
title: "Homework 10: Hypothesis Testing"
subtitle: "XDASI Fall 2021"
author: "YOUR NAME HERE"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
```

## COVID-19 cases in NYC

"NYC_covid_cases_by_age.csv" contains weekly data of cases in NYC over the course of the pandemic. 

### 1) Import and tidy the data

Convert the data to long format.

  + You should preserve the `week_ending` column, but get rid of the `age_all_ages` column and convert the rest of the columns to two new columns containing the `age_group` and `cases` information.
  + Turn the `age_groups` column into a factor.
  + Make sure the `week_ending` column is properly formatted as a Date type.
  
```{r}
# import file
nyc_cases<- read.csv("NYC_covid_cases_by_age.csv")
# convert to long format
nyc_cases_long <- nyc_cases %>% gather("age_group", "cases", -week_ending)
# format age_group as factor and week_ending as date
nyc_cases_long$age_group<- as.factor(nyc_cases_long$age_group)
nyc_cases_long$week_ending <- as.Date(nyc_cases_long$week_ending, "%m/%d/%y")
```


### 2) Plot cases over time by age group

Create a plot that shows the trajectory of cases over the course of the pandemic for each of the age groups and for all ages combined.

```{r}
cases_weekly_plot<- ggplot(nyc_cases_long,
                           aes(x=week_ending,
                               y = cases,
                               color= age_group)) +
  geom_line() +
  theme(panel.background=element_blank(), 
        strip.background = element_blank(),
        strip.placement = "outside",
        strip.text=element_blank(),
        panel.border=element_rect(colour="black",fill=NA),
        legend.title = element_blank())  

cases_weekly_plot
```


### 3) Significance testing

How would you determine whether there was a statistically significant difference in total cases among people aged 25-34 vs. those aged 75 and over?  

a) Explore the data visually using a boxplot, histograms, and qq plots, and perform a test for normality.

```{r}
# how are the data in both age categories distributed - boxplot
boxplot(nyc_cases$age_25_34, nyc_cases$age_75up)

# how are the data in both age categories distributed - histograms and qqnorm plots
par(mfrow = c(2,2))
hist(nyc_cases$age_25_34,
     breaks = 10)
hist(nyc_cases$age_75up,
     breaks = 10)

qqnorm(nyc_cases$age_25_34)
qqnorm(nyc_cases$age_75up)

# do the data in each age category pass the normality test?
shapiro.test(nyc_cases$age_25_34)  
shapiro.test(nyc_cases$age_75up)
```


b) If the data look highly skewed, it might be useful to normalize the data. Do this by converting the total number of cases in each group to a proportion of total cases in all age groups that week. Then, replot and retest for normality.

```{r}
nyc_cases_prop <- nyc_cases_long %>%
  drop_na() %>%
  group_by(week_ending) %>% 
  mutate(prop = cases/sum(cases))

cases_25_34 <- nyc_cases_prop %>%
  filter(age_group == "age_25_34") %>%
  pull(prop)

cases_75up <- nyc_cases_prop %>%
  filter(age_group == "age_75up") %>%
  pull(prop)

# how are the data in both age categories distributed - boxplot
boxplot(cases_25_34, cases_75up)

# how are the data in both age categories distributed - histograms and qqnorm plots
par(mfrow=c(2,2))

hist(cases_25_34)
qqnorm(cases_25_34)
qqline(cases_25_34)

hist(cases_75up)
qqnorm(cases_75up)
qqline(cases_75up)

# do the data in each age category pass the normality test?
shapiro.test(cases_25_34)
shapiro.test(cases_75up)
```

c) Did this help at all? Based on these results, run the appropriate statistical test to compare the proportional distributions of cases in these two groups.

```{r}
wilcox.test(cases_25_34, cases_75up)
```

### 4) Delta strain

Is there a significant difference between the proportion of infections among 25-34 year-olds before and after the rise of the Delta strain? (The Delta strain appeared 6/21/21.)

a) First, check for normality and choose the appropriate test.

```{r}
# filter the data
cases_25_34<- cases_25_34%>%
  mutate(Delta= ifelse(week_ending < "0021-06-26", "Before_Delta", "After_Delta"))
cases_25_34$Delta = factor(cases_25_34$Delta,
                           levels=c("Before_Delta","After_Delta"))

cases_25_34_before <- cases_25_34 %>%
  filter(Delta == "Before_Delta")

cases_25_34_after <- cases_25_34 %>%
  filter(Delta == "After_Delta")

# exploratory plots
par(mfrow = c(2,2))
hist(cases_25_34_before$prop)
qqnorm(cases_25_34_before$prop)
qqline(cases_25_34_before$prop)

hist(cases_25_34_after$prop)
qqnorm(cases_25_34_after$prop)
qqline(cases_25_34_after$prop)

shapiro.test(cases_25_34_before$prop) 
shapiro.test(cases_25_34_after$prop) 

# statistical test
wilcox.test(cases_25_34_before$prop, cases_25_34_after$prop)

```

b) Make a violin plot of the proportion of cases in the 25-34-year old group before and after the appearance of the Delta variant.

```{r}

# before and after delta - line plot
# cases_before_after_delta_plot<- ggplot(cases_25_34, aes(x=prop, color= Delta,
#                                                         group= Delta))+
#   geom_density() +
#   theme(panel.background=element_blank(), 
#         strip.background = element_blank(),
#         strip.placement = "outside",
#         strip.text=element_blank(),
#         panel.border=element_rect(colour="black",fill=NA),
#         legend.title = element_blank())  
# cases_before_after_delta_plot

# violin plot
ggplot(cases_25_34, aes(x= Delta, y=prop, fill= Delta)) +
  geom_violin() +
  geom_boxplot(width=0.1, fill="white") +
  theme(panel.background=element_blank(), 
        strip.background = element_blank(),
        strip.placement = "outside",
        strip.text=element_blank(),
        panel.border=element_rect(colour="black",fill=NA),
        legend.title = element_blank())  
```


### 5) Breakthrough cases

Consider the following data of the number of NYC documented Covid cases, hospitalizations, and deaths among vaccinated and unvaccinated individuals:

+ Cases:
  + Vaccinated: 820
  + Unvaccinated: 4140  

+ Hospitalizations:
  + Vaccinated: 32
  + Unvaccinated: 349

+ Deaths:
  + Vaccinated: 4
  + Unvaccinated: 39


What is an appropriate test you could use to analyze these data? Explain.

```{r eval=FALSE}
# your answer here
Either a Fishers Exact Test or a Chi-square test would work here.

Since the numbers are not so big, it is feasible to perform Fishers 
and not too computationally intensive. (People tend to forget that Fishers can
be performed on tables that are larger than two-by-two, but it works fine when
the numbers are not too big).
```


Use this data to create a contingency table, and then perform the appropriate test to determine whether there is a significant difference between the vaccinated and unvaccinated groups.

```{r}
breakthrough_cases = data.frame(Cases = c(820,4140),
                                Hospitalizations = c(32, 349),
                                Deaths = c(4, 39))
row.names(breakthrough_cases) = c("Vaccinated","Unvaccinated")
breakthrough_cases

knitr::kable(breakthrough_cases)

fisher.test(breakthrough_cases)
chisq.test(breakthrough_cases)
```

The tests work the same either way the table is constructed. By convention, we put the groups we want to compare in the rows with the "focal" group first, and the conditions in the columns.

```{r eval=FALSE}
vaccinated<- c(820,32,4)
unvaccinated<- c(4140,349,39)

breakthrough<- data.frame(vaccinated, unvaccinated)
rownames(breakthrough)<- c("cases", "hospitalizations", "deaths")

fisher.test(breakthrough)
chisq.test(breakthrough)
```








