---
title: "Homework 11: ANOVA"
subtitle: "XDASI Fall 2021"
author: "Wells Burrell"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
```


# Q0: Load the data set and examine it

For this exercise, we will be looking at brood size and fertilization defects in a *C. elegans* gene called *mel-28*, alone and in combination with a mutation in a second gene that suppresses these phenotypes. There are 30 biological replicates for each condition. The measurements taken were:  

+ **Brood size** - number of eggs laid by one worm in one day  
+ **Fertilized** - number of eggs laid that were fertilized embryos

First, load the data and take a look at it. Make sure that the `genotype` column is a factor and that you set "WT" as the reference level.

*Hint: you can use either the `factor()` or `relevel()` commands to set/reorder the levels; what weird thing happens if you use `level()` instead?*

```{r}
# load the data
elegens.data = read.csv("Ce_mel28_Emb_Ste.csv", stringsAsFactors = T)
# take a look at it (check / fix factor levels)
head(elegens.data)
```


# Q1: Brood size

For this first part, we will be looking just at the ***brood sizes*** in the wild-type N2 strain (WT), in single mutants (A), and in double mutants (A;B).

### a) Strip chart / boxplot

Plot the brood size data, including the individual data points. For readability, make sure that the WT is the leftmost boxplot, followed by the single mutant and then the double mutant.

```{r}

elegens.data$genotype = factor(elegens.data$genotype, levels = c("WT","A","A;B"))

ggplot(data = elegens.data,
       mapping = aes(x = genotype,
                     y = brood_size)) +
  geom_boxplot() +
  geom_jitter(height = 0)

```

### b) Histograms

Use histogram plots to visualize the distributions. and perform an appropriate statistical test to determine if the measurements from the different groups are normally distributed.  

```{r}

elegens.data.WT = elegens.data %>% filter(genotype == "WT")
elegens.data.A = elegens.data %>% filter(genotype == "A")
elegens.data.AB = elegens.data %>% filter(genotype == "A;B")

ggplot(data = elegens.data.WT,
       mapping = aes(x = brood_size)) +
  geom_histogram(bins = 10)

ggplot(data = elegens.data.A,
       mapping = aes(x = brood_size)) +
  geom_histogram(bins = 10)

ggplot(data = elegens.data.AB,
       mapping = aes(x = brood_size)) +
  geom_histogram(bins = 10)

shapiro.test(elegens.data.WT$brood_size)
shapiro.test(elegens.data.A$brood_size)
shapiro.test(elegens.data.AB$brood_size)
```

### c) ANOVA assumptions

In order to proceed with ANOVA, what two requirements should the data fulfill?

```{r eval=FALSE}
# each group sample must come from a normally distributed population, and groups must have common variances.
```

### d) Check for normality

Perform statistical tests to determine whether the data are normally distributed. (You may find it easier to subset the data at this point.)

```{r}
shapiro.test(elegens.data.WT$brood_size)
shapiro.test(elegens.data.A$brood_size)
shapiro.test(elegens.data.AB$brood_size)
```


### e) Check for homoscedasticity

Determine if the variation is the same between the groups. This is called ***"homoscedasticity"***.

+ First, use the `var.test()` function to make pairwise comparisons between groups.
+ Then, use `bartlett.test()` to perform Bartlett's test on all of the groups at once.

```{r}
# var.test
var.test(elegens.data.WT$brood_size, elegens.data.A$brood_size)
var.test(elegens.data.WT$brood_size, elegens.data.AB$brood_size)
var.test(elegens.data.A$brood_size, elegens.data.AB$brood_size)
# bartlett.test
bartlett.test(elegens.data$brood_size ~ elegens.data$genotype)
```


What is the test statistic used by `var.test()`? What is the value of the statistic when the null hypothesis is true?

```{r eval=FALSE}
# The test statistic for the var.test function is the f-statistic. When the null is true, we expect this value to be 1.
```


What distribution does Bartlett's test statistic follow?

```{r eval=FALSE}
# Chi-squared distribution
```


Are the assumptions for performing ANOVA met for this dataset?

```{r eval=FALSE}
# Yes but just barely. Data are normal enough but variances are a bit different. 
```


# Q2: Sum of Squares and F-statistic  

### a) SSE and SSG

Calculate the sum of squares for $SS_{error}$ and $SS_{group}$ (you may use either base R methods or dplyr, as you prefer). Recall the formulas:

$$SS_{total} = SS_{error}+SS_{group}$$ 
$$SS_{error} = \sum_{i}\sum_{j}(Y−\overline{Y}_i)^{2}$$    
$$SS_{group} = \sum_{i}n_{i}(\overline{Y}_i−\overline{Y})^{2}$$
```{r}
# SS_error
ss.error = sum((elegens.data.WT$brood_size - mean(elegens.data.WT$brood_size))^2,
    (elegens.data.A$brood_size - mean(elegens.data.A$brood_size))^2,
    (elegens.data.AB$brood_size - mean(elegens.data.AB$brood_size))^2)

## SS_group
mean.WT = mean(elegens.data.WT$brood_size)
mean.A = mean(elegens.data.A$brood_size)
mean.AB = mean(elegens.data.AB$brood_size)
grand.mean = mean(elegens.data$brood_size)

ss.groups = sum(length(elegens.data.A$brood_size) *
             (c(mean.WT,mean.A,mean.AB) - grand.mean)^2)

```

### b) F-statistic and p-value

Calculate the F-statistic and the p-value using the `pf() function.` Recall that:

$$F=\frac{MS_{groups}}{MS_{error}} = \frac{SS_{groups}/df_{groups}}{SS_{error}/df_{error}}$$

```{r}
# degrees of freedom for SS_error
df.ss.error = length(elegens.data$brood_size) - 3
# degrees of freedom for SS_groups
df.ss.groups = 3 - 1
# F-statistic
f.stat = (ss.groups / df.ss.groups) / (ss.error / df.ss.error)
# p-value using CDF for F-statistic
pf(f.stat, df.ss.groups, df.ss.error, lower.tail = F)
```


# Q3: Planned and Unplanned ANOVA

### a) Planned vs. Unplanned

Explain whether this experiment should be treated as a PLANNED or an UNPLANNED dataset. How would the analysis differ between these two scenarios?

```{r eval=FALSE}
# This experiment shoudl be treated as a planned dataset since it has clearly defined experimental groups adn control. In the case of a planned experiment, ANOVA is appropriate, otherwise we would need to use the aov() command. 
```


### b) Planned experiment

Assume the data is a planned experiment. Use the `lm()` function to perform the analysis. 

First, use the `anova()` function on the results of of the `lm()` function to look at the overall result. *Note: you can use the special notation `$"Pr(>F)" to get the exact p-value from the `anova()` function.*

```{r}
# anova for lm of brood_size ~ genotype
anova(lm(brood_size ~ genotype, data = elegens.data))
anova(lm(brood_size ~ genotype, data = elegens.data))$"Pr(>F)"
```

How do the $SS_error$, $SS_group$, $MS_error$, $MS_group$, $Fstat$, and $p$-value compare to the ones you calculated by  hand?

```{r eval=FALSE}
# They are equivalent
These should be exactly the same.
```

Now, use the `summary()` function on the results of the the linear model.

```{r}
summary(lm(brood_size ~ genotype, data = elegens.data))
```

What do the different parts of the `summary()` output tell you? What is R-squared and what does it mean?

```{r eval=FALSE}
# The summary() output tells me the distribution of my residuals, and the coefficients for the relationship between each genotype and brood size variable and their errors/p values. The R-squared value describes the overall strength and direction of the correlation genotype and brood size.
```

Is the model significant overall? How about the different groups: are they significantly different from the control?

```{r eval=FALSE}
# The model is significant and each group is significantly different from control, suggesting this is a good model to describe the relationship. 
```


### c) Unplanned experiment

Now assume the data are from an unplanned experiment. Use the `aov()` function to perform the analysis, and then use the `tukeyHSD()` function on the results of `aov()`.

```{r}
aov(lm(brood_size ~ genotype, data = elegens.data))
```

Use the `tukeyHSD()` function on the results of `aov()` to perform the Tukey-Kramer test to look at all pairwise comparisons.

```{r}
TukeyHSD(aov(lm(brood_size ~ genotype, data = elegens.data)))
```

What does Tukey's HSD test do? What do these results tell you? What is the biological interpretation of these results?

```{r eval=FALSE}
# The Tukey's HSD test calculates the difference in means between between all groups and creates confidence intervals for these values. The results tell me that each group is significantly different from one another. The biological interpretation is that the three genotypes tested (including control WT) all have significantly different brood sizes from one another. From this test we can also see that the greatest difference is between WT and genotype A.
```


### d) Residuals

Once you have performed an ANOVA, you can also check your assumption of normality by checking to see if the ANOVA residuals are normally distributed. 

+ Do this by first assigning the results of your `aov()` function to a variable (e.g. `res_aov`.
+ Then, make a histogram of the residuals, which can be found in `res_aov$residuals`.
+ Finally, perform a test for normality on the residuals.

```{r}
res.aov = aov(lm(brood_size ~ genotype, data = elegens.data))
hist(res.aov$residuals)
shapiro.test(res.aov$residuals)
```


### e) ANOVA for unequal variances

If the data look normal, but the variances are unequal, what kind of ANOVA could you do instead? *Hint: what kind of $t$-test would you use in this case?*

```{r eval=FALSE}
# You could doa Kruskal-Wallace test in this case. 
```

Perform this test below. Also extract the exact $p$-value from the test.

```{r}
kruskal.test(brood_size ~ genotype, data = elegens.data)
kruskal.test(brood_size ~ genotype, data = elegens.data)$p.value
```

How does the $p$-value compare with the ANOVA you performed above? Explain.

```{r eval=FALSE}
# The p-value is still significant, but less so using the Kruskal-Wallace test. This is because you sacrifice power when you have fewer assumptions in your statistical test. 
```


# Q4: Fertilization phenotype

Now we will look at the fertilization phenotype. 

### a) Percent fertilization

+ Add a column to your original data frame containing the ***percent fertilization***.
+ Take a look at the percent fertilization using a stip chart / boxplot (as you did for Q1).

```{r}
# add percent fertilization column
elegens.data$percent.fert = elegens.data$fertilized / elegens.data$brood_size * 100
# plot it
ggplot(data = elegens.data,
       mapping = aes(x = genotype,
                     y = percent.fert)) +
  geom_boxplot() +
  geom_jitter(height = 0)
```


### b) Check assumptions for ANOVA

Perform the appropriate tests to check whether you can perform an ANOVA on these data and draw histograms to visualize the distributions. (You may again find it useful at this point to subset the data by group.)

```{r}
# test for assumptions
elegens.data.WT = elegens.data %>% filter(genotype == "WT")
elegens.data.A = elegens.data %>% filter(genotype == "A")
elegens.data.AB = elegens.data %>% filter(genotype == "A;B")

shapiro.test(elegens.data.WT$percent.fert)
shapiro.test(elegens.data.A$percent.fert)
shapiro.test(elegens.data.AB$percent.fert)

var.test(elegens.data.WT$percent.fert, elegens.data.A$percent.fert)
var.test(elegens.data.A$percent.fert, elegens.data.AB$percent.fert)
var.test(elegens.data.AB$percent.fert, elegens.data.WT$percent.fert)

# histograms
hist(elegens.data.WT$percent.fert)
hist(elegens.data.A$percent.fert)
hist(elegens.data.AB$percent.fert)
```

Do these data meet the assumptions required for ANOVA? If not, what alternative tests could you perform? Explain the conditions under which either test would be appropriate.

```{r eval=FALSE}
# They do not meet the assumptions. An alternative test would be the Kruskal-Wallace test. For ANOVA, the variances must be equal, and the data must be normally distributed. In this case, neither hold true.
```

### c) Statistical test

Perform an appropriate alternative to ANOVA for these data.

```{r}
kruskal.test(percent.fert ~ genotype, data = elegens.data)
```
 
 
# Q5: Error, Power, and Sample size

Recall that the power is $1−\beta$, where $\beta$ is the Type II error. We can calculate the power of the ANOVA test using the command `power.anova.test()`, which is contained in the base `stats` package. The arguments for the command are:

+ **groups** = the number of groups in the dataset  
+ **n** = the number of observations per group  
+ **between.var** = between-group variance, $SS_{group}$
+ **within.var** = within-group variance, $SS_{error}$  
+ **sig.level** = significance cutoff $\alpha$ (Type I error)  
+ **power** = power of the test ($1−\beta = 1$ - Type II error)  

Given all of the arguments except for one, this function will return a value for the missing argument.

### a) Power of ANOVA for this dataset

Given the data that we have been provided and the $p$-value we obtained for the `brood_size`, what is the power of our dataset, given a Type I error rate (significance cutoff) of $\alpha=0.05$?

```{r}
# first, check the documentation
help(power.anova.test)

# compute the power
power.anova.test(groups= 3, 
                 n= 30, 
                 within.var= ss.error, 
                 between.var= ss.groups, 
                 power=NULL,
                 sig.level= 0.05)
```

### b) Sample size

If we would like to achieve a power of 95% with a Type I error rate of $\alpha=0.05$, how many observations for each group would we need? 

```{r}
# compute the required sample size
power.anova.test(groups= 3, 
                 n= NULL, 
                 within.var= ss.error, 
                 between.var= ss.groups, 
                 power=0.95,
                 sig.level= 0.05)

# sample size needed = 3

```

How do you feel about this result?

```{r eval=FALSE}
# This is a great result for the experimenter since many replicates can slow down progress. The small sample size needed is the result of such a great effect size, however, many of the effects we are interested in as biologists are mush smaller than the one observed here. 
```

### c) Effect size

The ratio of the between-groups variation to the within-groups variation ($SS_{group}/SS_{error}$) gives us an approximate effect size.

The `brood_size` data here has a pretty big effect size, and we have a lot of power because of this. 

If the effect size were 10 times smaller, how many animals would we need to get a power of 95% at a Type I error threshold of 5%?

```{r}
power.anova.test(groups= 3, 
                 n= NULL, 
                 within.var= ss.error, 
                 between.var= ss.groups / 10, 
                 power=0.95,
                 sig.level= 0.05)

# n = 16
```

<!-- That's it! Yay! Check to see if your file knits and then send this off! -->

<!-- Note: if you don't have LaTeX installed, or you had problems installing it, then the file may not knit. In that case, type `rm(list = ls())` in the console and then try running the entire file again to make sure that all of your final data structures are what you think they are. -->
