---
title: "Homework 11: ANOVA"
subtitle: "XDASI Fall 2021"
author: "Ozan Berk Imir"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
```


# Q0: Load the data set and examine it

For this exercise, we will be looking at brood size and fertilization defects in a *C. elegans* gene called *mel-28*, alone and in combination with a mutation in a second gene that suppresses these phenotypes. There are 30 biological replicates for each condition. The measurements taken were:  

+ **Brood size** - number of eggs laid by one worm in one day  
+ **Fertilized** - number of eggs laid that were fertilized embryos

First, load the data and take a look at it. Make sure that the `genotype` column is a factor and that you set "WT" as the reference level.

*Hint: you can use either the `factor()` or `relevel()` commands to set/reorder the levels; what weird thing happens if you use `level()` instead?*

```{r}
# load the data
c.elegans.brood = read.csv("Ce_mel28_Emb_Ste.csv", header = TRUE)
# take a look at it (check / fix factor levels)
c.elegans.brood
```


# Q1: Brood size

For this first part, we will be looking just at the ***brood sizes*** in the wild-type N2 strain (WT), in single mutants (A), and in double mutants (A;B).

### a) Strip chart / boxplot

Plot the brood size data, including the individual data points. For readability, make sure that the WT is the leftmost boxplot, followed by the single mutant and then the double mutant.

```{r}
ggplot(data = c.elegans.brood, aes(x = factor(genotype, levels = c("WT", "A", "A;B")), y = brood_size, color = genotype)) + 
  labs(x = "Genotypes", y = "Brood Size") +
  geom_boxplot()
```

### b) Histograms

Use histogram plots to visualize the distributions. and perform an appropriate statistical test to determine if the measurements from the different groups are normally distributed.  

```{r}
# histogram
ggplot(data = c.elegans.brood, aes(x = brood_size, fill = genotype)) + 
  geom_histogram()

#normality
shapiro.test(c.elegans.brood$brood_size)

qqnorm(c.elegans.brood$brood_size)
qqline(c.elegans.brood$brood_size)


```

### c) ANOVA assumptions

In order to proceed with ANOVA, what two requirements should the data fulfill?

```{r eval=FALSE}
# Equal variance and normally distributed data
```

### d) Check for normality

Perform statistical tests to determine whether the data are normally distributed. (You may find it easier to subset the data at this point.)

```{r}
# subset
WT.brood = c.elegans.brood %>% filter(genotype == "WT") %>% summarize(brood_size)
A.brood = c.elegans.brood %>% filter(genotype == "A") %>% summarize(brood_size)
AB.brood = c.elegans.brood %>% filter(genotype == "A;B") %>% summarize(brood_size)

# check for normality
shapiro.test(WT.brood$brood_size) # normally distributed (p-val = 0.4517)
shapiro.test(A.brood$brood_size)  # normally distributed (p-val = 0.1541)
shapiro.test(AB.brood$brood_size) # normally distributed (p-val = 0.6965)
```


### e) Check for homoscedasticity

Determine if the variation is the same between the groups. This is called ***"homoscedasticity"***.

+ First, use the `var.test()` function to make pairwise comparisons between groups.
+ Then, use `bartlett.test()` to perform Bartlett's test on all of the groups at once.

```{r}
# var.test
var.test(WT.brood$brood_size, A.brood$brood_size)
var.test(WT.brood$brood_size, AB.brood$brood_size)
var.test(A.brood$brood_size, AB.brood$brood_size)
# all variations the same

# bartlett.test
bartlett.test(c.elegans.brood$brood_size ~ c.elegans.brood$genotype, c.elegans.brood)
```


What is the test statistic used by `var.test()`? What is the value of the statistic when the null hypothesis is true?

```{r eval=FALSE}
# F statistic is used for var.test() which is the ratio of variances between different data groups. If the null hypothesis is true, then the ratio is close to 1.
```


What distribution does Bartlett's test statistic follow?

```{r eval=FALSE}
# Chi-square distribution
```


Are the assumptions for performing ANOVA met for this dataset?

```{r eval=FALSE}
# Yes, the assumptions that the variances are equal and that the distribution of the groups are normally distributed are both met.
```


# Q2: Sum of Squares and F-statistic  

### a) SSE and SSG

Calculate the sum of squares for $SS_{error}$ and $SS_{group}$ (you may use either base R methods or dplyr, as you prefer). Recall the formulas:

$$SS_{total} = SS_{error}+SS_{group}$$ 
$$SS_{error} = \sum_{i}\sum_{j}(Y−\overline{Y}_i)^{2}$$    
$$SS_{group} = \sum_{i}n_{i}(\overline{Y}_i−\overline{Y})^{2}$$
```{r}
# SS_error
anova_WT = mean(WT.brood$brood_size)
anova_A = mean(A.brood$brood_size)
anova_AB = mean(AB.brood$brood_size)

SS_error = 0
for ( i in 1:nrow(WT.brood)) {
  SS_error = SS_error + sum((WT.brood[i,]-anova_WT)**2)
}
for ( i in 1:nrow(A.brood)) {
  SS_error = SS_error + sum((A.brood[i,]-anova_A)**2)
}
for ( i in 1:nrow(AB.brood)) {
  SS_error = SS_error + sum((AB.brood[i,]-anova_AB)**2)
}
SS_error

## SS_group
brood.mean = mean(c.elegans.brood$brood_size)
SS_group = 0
for ( i in 1:3 ) {
 SS_group = SS_group + (nrow(WT.brood)*(-brood.mean)^2)
}
SS_group
```

### b) F-statistic and p-value

Calculate the F-statistic and the p-value using the `pf() function.` Recall that:

$$F=\frac{MS_{groups}}{MS_{error}} = \frac{SS_{groups}/df_{groups}}{SS_{error}/df_{error}}$$

```{r}
# degrees of freedom for SS_error
df.error = nrow(c.elegans.brood)-1
df.error
# degrees of freedom for SS_groups
df.group = nrow(WT.brood)-1
df.group
# F-statistic
F.stat = ((SS_group)/(df.group))/((SS_error)/(df.error))
F.stat
# p-value using CDF for F-statistic
pf(F.stat, df.error, df.group, lower.tail = F)
```


# Q3: Planned and Unplanned ANOVA

### a) Planned vs. Unplanned

Explain whether this experiment should be treated as a PLANNED or an UNPLANNED dataset. How would the analysis differ between these two scenarios?

```{r eval=FALSE}
# This should be a planned experiment because we have a wild type / control group. If we had an unplanned experiment, we should utilize the Tukey's test in order to analyze, but since it is planned, then we can simply utilize a one-way ANOVA. 
```


### b) Planned experiment

Assume the data is a planned experiment. Use the `lm()` function to perform the analysis. 

First, use the `anova()` function on the results of the `lm()` function to look at the overall result. *Note: you can use the special notation `$"Pr(>F)" to get the exact p-value from the `anova()` function.*

```{r}
# anova for lm of brood_size ~ genotype
anova(lm(formula = c.elegans.brood$brood_size ~ c.elegans.brood$genotype, 
         data = c.elegans.brood))
```

How do the $SS_error$, $SS_group$, $MS_error$, $MS_group$, $Fstat$, and $p$-value compare to the ones you calculated by  hand?

```{r eval=FALSE}
# These should be exactly the same.
```

Now, use the `summary()` function on the results of the linear model.

```{r}
summary(lm(formula = c.elegans.brood$brood_size ~ c.elegans.brood$genotype, 
         data = c.elegans.brood))
```

What do the different parts of the `summary()` output tell you? What is R-squared and what does it mean?

```{r eval=FALSE}
# R-squared is the strength of the variance analysis between the values of 0 and 1, with 1 indicating the highest strength.
```

Is the model significant overall? How about the different groups: are they significantly different from the control?

```{r eval=FALSE}
# The model significance, based on the p value obtained, seems to be significant enough to interpret the results. According to the results of the ANOVA, both the difference between the wild type and A mutant and the wildtype and the AB mutants are both significantly different. 
```


### c) Unplanned experiment

Now assume the data are from an unplanned experiment. Use the `aov()` function to perform the analysis, and then use the `tukeyHSD()` function on the results of `aov()`.

```{r}
aov(lm(formula = c.elegans.brood$brood_size ~ c.elegans.brood$genotype, 
         data = c.elegans.brood))
```

Use the `tukeyHSD()` function on the results of `aov()` to perform the Tukey-Kramer test to look at all pairwise comparisons.

```{r}
TukeyHSD(aov(lm(formula = c.elegans.brood$brood_size ~ c.elegans.brood$genotype, 
         data = c.elegans.brood)))
```

What does Tukey's HSD test do? What do these results tell you? What is the biological interpretation of these results?

```{r eval=FALSE}
# Tukey's HSD test allows a pairwise comparison between each group as there might not be proper establishment of 
```


### d) Residuals

Once you have performed an ANOVA, you can also check your assumption of normality by checking to see if the ANOVA residuals are normally distributed. 

+ Do this by first assigning the results of your `aov()` function to a variable (e.g. `res_aov`.
+ Then, make a histogram of the residuals, which can be found in `res_aov$residuals`.
+ Finally, perform a test for normality on the residuals.

```{r}
res_aov = aov(lm(formula = c.elegans.brood$brood_size ~ c.elegans.brood$genotype, 
         data = c.elegans.brood))
hist(res_aov$residuals)
shapiro.test(res_aov$residuals)
```


### e) ANOVA for unequal variances

If the data look normal, but the variances are unequal, what kind of ANOVA could you do instead? *Hint: what kind of $t$-test would you use in this case?*

```{r eval=FALSE}
# Welch's t-test
```

Perform this test below. Also extract the exact $p$-value from the test.

```{r}
t.test(res_aov$residuals, alternative="less")
t.test(res_aov$residuals, alternative="less")$p.val
```

How does the $p$-value compare with the ANOVA you performed above? Explain.

```{r eval=FALSE}
# The p-value calculated through the residuals are not significant whereas the actual data came out to have a significant difference, which is to be expected as residuals are differences between each value in the dataset compared to the mean of the group they are assigned to, so the differences tend to be more normally distributed and less varied compared to the actual data. 
```


# Q4: Fertilization phenotype

Now we will look at the fertilization phenotype. 

### a) Percent fertilization

+ Add a column to your original data frame containing the ***percent fertilization***.
+ Take a look at the percent fertilization using a stip chart / boxplot (as you did for Q1).

```{r}
# add percent fertilization column
c.elegans.brood$percent_fertile = ((c.elegans.brood$fertilized) / (c.elegans.brood$brood_size)) * 100
# plot it
ggplot(data = c.elegans.brood, aes(x = factor(genotype, levels = c("WT", "A", "A;B")), y = percent_fertile, color = genotype)) + 
  labs(x = "Genotypes", y = "Percent Fertilization") +
  geom_boxplot()
```


### b) Check assumptions for ANOVA

Perform the appropriate tests to check whether you can perform an ANOVA on these data and draw histograms to visualize the distributions. (You may again find it useful at this point to subset the data by group.)

```{r}
# test for assumptions

# subset
WT.brood.percent = c.elegans.brood %>% filter(genotype == "WT") %>% summarize(percent_fertile)
A.brood.percent = c.elegans.brood %>% filter(genotype == "A") %>% summarize(percent_fertile)
AB.brood.percent = c.elegans.brood %>% filter(genotype == "A;B") %>% summarize(percent_fertile)

# check for normality
shapiro.test(WT.brood.percent$percent_fertile) # normally distributed (p-val = 0.4517)
shapiro.test(A.brood.percent$percent_fertile)  # normally distributed (p-val = 0.1541)
shapiro.test(AB.brood.percent$percent_fertile) # normally distributed (p-val = 0.6965)

# check for homoscedasticity
bartlett.test(c.elegans.brood$percent_fertile ~ c.elegans.brood$genotype, c.elegans.brood)

# histograms
ggplot(data = c.elegans.brood, aes(x = percent_fertile, fill = genotype)) + 
  geom_histogram(position = "dodge")

```

Do these data meet the assumptions required for ANOVA? If not, what alternative tests could you perform? Explain the conditions under which either test would be appropriate.

```{r eval=FALSE}
# Because this data is neither normally distributed in any group nor the group variances are equal, we can perform either a Kruskal-Wallace or a Wilcoxon rank test to test for hypothesis testing. In order to prepare the data to the tests, we need to perform corrections to the data in order for there to be assumptions that get met for Kruskal-Wallace and Wilcoxon. Bonferroni correction to control for FWR or Benjamini-Hochberg for FDR controlling can be done prior to our tests. 
```

### c) Statistical test

Perform an appropriate alternative to ANOVA for these data.

```{r}
kruskal.test(c.elegans.brood$percent_fertile ~ c.elegans.brood$genotype, data = c.elegans.brood)
pairwise.wilcox.test(c.elegans.brood$percent_fertile, c.elegans.brood$genotype,
                     p.adjust.method = "BH")
```
 
 
# Q5: Error, Power, and Sample size

Recall that the power is $1−\beta$, where $\beta$ is the Type II error. We can calculate the power of the ANOVA test using the command `power.anova.test()`, which is contained in the base `stats` package. The arguments for the command are:

+ **groups** = the number of groups in the dataset  
+ **n** = the number of observations per group  
+ **between.var** = between-group variance, $SS_{group}$
+ **within.var** = within-group variance, $SS_{error}$  
+ **sig.level** = significance cutoff $\alpha$ (Type I error)  
+ **power** = power of the test ($1−\beta = 1$ - Type II error)  

Given all of the arguments except for one, this function will return a value for the missing argument.

### a) Power of ANOVA for this dataset

Given the data that we have been provided and the $p$-value we obtained for the `brood_size`, what is the power of our dataset, given a Type I error rate (significance cutoff) of $\alpha=0.05$?

```{r}
# first, check the documentation
help(power.anova.test)

# compute the power
power.anova.test(groups=3, 
                 n=30, 
                 within.var=SS_error, 
                 between.var= SS_group, 
                 power=NULL,
                 sig.level=0.05)
```

### b) Sample size

If we would like to achieve a power of 95% with a Type I error rate of $\alpha=0.05$, how many observations for each group would we need? 

```{r}
# compute the required sample size
power.anova.test(groups=3, 
                 n=NULL, 
                 within.var=brood.mean, 
                 between.var= mean(c(anova_A, anova_AB, anova_WT)),
                 sig.level=0.05, power = 0.95)

# sample size needed = 8.815

```

How do you feel about this result?

```{r eval=FALSE}
# I think this is encouraging that with our particular results, even if we have 8 as our sample size, we would still be in good shape to get a power of 95% which means our results are robust.
```

### c) Effect size

The ratio of the between-groups variation to the within-groups variation ($SS_{group}/SS_{error}$) gives us an approximate effect size.

The `brood_size` data here has a pretty big effect size, and we have a lot of power because of this. 

If the effect size were 10 times smaller, how many animals would we need to get a power of 95% at a Type I error threshold of 5%?

```{r}
SS_group/SS_error
# If effect size was 2.6 instead...
# n would have to be 10 times larger almost...
power.anova.test(groups=3, 
                 n=NULL, 
                 within.var=10*brood.mean, 
                 between.var= mean(c(anova_A, anova_AB, anova_WT)),
                 sig.level=0.05, power = 0.95)
# n = 78
```

<!-- That's it! Yay! Check to see if your file knits and then send this off! -->

<!-- Note: if you don't have LaTeX installed, or you had problems installing it, then the file may not knit. In that case, type `rm(list = ls())` in the console and then try running the entire file again to make sure that all of your final data structures are what you think they are. -->
