---
title: "Homework 11: ANOVA"
subtitle: "XDASI Fall 2021"
author: "Eric Laderman"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
```


# Q0: Load the data set and examine it

For this exercise, we will be looking at brood size and fertilization defects in a *C. elegans* gene called *mel-28*, alone and in combination with a mutation in a second gene that suppresses these phenotypes. There are 30 biological replicates for each condition. The measurements taken were:  

+ **Brood size** - number of eggs laid by one worm in one day  
+ **Fertilized** - number of eggs laid that were fertilized embryos

First, load the data and take a look at it. Make sure that the `genotype` column is a factor and that you set "WT" as the reference level.

*Hint: you can use either the `factor()` or `relevel()` commands to set/reorder the levels; what weird thing happens if you use `level()` instead?*

```{r}
# load the data

cel_data = read.csv('/Users/eladerman/NYU/Classes/Biostatistics/Week12/XDASI2021-HW11_ANOVA_files/Ce_mel28_Emb_Ste.csv', header = TRUE)

# take a look at it (check / fix factor levels)
unique(cel_data$genotype)
cel_data$genotype = factor(cel_data$genotype, levels = c('WT', 'A', 'A;B'))
cel_data
```


# Q1: Brood size

For this first part, we will be looking just at the ***brood sizes*** in the wild-type N2 strain (WT), in single mutants (A), and in double mutants (A;B).

### a) Strip chart / boxplot

Plot the brood size data, including the individual data points. For readability, make sure that the WT is the leftmost boxplot, followed by the single mutant and then the double mutant.

```{r}

ggplot(data = cel_data, mapping = aes(x = genotype, y = brood_size)) + geom_boxplot(mapping = aes(fill = genotype)) + geom_jitter(width = 0.25, height = 0)

```

### b) Histograms

Use histogram plots to visualize the distributions. and perform an appropriate statistical test to determine if the measurements from the different groups are normally distributed.  

```{r}

wt_data = cel_data %>% filter(genotype == 'WT')
a_data = cel_data %>% filter(genotype == 'A')
ab_data = cel_data %>% filter(genotype == 'A;B')



hist(wt_data$brood_size)
hist(a_data$brood_size)
hist(ab_data$brood_size)

shapiro.test(wt_data$brood_size)
shapiro.test(a_data$brood_size)
shapiro.test(ab_data$brood_size)


```

### c) ANOVA assumptions

In order to proceed with ANOVA, what two requirements should the data fulfill?

```{r eval=FALSE}
# The distribution of the data must be normal and the variance must be the same across different groups
```

### d) Check for normality

Perform statistical tests to determine whether the data are normally distributed. (You may find it easier to subset the data at this point.)

```{r}
shapiro.test(wt_data$brood_size)
shapiro.test(a_data$brood_size)
shapiro.test(ab_data$brood_size)
```


### e) Check for homoscedasticity

Determine if the variation is the same between the groups. This is called ***"homoscedasticity"***.

+ First, use the `var.test()` function to make pairwise comparisons between groups.
+ Then, use `bartlett.test()` to perform Bartlett's test on all of the groups at once.

```{r}
# var.test

var.test(wt_data$brood_size, a_data$brood_size)
var.test(wt_data$brood_size, ab_data$brood_size)
var.test(a_data$brood_size, ab_data$brood_size)

# bartlett.test

bartlett.test(brood_size ~ genotype, data = cel_data)

```


What is the test statistic used by `var.test()`? What is the value of the statistic when the null hypothesis is true?

```{r eval=FALSE}
# F statistic, the value is 1 if the null hypothesis is true
```


What distribution does Bartlett's test statistic follow?

```{r eval=FALSE}
# The test statistic follows an approximately chisquare distribution
```


Are the assumptions for performing ANOVA met for this dataset?

```{r eval=FALSE}
# Yes as the p values for all of the above tests are greater than 0.05
```


# Q2: Sum of Squares and F-statistic  

### a) SSE and SSG

Calculate the sum of squares for $SS_{error}$ and $SS_{group}$ (you may use either base R methods or dplyr, as you prefer). Recall the formulas:

$$SS_{total} = SS_{error}+SS_{group}$$ 
$$SS_{error} = \sum_{i}\sum_{j}(Y−\overline{Y}_i)^{2}$$    
$$SS_{group} = \sum_{i}n_{i}(\overline{Y}_i−\overline{Y})^{2}$$
```{r}
grand_mean = mean(cel_data$brood_size)

# SS_error

ss_error_table = cel_data %>% group_by(genotype) %>% mutate(error_col = (brood_size - mean(brood_size)) ^ 2)
ss_error = sum(ss_error_table$error_col)

## SS_group

summary_table = cel_data %>% group_by(genotype) %>% summarise(group_mean = mean(brood_size), individuals = n())
ssgroup = sum(summary_table$individuals * (summary_table$group_mean - grand_mean) ^ 2)

```

### b) F-statistic and p-value

Calculate the F-statistic and the p-value using the `pf() function.` Recall that:

$$F=\frac{MS_{groups}}{MS_{error}} = \frac{SS_{groups}/df_{groups}}{SS_{error}/df_{error}}$$

```{r}
# degrees of freedom for SS_error
error_df = 90 - 3
# degrees of freedom for SS_groups
groups_df = 3 - 1

# F-statistic
fstat = (ssgroup/ groups_df) / (ss_error / error_df)
fstat
# p-value using CDF for F-statistic

summary(aov(brood_size ~ genotype, data = cel_data))

pf(fstat, groups_df, error_df, lower.tail = FALSE)

```


# Q3: Planned and Unplanned ANOVA

### a) Planned vs. Unplanned

Explain whether this experiment should be treated as a PLANNED or an UNPLANNED dataset. How would the analysis differ between these two scenarios?

```{r eval=FALSE}
# This is a planned dataset as the WT is effectively a control and we are interested if either of the genotypes are different than WT
```


### b) Planned experiment

Assume the data is a planned experiment. Use the `lm()` function to perform the analysis. 

First, use the `anova()` function on the results of of the `lm()` function to look at the overall result. *Note: you can use the special notation `$"Pr(>F)" to get the exact p-value from the `anova()` function.*

```{r}
# anova for lm of brood_size ~ genotype
anova_results = anova(lm(brood_size ~ genotype, data = cel_data))
anova_results$`Pr(>F)`
```

How do the $SS_error$, $SS_group$, $MS_error$, $MS_group$, $Fstat$, and $p$-value compare to the ones you calculated by  hand?

```{r eval=FALSE}
# your answer here
They are exactly the same.
```

Now, use the `summary()` function on the results of the the linear model.

```{r}

summary(lm(brood_size ~ genotype, data = cel_data))

```

What do the different parts of the `summary()` output tell you? What is R-squared and what does it mean?

```{r eval=FALSE}
# The r squared value tells you how much of the variance in the observed data is explained by the groups
```

Is the model significant overall? How about the different groups: are they significantly different from the control?

```{r eval=FALSE}
# Yes, the model is significant overall and both groups are significantly different than the control
```


### c) Unplanned experiment

Now assume the data are from an unplanned experiment. Use the `aov()` function to perform the analysis, and then use the `tukeyHSD()` function on the results of `aov()`.

```{r}

summary(aov(brood_size ~ genotype, data = cel_data))

```

Use the `tukeyHSD()` function on the results of `aov()` to perform the Tukey-Kramer test to look at all pairwise comparisons.

```{r}
TukeyHSD(aov(brood_size ~ genotype, data = cel_data))
```

What does Tukey's HSD test do? What do these results tell you? What is the biological interpretation of these results?

```{r eval=FALSE}
# The Tukey HSD test compares all of the groups pairwise with each other. The biological interpretation of this is that all of the groups are different than each other.
```


### d) Residuals

Once you have performed an ANOVA, you can also check your assumption of normality by checking to see if the ANOVA residuals are normally distributed. 

+ Do this by first assigning the results of your `aov()` function to a variable (e.g. `res_aov`.
+ Then, make a histogram of the residuals, which can be found in `res_aov$residuals`.
+ Finally, perform a test for normality on the residuals.

```{r}

res_aov = aov(brood_size ~ genotype, data = cel_data)
hist(res_aov$residuals)
```


### e) ANOVA for unequal variances

If the data look normal, but the variances are unequal, what kind of ANOVA could you do instead? *Hint: what kind of $t$-test would you use in this case?*

```{r eval=FALSE}
# You can do a welch's correction of an anova
```

Perform this test below. Also extract the exact $p$-value from the test.

```{r}
welches = oneway.test(brood_size ~ genotype, data = cel_data)
welches$p.value
```

How does the $p$-value compare with the ANOVA you performed above? Explain.

```{r eval=FALSE}
# It is slightly larger because this test has lower power so it will overestimate the p-value
```


# Q4: Fertilization phenotype

Now we will look at the fertilization phenotype. 

### a) Percent fertilization

+ Add a column to your original data frame containing the ***percent fertilization***.
+ Take a look at the percent fertilization using a stip chart / boxplot (as you did for Q1).

```{r}
# add percent fertilization column

cel_data_percent_fertile = cel_data %>% mutate(percent_fertilization = fertilized / brood_size)
cel_data_percent_fertile
# plot it

boxplot(percent_fertilization ~ genotype, data = cel_data_percent_fertile)

```


### b) Check assumptions for ANOVA

Perform the appropriate tests to check whether you can perform an ANOVA on these data and draw histograms to visualize the distributions. (You may again find it useful at this point to subset the data by group.)

```{r}
# test for assumptions

wt_data = cel_data_percent_fertile %>% filter(genotype == 'WT')
a_data = cel_data_percent_fertile %>% filter(genotype == 'A')
ab_data = cel_data_percent_fertile %>% filter(genotype == 'A;B')

hist(wt_data$percent_fertilization)
hist(a_data$percent_fertilization)
hist(ab_data$percent_fertilization)

shapiro.test(wt_data$percent_fertilization)
shapiro.test(a_data$percent_fertilization)
shapiro.test(ab_data$percent_fertilization)

# histograms

var.test(wt_data$percent_fertilization, a_data$percent_fertilization)
var.test(wt_data$percent_fertilization, ab_data$percent_fertilization)
var.test(a_data$percent_fertilization, ab_data$percent_fertilization)

```

Do these data meet the assumptions required for ANOVA? If not, what alternative tests could you perform? Explain the conditions under which either test would be appropriate.

```{r eval=FALSE}
# These data do not meet the appropriate assumptions required for ANOVA
```

### c) Statistical test

Perform an appropriate alternative to ANOVA for these data.

```{r}
kruskal.test(cel_data_percent_fertile$percent_fertilization, cel_data_percent_fertile$genotype)
```
 
 
# Q5: Error, Power, and Sample size

Recall that the power is $1−\beta$, where $\beta$ is the Type II error. We can calculate the power of the ANOVA test using the command `power.anova.test()`, which is contained in the base `stats` package. The arguments for the command are:

+ **groups** = the number of groups in the dataset  
+ **n** = the number of observations per group  
+ **between.var** = between-group variance, $SS_{group}$
+ **within.var** = within-group variance, $SS_{error}$  
+ **sig.level** = significance cutoff $\alpha$ (Type I error)  
+ **power** = power of the test ($1−\beta = 1$ - Type II error)  

Given all of the arguments except for one, this function will return a value for the missing argument.

### a) Power of ANOVA for this dataset

Given the data that we have been provided and the $p$-value we obtained for the `brood_size`, what is the power of our dataset, given a Type I error rate (significance cutoff) of $\alpha=0.05$?

```{r}
# first, check the documentation
help(power.anova.test)



# compute the power
power.anova.test(groups= 3, 
                 n= 30, 
                 within.var= ss_error, 
                 between.var= ssgroup, 
                 power=NULL,
                 sig.level= 0.05)
```

### b) Sample size

If we would like to achieve a power of 95% with a Type I error rate of $\alpha=0.05$, how many observations for each group would we need? 

```{r}
# compute the required sample size

power.anova.test(groups= 3, 
                 n= NULL, 
                 within.var= ss_error, 
                 between.var= ssgroup, 
                 power=0.95,
                 sig.level= 0.05)


# sample size needed = 2.74

```

How do you feel about this result?

```{r eval=FALSE}
# I think we should still have a bigger sample size as the chance of human error messing up the experiment is very high with only 2 or 3 individuals in each group
```

### c) Effect size

The ratio of the between-groups variation to the within-groups variation ($SS_{group}/SS_{error}$) gives us an approximate effect size.

The `brood_size` data here has a pretty big effect size, and we have a lot of power because of this. 

If the effect size were 10 times smaller, how many animals would we need to get a power of 95% at a Type I error threshold of 5%?

```{r}
new_ssgroup = ssgroup / 10

power.anova.test(groups= 3, 
                 n= NULL, 
                 within.var= ss_error, 
                 between.var= new_ssgroup, 
                 power=0.95,
                 sig.level= 0.05)

n = 15.225
```

<!-- That's it! Yay! Check to see if your file knits and then send this off! -->

<!-- Note: if you don't have LaTeX installed, or you had problems installing it, then the file may not knit. In that case, type `rm(list = ls())` in the console and then try running the entire file again to make sure that all of your final data structures are what you think they are. -->
