

## Exercise

22,071 physicians served as subjects to study effect of aspirin on incidence of heart attacks. There are two groups: those who took aspirin regularly for 5 years and others (11,043 of the 22,071) received placebo instead of aspirin. 189 of the group that took placebo suffered a heart attack and 104 of those who took aspirin suffered a heart attack. Does aspirin have a significant association with physicians suffering from a heart attack?
  
  In last week's class you used a Chi-square test to answer this question. Here let's use Fisher's exact test instead.

1. Create a contingency table for these data.

```{r}
# aspirin 
# heart attach
N = 22071
a = 189 # took placebo and had heart attack
b = 104 # took aspirin and had heart attack
c = 
d = 

placebo_total = 11043
placebo_heart = 189
placebo_healthy = placebo_total - placebo_heart

aspirin_total = N - placebo_total
aspirin_heart = 104
aspirin_healthy = aspirin_total - aspirin_heart


aspirindata = rbind(c(placebo_heart,aspirin_heart),
                    c(placebo_healthy, aspirin_healthy)
                    )
aspirindata

heart_total = placebo_heart + aspirin_heart
healthy_total = placebo_healthy + aspirin_healthy
```

2. Calculate the expected values.

```{r}
placebo_heart_expected = (placebo_total/N) * (heart_total/N) * N
placebo_heart_expected

placebo_healthy_expected = (placebo_total/N) * (healthy_total/N) * N
placebo_healthy_expected

aspirin_healthy_expected = (aspirin_total/N) * (healthy_total/N) * N
aspirin_healthy_expected

aspirin_heart_expected = (aspirin_total/N) * (heart_total/N) * N
aspirin_heart_expected


```


3. Calculate the $X^2$ value using your observed and expected counts.

```{r}

chisqvalue = sum( ((placebo_healthy - placebo_healthy_expected)^2)/placebo_healthy_expected +
                  ((placebo_heart - placebo_heart_expected)^2)/placebo_heart_expected +
                  ((aspirin_healthy - aspirin_healthy_expected)^2)/aspirin_healthy_expected +
                  ((aspirin_heart - aspirin_heart_expected)^2)/aspirin_heart_expected
                  )

chisqvalue

```


4. What is the $p$-value of the test statistic?

```{r}

pchisq(chisqvalue, df = 1, lower.tail = F)
chisq.test(aspirindata)
```


5. Calculate the $p$-value using Fisher's exact test.

```{r}
fisher.test(aspirindata)

```

6. Calculate the $p$-value using the `phyper` function.

```{r}
# phyper(q, m, n, k, lower.tail = TRUE, log.p = FALSE)
# 
# 
# x = , q	
# vector of quantiles representing the number of white balls drawn without replacement from an urn which contains both black and white balls.
# 
# m	
# the number of white balls in the urn.
# 
# n	
# the number of black balls in the urn.
# 
# k	
# the number of balls drawn from the urn.
# 
# p	
# probability, it must be between 0 and 1.
# 
# ```{r}
# x = aspirin_heart
# m = aspirin_total
# n = placebo_total
# k = heart_total
# 
# phyper(x,m,n,k,lower.tail = T )

```

---
## Exercise

22,071 physicians served as subjects to study effect of aspirin on incidence of heart attacks. There are two groups: those who took aspirin regularly for 5 years and others (11,043 of the 22,071) received placebo instead of aspirin. 189 of the group that took placebo suffered a heart attack and 104 of those who took aspirin suffered a heart attack. Does aspirin have a significant association with physicians suffering from a heart attack ? 

1. Create a contingency table for these data.

```{r}

```

2. Calculate the expected values.

```{r}

```


3. Calculate the $\chi^2$ value using your observed and expected counts.

```{r}



```


4. What is the $p$-value of the test statistic?

```{r}


```

