```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This covid.csv dataset contains the details of covid infections reported during the first two months of the pandemic.


Import the data and remove any rows containing 'NA'
```{r}



```

We suspect that most people who dont survive a covid infection are over the age of 60. Test the hypothesis that the mean age of covid patients who die is greater than 60.
```{r}
#What type of t.test is appropriate to test this hypothesis?



#Write the null and alternative hypothesis:



#Calculate the mean age of the people who dont survive:



#Calculate the sem for people who dont survive:



#Calculate the t-statistic for the hypothesized mean age of 60:



#Calculate the p.value  and 95% confidence interval



#Does the confidence interval contain the target value of 60 years and what does that tell us?



#Perform the test with the t.test function to confirm previous answers



#Use bootstrapping with 10000 trials to find the previously calculated p.value and 95% C.I 





```

Test the hypothesis that the mean age of patients who survive the infection is different from those who do not survive
```{r}

#Calculate the mean, variance and no of observations in the surviving and non-surviving groups


#Which type of t.test should be used to test this hypothesis?



#Calculate the standard error 



#Calculate the t.statistic 



#Calculate the p.value (Use 72.076 as degrees of freedom)




#Calculate the 95% confidence interval (Use 72.076 as degrees of freedom)


#Confirm your answers using the t.test function



```

Does the *gender* of a patient affect their chances of *survival*? Use any appropriate test to test this hypothesis.
```{r}


```

Is one gender more likely to be *infected* by covid than the other? Use any appropriate test to test this hypothesis.
```{r}


```

