---
title: 'Exercise: Exponential and Poisson'
subtitle: XDASI Fall 2021
date: "10/21/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Example: Lizard mortality

For some animals, like lizards (or hydra), their mortality rate is independent of their age. Over an extended trip to Costa Rica, you studied a large population of lizards and found that their rate of mortality is $\lambda$ = 0.05 deaths per month.

$\Rightarrow$ ***What is the expected number of deaths per year?***

```{r}
# Poisson: E(X) = SD(X) = lambda = 0.05 per month
0.05 * 12
```


$\Rightarrow$ ***What is the expected lifespan of one of these lizards?*** (In other words, this is the typical "wait time" until one death occurs?)

```{r}
# Exponential: E(X) = SD(X) = 1/lambda = 20 months
1/0.05
```


$\Rightarrow$ ***What is the probability an animal will survive more than 20 months?*** 

Hint: 

+ This is given by the the exponential survivorship function.

$$P_{exp}(X > 20) = e^{-0.05*20} = 0.3679 $$
+ Alternatively, it is given by the lower-tail Poisson probability for $\mu = \lambda * t = 0.05 * 20$:

$$P_{pois}(X = 0 ) = e^{-\lambda t}\frac{\lambda t^x}{x!} = e^{-\mu}\frac{\mu^x}{x!} = e^{-\lambda t}$$

```{r, collapse = T}
# with exponential
exp(-0.05*20)                       # manual calculation
pexp(20, rate = 0.05, lower.tail=F) # with CDF

# with Poisson: need to multiply units out to mu = lambda * t = 0.05/month * 20 months => 1 new unit
mu = 0.05*20 # number of deaths in t=20 months
ppois(0, lambda = mu, lower.tail=T)
dpois(0, lambda = mu) # same, since just one value for x
```

$\Rightarrow$ ***What is the probability an animal will survive 20 months or less?***

Hint:

+ This is given by the lower-tail CDF of the exponential distribution
+ This is equivalent to the upper-tail CDF of the Poisson distribution.


$$ P(X \le 20) = 1 - e^{-0.05*20} = 1 - 0.3679 = 0.6321 $$

```{r, collapse = T}
1-exp(-0.05*20) # manual calculation
pexp(20,0.05)   # with CDF

ppois(0, lambda = mu, lower.tail=F) # P(no death > 20 months)
```


$\Rightarrow$ ***What is the probability an animal will survive between 10 and 15 months?*** 

Hint: 

+ To get the total probability across an interval, we subtract the total probability of the lower value from that of the higher value.
+ We can't answer this with Poisson, since can only vary # events (not time)!


$$ P(10 \le X \le 15) = (1 - e^{-0.05*15}) - (1 - e^{-0.05*10}) = e^{-0.05*10} - e^{-0.05*15} =  0.165 $$

```{r, collapse = T}
exp(-0.05*10) - exp(-0.05*15) # manual calculation
pexp(15,0.05) - pexp(10,0.05) # with CDF
```

