```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Scenario

You want to clone a gene and you have some reporter for success of transformation in *E. coli*. Positive colonies are blue because they express $\beta$-galactosidase when plated on medium with IPTG. The probability of a successful transformation is $0 \le P(success) \le 1$. For the following exercise, you may consider one plate, many plates, or a truly gargantuan number of plates.

## Exercise

The class will split into 8 groups and each one will take one of the following distributions:

- Bernoulli
- Binomial
- Geometric
- Hypergeometric
- Negative binomial
- Poisson
- Normal
- Exponential

For your assigned distribution, answer the following questions. At the end of the class, each group will present their distribution and how it applies to the scenario above to the rest of the class.

#### a. What is your function? Describe it in terms of the kind of problem that it applies to. Is it discrete or continuous?

<!-- Hyper Geometric
It is a discrete function and you sample without replacement from a universal value. Each trial outcome affects the outcome of the next trial. -->


#### b. Describe a question that you can ask with your distribution as applied to the transformation scenario given above.

<!-- There are 100 bacterial colonies on a plate. 25 colonies are blue and 75 colonies are white. If you take a sample 10 colonies, whats the probability of getting 10 blue transformed colonies after 10 trials? -->


#### c. Describe the parameterization of the equation. What parameters does it take, and what do they mean?

<!-- Hyper Geometric parameters:
N = 100
n = 10
x = 10
M = 25  <- blue
M = 75  <- white
 -->


#### d. Pick some appropriate parameters for your distribution as applied to the transformation scenario and graph the PDF and CDF.

```{r}

library(ggplot2)
library(reshape2)

blue <- c(0,10,20,30,40,50,60,70,80,90,100) # our parameter

# make PDF
pdf <- NULL
for (i in blue){
  xx <- dhyper(1:10, i, 100-i, 10)
  pdf <- cbind(pdf, xx) 
}

colnames(pdf) <- paste("blue", blue, sep = "")

# make CDF
cdf <- NULL
for (i in blue){
  xx <- phyper(1:10, i, 100-i, 10)
  cdf <- cbind(cdf, xx) 
}

colnames(cdf) <- paste("blue", blue, sep = "")

# make dataframe for ggplot
pdf2 <- data.frame(pdf, number=1:10)
pdf2 <- melt(pdf2, "number")
cdf2 <- data.frame(cdf, number=1:10)
cdf2 <- melt(cdf2, "number")

# plot PDF with 10 different parameters 
ggplot()+
  geom_line(aes(y = value, x = number, colour = variable),
                     data = pdf2, stat="identity") +
  theme(legend.title = element_blank(),
        legend.position="top")
 
# plot CDF with 10 different parameters
ggplot()+
  geom_line(aes(y = value, x = number, colour = variable),
            data = cdf2, stat="identity") +
  theme(legend.title = element_blank(),
        legend.position="top")


```


#### e. Compute a cumulative probability for some appropriate value and describe in words what it means. Be prepared to explain your answer to the class.

```{r}
# when there are 50 blue colonies among 100 colonies, we will sample 10 colonies. 
blue50situation_PDF <- pdf[,5]
blue50situation_CDF <- cdf[,5]

plot(blue50situation_CDF)
# according to the plot of CDF, we can obtain at least 6 colonies by 95.42% when sampling 10 colonies.
```




#### f. Identify at least one other distribution that yours relates to and be prepared to discuss the relationship between them. This could be a special case, a limiting case, an approximation, or an inverse CDF (survival) relationship.

<!-- When the number of colonies is gargantuan, we can consider binominal distribution. This is because whether replacing or not doesn't affect outcomes after a certain of trials.  -->
