```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Scenario

You want to clone a gene and you have some reporter for success of transformation in *E. coli*. Positive colonies are blue because they express $\beta$-galactosidase when plated on medium with IPTG. The probability of a successful transformation is $0 \le P(success) \le 1$. For the following exercise, you may consider one plate, many plates, or a truly gargantuan number of plates.

## Exercise

The class will split into 8 groups and each one will take one of the following distributions:

- Bernoulli
- Binomial
- Geometric
- Hypergeometric
- Negative binomial
- Poisson
- Normal
- Exponential

For your assigned distribution, answer the following questions. At the end of the class, each group will present their distribution and how it applies to the scenario above to the rest of the class.


# Marcel and Avrami_Poisson Distribution

#### a. What is your function? Describe it in terms of the kind of problem that it applies to. Is it discrete or continuous?

<!--Insert your answer below. -->

Our function is the prpbability, p, of geting a number, x, of blue cells per plate. A blue cells means the plasmid harboring the gene for beta galactosidase was transformed, and is considered a success. IOur distirbution is discrete, and there are only two possibilites, while cells or blue cells (0 or 1).  

#### b. Describe a question that you can ask with your distribution as applied to the transformation scenario given above.

<!-- Insert your answer below. -->

Knowing the average number of blue cells per plate, what is the PMF of blue cells per plate within the range of 0:100. Cells that are blue have been succesfully transformed.

#### c. Describe the parameterization of the equation. What parameters does it take, and what do they mean?

<!-- Insert your answer below. -->

Lamba, which is a rate of events/time or events/space. We chose events/space because blue white screening works better with space. In other words, you either have blue or white cells (events) on the plate (space). This is the only parameter we need. The possion distirubution will give us the probability of each number of blue cells within on our range based on the lambda parameter. 

#### d. Pick some appropriate parameters for your distribution as applied to the transformation scenario and graph the PDF and CDF.

```{r}
Lamnda <- 40 #blue cells per plate
Range.vector <- 0:100

# We chose our rate to be 40 blue cells/plate because based on our lab experience, that is usually how many blue cells we've gotten when transforming lac z plasmids.
```


#### e. Compute a cumulative probability for some appropriate value and describe in words what it means. Be prepared to explain your answer to the class.

```{r}

par(mfrow = c(1,3))

PDF.Poisson <- dpois(range.vector, lambda = blue.cells.per.plate)
plot(PDF.Poisson, type = "l", main = "Marcel's Swag", xlab = "# Blue cells/plate", ylab = "PDF")

CDF.Poisson <- ppois(range.vector, lambda = blue.cells.per.plate)
plot(CDF.Poisson, type = "l", main = "Avrami's Swag", xlab = "# Blue cells/plate", ylab = "CDF")

```

<!-- Insert your answer below. -->

Between 0 blue cells and 100 blue cells, we have the highest probability of getting 40 blue cells. That probability is still low, lying at 0.06. 

#### f. Identify at least one other distribution that yours relates to and be prepared to discuss the relationship between them. This could be a special case, a limiting case, an approximation, or an inverse CDF (survival) relationship.

<!-- Insert your answer below. -->

One other distribution that relates for Poisson is exponential, as it can also account for series of events that account for space & time. Exponential can be used for approximations or survival relationships.
