---
title: "Week 02 - Writing R Functions"
author: "Manpreet S. Katari"
output:
  html_document:
    toc: yes
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Why write your own function?

+ Allows you to access code that you or someone has already written
  + Many functions are already written so it is possible that you may never have to but there are cases where small functions become helpful.
+ Can include complicated steps of an algorithm
  + For example implementing a new method that you have created to calculate confidence in probe signals.
+ Can include simply steps of a workflow
  + For example running all the steps to normalizing data.	
+ You can use the apply function to make your code more efficient.


## Architecture of a function

+ A function is divided into 3 parts:
  + Name – the name of the function that has to be used when calling the function
  + Arguments – the required and optional objects and settings that the software can/will use
  + Expression – the code that does the work
  + It is a good idea to include a return command in the expression block


```{r eval=FALSE}
#this is a template of how you would write a function
function_name<-function(argument1, argument2, ....) {
  expression
  ....
  return(result)
}

#this is how you would run your function and save the result
myresult = function_name("arg1","arg2")

```

Here's an example:

```{r}
squareme <- function(x) {
  result = x**2
  return(result)
}

squareme(3)
```

```{r eval=FALSE}
squareme("hello") # this will give an error

```

```{r}
squareme(c(3,4,5,6))

```

## Control Statements

Now let's give our code some more structure and decision making power.

### If-else statements

**if-else** is a control statement that is used to evaluate a condition.
+ '`if-else` evaluates a logical test expression.
+ If the condition is **TRUE**, a given statement will be executed; if it is **FALSE**, then a different statement will be executed.
+ `if-else` statements can be laddered and nested as well.


```{r eval=F}
if ( condition ) {
  then do this
} else {
  do something else
}

```

#### Simple working example

```{r}

a<-20
if(a>5) {
  print(a) 
} else {
  print("a is less than 5")
}
```

```{r}
a<-2
if(a>5) {
  print(a) 
} else {
  print("a is less than 5")
}
```

#### Nested if-else

If we need to check for more than one condition, we can create a nested `if-else` statement. You can use as many nested layers as you like. Just be careful that all of your curly brackets are correctly matched!

```{r}

foo<-function(a) {
	if(a<5) {
		print("a is less than 5")
	} else {
	  if(a<10) {
			print("a is less than 10 but greater than 5")
	  } else {
			print ("a is greater than or equal to 10")
	  }
	}
}

foo(2)
foo(7)
foo(20)
```


#### Laddered if-else

The code above can be simplified into a single-layered expression by combining the second `if` statement with the `else` statement that comes right before it: 

```{r}

foo<-function(a) {
	if(a<5) {
		print("a is less than 5")
	} else if(a<10) {
			print("a is less than 10 but greater than 5")
	} else {
			print ("a is greater than or equal to 10")
	}
}

foo(2)
foo(7)
foo(20)
```


#### ifelse command

There is also an `ifelse` command that can be used to evaluate a vector rather than just one condition. 
+ The first argument is the test condition, a logical expression involving the vector.
+ The 2nd and 3rd arguments are assignment statements to be executed if the condition is **TRUE** or **FALSE**, respectively.
+ The test condition is evaluated for each element of the vector.
+  The `ifelse` statement returns a vector of the same length as the input vector, containing values assigned according to the result of each comparison.
+ The values of the original vector are not changed (unless you assign the result of the statement back to the same vector).

In the example below, note that the result of the `ifelse` statement is a character vector rather than a numeric vector because a vector can contain only one data type. Since some of the elements are assigned a string, the numeric elements are coerced into strings as well.

```{r}
a<-c(1:8)
b = ifelse(a <= 5, a, "greater than 5")
a  # stays the same
b  # a character vector of the same length as a
```

## Loops

There are cases when you need to loop through each element of a data object. Loops run repeatedly through the code that is provided within curly brackets until some termination condition occurs, at which point the loop exits and flow control is returned.

+ `for` loops are a great option when you know the number of iterations.
+ `while` loops evaluate a condition and loop until it is no longer true.
+ `repeat` loops repeat a block of code until the `break` command is used to exit the loop.

Let's create random matrix and look at all the ways we can calculate the mean of each row.

```{r}
# The statemwnt below uses the 'sample' function to 20 random values
# ranging from 20 to 160, then divides them by 10.
# Finally, it assigns these values to a matrix called 'sampleData',
# which has 5 rows and 4 columns.

sampleData<-matrix(sample(20:160, 20, replace=T)/10, ncol=4, nrow=5)
sampleData # inspect the matrix

# Your sampleData will be different because it is generated randomly!

# Since we know that the number of rows is 4, we can use the 'for' loop.

# 'i' is the iterator: on each cycle through the code, it will change in value. 
# Here, the iterator ranges from 1 to the number of rows in the matrix, 
# incrementing by 1 on each iteration through the loop.

# Let's save the results in a vector called 'sampleMean'.
# Initialize an empty vector (needs to be defined so you can assign to it)
sampleMean=numeric()

# store the row means of 'sampleData' in a vector called 'sampleMean'
# The 'i'th row of sampleData is stored in the 'i'th element of 'sampleMean'. 
for ( i in 1:nrow(sampleData) ) {
    sampleMean[i] = mean(sampleData[i,]) # It is important to use i to 
                                         # specify where the value should be
                                         # retrieved and where it should be 
                                         # saved.
}
sampleMean

```


```{r}
# In cases when we want to use a 'while' loop, we need to use a counter that 
# will keep track of which row we are analyzing on each iteration.
# The counter needs to be updated explicitly, or else you may end up in an infinite loop.

counter=1

while( counter <= nrow(sampleData)) { # I want to loop while this cond is true
    sampleMean[counter] = mean(sampleData[counter,]) # now instead of 'i' we use 
                                                     # 'counter' to keep track
    counter = counter + 1 # don't forget to add 1 to the counter so you move on 
                          # to the next row; otherwise the loop will be stuck!
  }

counter
sampleMean

```


## Apply functions

Instead of looping through a matrix or vector of values to perform the same operation, R has commands that will allow you to apply a function to each row, column, every element of a list, or simply on a vector that is grouped using a factor.

+ `apply()` – first provide the array; then whether to apply the function by row (1), column (2), or both (c(1,2)); then the name of the function; and then optionally any options for the function that is being applied.
+ `tapply()` – similar to apply, but pass a factor vector instead of row or column.
+ `lapply()` – simply provide a list and the function to apply to each vector in the list. The result is also a list.
+ `sapply()` – same as lapply(), but returns the result in its original data structure if possible ("simplified" apply).

For the example above, let's use the `apply` function to calculate the mean.

```{r}
apply(sampleData, 1, mean) # the "1" tells apply to take the mean of each row

```

Let's assume the data was actually two replicates of control in the first two columns and two replicates of treatment in the last two columns. Let's create a factor that defines the different groups.

```{r}
colnames(sampleData)<-c("ctr1l", "ctrl2", "trt1", "trt2")
rownames(sampleData)<-c("gene1", "gene2", "gene3", "gene4", "gene5")

expgroups = factor(c("ctrl", "ctrl", "trt", "trt"))

# check it out
sampleData
str(sampleData) 
```

We can use `tapply()` to calculate the average among the `ctrl` and `trt` groups for a given gene:

```{r}
sampleData[1,]

# mean of ctrl and trt groups for first row of sampleData (gene1)
tapply(sampleData[1,], expgroups, mean)

```

We could do this manually for each row of the matrix:

```{r}
# apply 'mean()' to each row of SampleData, grouping data by expgroups:
# This takes the row mean across both replicates for each factor ("ctrl" and "trt")
tapply(sampleData[1,], expgroups, mean)
tapply(sampleData[2,], expgroups, mean)
tapply(sampleData[3,], expgroups, mean)
tapply(sampleData[4,], expgroups, mean)

```

However, this is not so useful in practice, since you will usually be working with a lot more than four rows!

We could automate this using a `for()` loop:

```{r}
sampleMeans = matrix(NA, ncol=2, nrow(sampleData)) # initialize a matrix

for (i in 1:nrow(sampleData)) { # iterate over the rows
    sampleMeans[i,] = tapply(sampleData[i,], expgroups, mean)
}
sampleMeans
```

But wait! We can make this a lot easier!

We can do this a lot more efficiently using `apply()` to call the `tapply()` function above on each row of the matrix. We can do this in two different ways:
1. Create a **function** that uses `tapply()` to take the mean of a single vector. Then, call this function with `apply()` to get the mean for each row.
2. Use `apply()`  to pass the matrix data directly to `tapply()`, one row at a time. This is a one-liner! (much simpler :)

Let's unpack Option 1 first:

```{r}
## Option 1: 

# First, create a function that uses `tapply()' to compute the mean of the replicates
# for the ctrl and trt groups (i.e. the mean for each factor).
# Then, use it to get the means for all the rows using 'apply()'.

# The function will take two arguments:
# 1) the values that are needed to perform the calculation (a numerical vector)
# 2) the grouping of the values (a factor)
calcmean = function(genevalues, expgroup) {
  result = tapply(genevalues, expgroup, mean)
  return(result)
}
apply(sampleData, 1, calcmean, expgroups)

## Note on 'apply()' syntax:
# 1) 'sampleData' = the first argument that gets passed to 'calcmean()'
# 2) 1 = apply 'calmean()' to each row of 'sampleData'
# 3) 'calcmean' = the function to apply here
# 4) 'expgroups' (the factor groupings) = the 2nd argument passed to 'calcmean()'

```

Above is a great example of how creating your own function can help speed up the analysis because you can simply apply it to each row.

However, there is an even simpler way!

**Option 2:** You can do the same thing as above in one line, using `apply()` to call `tapply()` directly. This works just as well, and is simpler syntactically. Now, you can get the row means for each factor across the entire matrix, all in one go.

Remember the following when using `apply()`:
- 1st argument: will be applied to the function call you are using
- 2nd argument: specifies how to apply the function: 1=row-wise, 2=column-wise, c(1,2)=both.
- 3rd argument: the name of the function you want to apply. Note that you do not include parentheses here.
- 4th+ arguments: any additional arguments to be passed to the function

```{r}
# three arguments will be passed to 'tapply()':
# sampleData, expgroups, mean
apply(sampleData, 1, tapply, expgroups, mean) 

# you can transpose the matrix to get the same format back
t(apply(sampleData, 1, tapply, expgroups, mean))

```

A final note about functions and scoping:

```{r}
#####################################################################################
## A note about functions:
# A function's arguments are just *containers* for the data that gets passed to it --
# These variables only exist within the function. This is called "lexical scoping".
# If you want to change the contents of a variable by calling a function on it,
# you will need to reassign the output of the function to that variable.
# Otherwise, it will remain unchanged after the function call. Here is an example:
x = 1
y = 2
test_fctn <- function(x, y) {
  x = y + 1
  y = y**2
  return(c(x,y)) # R only lets you return one variable at a time! ;-(
}
# first call: only z gets changed
z = test_fctn(x,y)
x
y
z
# second call: only x gets changed
x = test_fctn(y,z)
x
y
z
#####################################################################################
```
