---
title: "Tabular Statistics"
author: "Manpreet S. Katari / Kris Gunsalus"
date: "October 23, 2019"
output:
  pdf_document:
    toc: yes
    toc_depth: '2'
  html_document:
    toc: yes
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Tabular Statistics in-class Exercise (excerpted from TabularStats_2019.Rmd)

22,071 physicians served as subjects to study effect of aspirin on incidence of heart attacks. There are two groups: those who took aspirin regularly for 5 years, and others (11,043 of the 22,071) who received placebo instead of aspirin. 189 of the group that took placebo suffered a heart attack and 104 of those who took aspirin suffered a heart attack. Does aspirin have a significant association with physicians suffering from a heart attack ? 

1. Create a contingency table for these data.

```{r}
# data (quadrants a-d)
a = 189    # took placebo and heart attack
b = 104    # took aspirin and heart attack
c = 10854  # took placebo and healthy
d = 10924  # took aspirin and healthy

N = 22071  # total

placebo_total = 11043  # col1 margin
placebo_heart = 189
placebo_healthy = placebo_total - placebo_heart

aspirin_total = N - placebo_total
aspirin_heart = 104
aspirin_healthy = aspirin_total - aspirin_heart

heart_total = placebo_heart + aspirin_heart
healthy_total = placebo_healthy + aspirin_healthy

aspirindata = rbind(c(placebo_heart, aspirin_heart),
                    c(placebo_healthy, aspirin_healthy))
aspirindata
```

2. Calculate the expected values.
```{r}

placebo_heart_expected = (placebo_total/N) * (heart_total/N) * N
placebo_heart_expected

aspirin_heart_expected = (aspirin_total/N) * (heart_total/N) * N
aspirin_heart_expected

placebo_healthy_expected = (placebo_total/N) * (healthy_total/N) * N
placebo_healthy_expected

aspirin_healthy_expected = (aspirin_total/N) * (healthy_total/N) * N
aspirin_healthy_expected
```

3. Calculate the $X^2$ value using your observed and expected counts.
```{r}
chisq = sum(
  (placebo_healthy - placebo_healthy_expected)**2 / placebo_healthy_expected + 
  (placebo_heart   - placebo_heart_expected)**2   / placebo_heart_expected +

  (aspirin_healthy - aspirin_healthy_expected)**2 / aspirin_healthy_expected +
  (aspirin_heart   - aspirin_heart_expected)**2   / aspirin_heart_expected
)
```


4. What is the $p$-value of the test statistic?
```{r}
pchisq(chisqvalue, df = 1, lower.tail = F)
chisq.test(aspirindata)
```

5. Calculate the $p$-value using Fisher's exact test.
```{r}
fisher.test(aspirindata)
```

6. Calculate the $p$-value using the `phyper` function.
```{r}

# this is in a wrong orientation
# relative to how Manny defined the data
# but it doesn't matter -- you just have to choose the correct tail later

# same as 'data_matrix' used above
x = aspirin_heart  # 104
m = aspirin_total  # 293    # = row 1 margin
n = placebo_total  # 10245  #  = row 2 margin = c+d
k = heart_total    # 2181   # = col 1 margin = a+c
N = 22071

# (although table orientation doesn't matter for the Fisher exact test)
aspirindata = matrix(c(x,   m-x,      # a, b
                       k-x, n-(k-x)), # c, d
                     2,2,
                     byrow = TRUE) # matrices fill by columns by default

# Fisher test, alternative = 'less'
fisher.test(aspirindata, alternative='less')$p.value

# Hypergeometric
q = x
phyper(q, m, n, k, lower.tail = T)
```


