---
title: "PCA"
author: "Chris"
date: "12/06/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(ggplot2)
require(RColorBrewer)
require(cowplot)
require(scatterplot3d)
require(MASS)
ggplot2::theme_set(theme_classic())
```

#### Load Data

```{r}
data <- read.delim("HW9_EXPRESSION_DATA.tsv", check.names = F)
meta <- read.delim("HW9_SAMPLE_DATA.tsv")
```

#### Remove zeros

```{r}
data <- data[rowSums(data) != 0, ]
```

#### PCA

Let's create a toy example

```{r}
straight.line.data <- data.frame(x = 1:10, y=1:10)
qplot(x, y, data=straight.line.data, geom="point")
```

Can we do PCA on this example?

```{r}
straight.line.pca <- prcomp(straight.line.data, center=T, scale=T)
summary(straight.line.pca)
```

What does this mean? Can we plot the resulting projection?require(cowplot

```{r}
qplot(PC1, PC2, data=as.data.frame(straight.line.pca$x), geom="point") + scale_y_continuous(limit=c(-1,1))
```

Let's add a third dimension

```{r}
threed.data <- cbind(straight.line.data, data.frame(z=10:1))
plot_grid(qplot(x, y, data=threed.data, geom="point"),
          qplot(x, z, data=threed.data, geom="point"),
          qplot(y, z, data=threed.data, geom="point"), 
          ncol=2)

scatterplot3d(threed.data, color="navyblue", pch=16)
```

Aight, now how does the PCA look?

```{r}
threed.pca <- prcomp(threed.data, center=T, scale=T)
summary(threed.pca)

qplot(PC1, PC2, data=as.data.frame(threed.pca$x), geom="point") + scale_y_continuous(limit=c(-1,1))
```

#### PCA on Noisy Data

```{r}
noisy.data <- data.frame(x=rnorm(10000, mean=0, sd=1),
                         y=rnorm(10000, mean=0, sd=10),
                         z=rnorm(10000, mean=0, sd=100))
scatterplot3d(noisy.data, color="navyblue", pch=16)
```

```{r}
noisy.data.pca <- prcomp(noisy.data, center=T, scale=F)
summary(noisy.data.pca)
qplot(PC1, PC2, data=as.data.frame(noisy.data.pca$x), geom="point") + labs(x="PC1 (99%)", y="PC2 (1%)")
```

How does this affect distance between the points?

```{r}
dist.data <- dist(noisy.data)
dist.pca.data <- dist(noisy.data.pca$x)
sum(dist.data)
sum(dist.pca.data)
sum(dist.data - dist.pca.data)
```

Can I rotate this back?

```{r}
# Original transformation
noisy.rotated <- as.matrix(scale(noisy.data, scale=F)) %*% noisy.data.pca$rotation
sum(noisy.rotated - noisy.data.pca$x)

# Rotate backwards
noisy.rotated.back <- noisy.data.pca$x %*% MASS::ginv(noisy.data.pca$rotation)
sum(noisy.rotated.back - scale(noisy.data, scale=F))
```

How about scaling?

```{r}
noisy.data.pca.scaled <- prcomp(noisy.data, center=T, scale=T)
summary(noisy.data.pca.scaled)
qplot(PC1, PC2, data=as.data.frame(noisy.data.pca.scaled$x), geom="point") + labs(x="PC1 (33%)", y="PC2 (33%)")
```

#### PCA on Real Data

```{r}
data.pca <- prcomp(data, center=T, scale=T)
summary(data.pca)
qplot(PC1, PC2, data=as.data.frame(data.pca$x), geom="point")
```

```{r}
data.pca.t <- prcomp(t(data), center=T, scale=T)
summary(data.pca.t)
qplot(PC1, PC2, data=as.data.frame(data.pca.t$x), geom="point")
```

#### Attach metadata

```{r}
data.pca.t.df <- cbind(as.data.frame(data.pca.t$x), meta)
sum(rownames(data.pca.t.df) == data.pca.t.df$Sample_ID)
```

#### Plot with metadata

```{r}
ggplot(data.pca.t.df, aes(x=PC1, y=PC2, shape=Experiment, color=TF)) +
  geom_point() +
  scale_color_brewer(palette = "Set3") +
  labs(x="PC1 (43.7%)", y="PC2 (16.1%)")
```

#### How much does each gene contribute to PC1

```{r}
# What does PC1 mean for expression data
# Do a matrix inverse on PC1
rotation.inverse <- MASS::ginv(data.pca.t$rotation[, "PC1", drop=F])
colnames(rotation.inverse) <- row.names(data.pca.t$rotation)

# Calculate the expression values from the PC1 loading and the PC1 matrix inverse
data.pc1 <- as.matrix(data.pca.t.df[, "PC1", drop=F]) %*% rotation.inverse

# Unscale
data.pc1 <- sweep(data.pc1, 2, data.pca.t$scale, FUN = "*")
# Uncenter
data.pc1 <- sweep(data.pc1, 2, data.pca.t$center, FUN = "+")
```
