---
title: "Exercise: Chi-Square Test of Independence"
subtitle: "XDASI Fall 2021"
date: "10/7/2021"
output:
  html_document:
    toc: no
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Effect of aspirin on heart attack risk

22,071 physicians served as subjects to study the effect of aspirin on the incidence of heart attacks. There were two groups: 

+ 11,028 subjects took aspirin regularly for 5 years
+ 11,043 received a placebo instead of aspirin 

At the end of the study, 189 in the placebo group and 104 of those who took aspirin had suffered a heart attack. 

***Does aspirin have a significant association with physicians suffering from a heart attack?***

### 1. Create a contingency table for these data.

```{r}
# data (quadrants a-d)
placebo_heart   = 189    # a: took placebo and heart attack
aspirin_heart   = 104    # b: took aspirin and heart attack
placebo_healthy = 10854  # c: took placebo and healthy
aspirin_healthy = 10924  # d: took aspirin and healthy

placebo_total = 11043    # column 1 margin
aspirin_total = 11028    # column 2 margin

heart_total   = placebo_heart   + aspirin_heart    # row 1 margin
healthy_total = placebo_healthy + aspirin_healthy  # row 2 margin

Observed = rbind(c(placebo_heart,   aspirin_heart),
                 c(placebo_healthy, aspirin_healthy))
rownames(Observed) = c("Heart",  "Healthy")
colnames(Observed) = c("Placebo","Aspirin")

Observed
```

### 2. Calculate the expected values for each group.

```{r}
N = 22071  # total

placebo_heart_expected   = (placebo_total/N) * (heart_total/N)   * N
aspirin_heart_expected   = (aspirin_total/N) * (heart_total/N)   * N
placebo_healthy_expected = (placebo_total/N) * (healthy_total/N) * N
aspirin_healthy_expected = (aspirin_total/N) * (healthy_total/N) * N

Expected = rbind(c(placebo_heart_expected,   aspirin_heart_expected),
                 c(placebo_healthy_expected, aspirin_healthy_expected))
rownames(Expected) = c("Heart",  "Healthy")
colnames(Expected) = c("Placebo","Aspirin")

Expected
```

### 3. Calculate the $X^2$ test statistic using observed and expected counts.

```{r}
# chisq = sum(
#   (placebo_healthy - placebo_healthy_expected)**2 / placebo_healthy_expected + 
#   (placebo_heart   - placebo_heart_expected)**2   / placebo_heart_expected +
# 
#   (aspirin_healthy - aspirin_healthy_expected)**2 / aspirin_healthy_expected +
#   (aspirin_heart   - aspirin_heart_expected)**2   / aspirin_heart_expected
# )
# chisq

# same, but much easier!
chisq = sum( abs(Observed - Expected)^2 / Expected)
chisq
```


### 4. Compute the $p$-value of the test statistic using the $\chi^2$ CDF.

```{r}
pchisq(chisq, df = 1, lower.tail = F)
```

### 5. Calculate the $p$-value using the `chisq.test()` function.

```{r}
chisq.test(Observed)
```

### 6. Apply Yates' continuity correction to your manual calculation of the $X^2$ test statistic.

```{r}
new.chisq = sum( (abs(Observed - Expected) - 0.5)^2 / Expected)
new.chisq
```

### 7. Recalculate the $p$-value using `pchisq()`.

```{r}
pchisq(new.chisq, df = 1, lower.tail = F)
```

### 8. Why did we need to apply Yates' correction?

```{r}
# The normal distribution is a continuous approximation and underestimates the 
# significance of the discrete differences without this correction.
```

### 9. Use a two-sample $z$-test to perform the same test.

```{r}
prop.test(Observed)
```

### 10. How do the results compare? 

```{r}
# they are the same since they both use the normal approximation
# they just use a different formulation and different test statistic
```

